Estradiol elicits distinct firing patterns in arcuate nucleus kisspeptin neurons of females through altering ion channel conductances

Version 1

Abstract

Abstract

Hypothalamic kisspeptin (Kiss1) neurons are vital for pubertal development and reproduction. Arcuate nucleus Kiss1 (Kiss1ARH) neurons are responsible for the pulsatile release of Gonadotropin-releasing Hormone (GnRH). In females, the behavior of Kiss1ARH neurons, expressing Kiss1, Neurokinin B (NKB), and Dynorphin (Dyn), varies throughout the ovarian cycle. Studies indicate that 17β-estradiol (E2) reduces peptide expression but increases Vglut2 mRNA and glutamate neurotransmission in these neurons, suggesting a shift from peptidergic to glutamatergic signaling. To investigate this shift, we combined transcriptomics, electrophysiology, and mathematical modeling. Our results demonstrate that E2 treatment upregulates the mRNA expression of voltage-activated calcium channels, elevating the whole-cell calcium current and contributing to high-frequency firing. Additionally, E2 treatment decreased the mRNA levels of Canonical Transient Receptor Potential (TPRC) 5 and G protein-coupled K+ (GIRK) channels. When TRPC5 channels in Kiss1ARH neurons were deleted using CRISPR, the slow excitatory postsynaptic potential (sEPSP) was eliminated. Mathematical modeling confirmed the importance of TRPC5 channels for initiating and sustaining synchronous firing, while GIRK channels, activated by Dyn binding to kappa opioid receptors, were responsible for repolarization. Our findings suggest that E2 modifies ionic conductance in Kiss1ARH neurons, enabling the transition from high frequency synchronous firing through NKB-driven activation of TRPC5 channels to a short bursting mode facilitating glutamate release. In a low E2 milieu, synchronous firing of Kiss1ARH neurons drives pulsatile release of GnRH, while the transition to burst firing with high, preovulatory levels of E2 facilitates the GnRH surge through its glutamatergic synaptic connection to preoptic Kiss1 neurons.

Introduction

Hypothalamic kisspeptin (Kiss1) neurons and its cognate receptor (GPR 54 or Kiss1 R) are essential for pubertal development and reproduction, and may also be involved in the control of energy homeostasis (Kotani et al., 2001) (De Roux et al., 2003) (Seminara et al., 2003) (Messager et al., 2005) (Shahab et al., 2005) (d’Anglemont de Tassigny et al., 2008) (Qiu J. et al., 2018) (Rønnekleiv et al., 2022). Kisspeptin neurons within the arcuate nucleus of the hypothalamus (Kiss1ARH) co-express Kiss1, Neurokinin B (NKB) and Dynorphin (Dyn), which are all down-regulated by 17β-estradiol (E2) (Goodman et al., 2007; Navarro V. M. et al., 2009). However, Kiss1ARH neurons also express vesicular glutamate transporter 2 (vGlut2) and release glutamate, and both vGlut2 expression and glutamate release are upregulated by E2 in females (Qiu J. et al., 2018). This indicates that peptides and glutamate in Kiss1ARH neurons are differently modulated by E2 and suggests that there is a complex E2 regulation in these neurons such that they transition from predominantly peptidergic to glutamatergic neurotransmission and hence from a “pulsatile” to “surge” mode (Figure 1). It has been known for decades that neurons located within the arcuate nucleus are responsible for the pulsatile release of GnRH and subsequently pulsatile release of LH from the pituitary gland (O’Byrne et al., 1991) (Moenter et al., 1993). In this respect, it is now generally accepted that Kiss1ARH neurons are the main neurons responsible for the generation of pulsatile LH release, but the underlying cellular conductances generating this activity have not been elucidated.

LH pulse and surge profiles in mice with associated firing activity of Kiss1ARH neurons.

A, Representative example of an LH pulse profile in an ovariectomized (OVX) mouse in the absence of gonadal steroid feedback. Pulses detected by the DynPeak algorithm are indicated with an asterisk (Lin et al., 2021). B, LH surge profile (closed symbol) in mice undergoing the OVX+E2+E2 surge inducing protocol (6-10 days post OVX and implantation of 17β-estradiol capsule delivering diestrous levels of steroid, mice received subcutaneous injections of estradiol benzoate (1µg/20g body weight) on 2 consecutive days at 08:30 h with blood samples collected every 30 min from 14:30 - 20:30 h for LH measurement (Lin et al., 2021); mean ± SEM; (n=9). Open symbol representing control mice not receiving the second dose of estradiol benzoate (OVX+E2+oil). The expected LH surge occurred approximately 1 h before lights off (19:00 h). Values significantly different from basal LH concentrations are indicated by # (#p<0.05, ##p<0.01, repeated measures ANOVA). C, illustration of synchronized firing of a Kiss1ARH neuron induced by TACR3 agonist senktide in an OVX female. D, demonstration of burst firing of a Kiss1ARH neuron induced by glutamate in an E2-treated, OVX female. The spike activities in C and D have been expanded to emphasize the notable effects of senktide or glutamate on the firing activity of Kiss1ARH neurons.

Morphological studies have provided evidence that Kiss1ARH neurons can communicate directly with each other (Lehman et al., 2010) (Navarro V. M. et al., 2009) (Navarro V.M. et al., 2011). Furthermore, Kiss1ARH neurons express the NKB receptor, Tacr3, as well as the kappa (κ) opioid receptor (KOR), whereas GPR54 is not expressed in Kiss1ARH neurons, rendering them unresponsive to kisspeptin (d’Anglemont de Tassigny et al., 2008; Navarro V. M. et al., 2009; Wakabayashi et al., 2010). Using optogenetics and whole-cell recordings we demonstrated that high-frequency photoactivation of Kiss1ARH neurons induces a NKB-mediated slow excitatory postsynaptic potential (EPSP), which is mediated by the recruitment of canonical transient receptor potential (TRCP5) channels. The release of NKB is limited by co-released dynorphin, which acts presynaptically to inhibit further release. Together the two peptides cause synchronized firing of Kiss1ARH neurons (Kelly et al., 2018; Qiu J. et al., 2016; Qiu Jian et al., 2021), whereas kisspeptin and glutamate appear to be the main output signals from Kiss1ARH neurons (Qiu J. et al., 2016; Qiu J. et al., 2018) (Voliotis et al., 2021) (Liu et al., 2021).

Although single action potential-generated calcium influx is sufficient to trigger the release of classical neurotransmitters such as glutamate, high frequency (10-20 Hz) is required for the release of neuropeptides such as kisspeptin, NKB and dynorphin (Qiu J. et al., 2016). Indeed, the slow EPSP, which underlies the synchronization, is similar to the “plateau potential” that has been described in hippocampal and cortical neurons (Arboit et al., 2020; Zhang Z. et al., 2011). Many neurons, including Kiss1ARH neurons, express the biophysical properties that allow them to continue to persistently fire even after a triggering synaptic event has subsided (Zylberberg and Strowbridge, 2017) (Qiu J. et al., 2016). Moreover, the intrinsic bi-stability of neurons that generates persistent firing activity has been linked to a calcium-activated, non-selective cation current (ICAN) (Zylberberg and Strowbridge, 2017), and TRPC channels, specifically TRPC5 channels, are thought to be responsible for the ICAN in cortical neurons (Zhang Z. et al., 2011). Therefore, we postulate that TacR3 activation via NKB drives influx of Ca+2 through TRPC5 channels leading to greater build-up of [Ca2+]i that facilitates the opening of more TRPC5 channels in a self-sustaining manner. Indeed, using the fast intracellular calcium chelator BAPTA, which has been shown to robustly inhibit TRPC5 channel activation in heterologous cells (Blair et al., 2009), we have been able to abolish the slow EPSP and persistent firing in Kiss1ARH neurons following optogenetic stimulation in female mice (Qiu Jian et al., 2021).

Although the expression of peptides in Kiss1ARH neurons are downregulated by high circulating levels (late follicular levels) of E2, the intrinsic excitability of and the glutamate release by Kiss1ARH neurons are increased by Cacna1g (Cav3.1, T-type calcium channel), Hcn1 and Hcn2 (Hyperpolarization-activated, Cyclic Nucleotide Gated channels) mRNA expression and Vglut2 mRNA expression, respectively (Qiu J. et al., 2018). Burst firing in CNS neurons, which efficiently releases fast amino acid transmitters like glutamate, is generated primarily by the T-type calcium channel current (IT) (e.g., in thalamic relay neurons), and the rhythmicity of this burst firing is dependent on the h-current (Ih) (for review, see (Lüthi and McCormick, 1998; Zagotta and Siegelbaum, 1996)). Ih is mediated by the HCN channel family, which includes channel subtypes 1-4, of which Hcn1 and Hcn2 are the main channels in Kiss1ARH neurons. Ih depolarizes neurons from hyperpolarized states, raising the membrane potential into the range of IT activation (Erickson et al., 1993a, 1993b; Kelly and Rønnekleiv, 1994; Lüthi and McCormick, 1998; Zhang C. et al., 2009). IT is mediated by the low-threshold voltage-gated calcium channels, CaV3.1-3.3 (for review see (Perez-Reyes, 2003)). IT initiates a transient Ca2+-driven depolarization above the threshold for action potential initiation (i.e., a low threshold spike) (Llinás, 1988; Tsien et al., 1987). This depolarization then drives neurons to fire an ensemble (burst) of Na+-driven action potentials.

Based on the above compelling evidence we postulated that Kiss1ARH neurons transition from peptidergic neurotransmission, driving the pulsatile release of GnRH via kisspeptin release into the median eminence, to glutamatergic transmission that facilitates in the preovulatory surge of GnRH (Lin et al., 2021) through their projection to the Kiss1AVPV neurons (Qiu J. et al., 2016). Therefore, we initiated studies to thoroughly characterize effects of high circulating (late follicular) levels of E2 on the expression of the full complement of voltage-activated calcium channels (and currents) and the opposing K+ channels, involved in the repolarization, on the excitability of Kiss1ARH neurons. We performed whole-cell recordings and single cell RT-PCR analysis of Kiss1ARH neurons to determine which channels are involved in the physiological transition from peptidergic to glutamatergic neurotransmission. Our physiological findings were incorporated into a mathematical model that accounts for the E2 effects on the firing activity of Kiss1ARH neurons and validates our hypothesis that high levels of E2 facilitate the transition from peptidergic to glutamatergic neurotransmission.

Results

Whole-cell current of voltage-activated Ca2+ channels in Kiss1ARH neurons

Previously, we have shown that an increase in the intracellular calcium concentration can potentiate TRPC5 channel current in POMC neurons (Qiu J. et al., 2010). Additionally, chelating intracellular calcium with BAPTA abolishes the slow excitatory postsynaptic potential (EPSP) and persistent firing in Kiss1ARH neurons (Qiu Jian et al., 2021). Here, to investigate the contributions of voltage-activated calcium channels (VGCCs) to the increase in intracellular calcium, we measured the peak calcium current contributed by both the low and high voltage-activated calcium channels. To assess VGCC activity, we employed 150-ms test pulses starting from a holding potential of -80 mV with 10-mV increments, ultimately reaching a test potential of +40 mV in Kiss1ARH neurons. The inward currents evoked by the voltage pulses were identified as calcium (Ca2+) currents, as they were blocked by the universal VGCC inhibitor Cd2+ (200 μM) (McNally et al., 2020) (Figures 2A-E). The maximum total inward and Cd2+-sensitive currents reached their peak amplitudes at -10 mV. To differentiate between various calcium channel subtypes present in Kiss1ARH neurons, we applied selective antagonists individually, allowing us to isolate the drug-sensitive current for each cell. When we individually applied specific antagonists, we observed partial inhibition of the Ca2+ currents. Treatment with 10 μM nifedipine (Hiraizumi et al., 2008; Kato et al., 2003; Lee et al., 2002), an L-type Ca2+ channel inhibitor, resulted in a partial inhibition (26.0%) of the whole-cell calcium current. Additionally, application of 1 μM ω-conotoxin MVIIC (ConoMVIIC, targeting N/P/Q-type channels) (Nunemaker et al., 2003), 2 μM ω-conotoxin GVIA (conoGVIA, targeting N-type channels) (Lee et al., 2002), 200 nM ω-agatoxin IVA (AgaIVA, targeting P/Q-type channels) (Kato et al., 2003; McNally et al., 2020), 100 nM SNX-482 (targeting R-type channels) (Hiraizumi et al., 2008), or 1 μM TTA-P2 (TTAP2, targeting T-type channels) (McNally et al., 2020) also led to partial inhibition of the Ca2+ currents (Figure 2F). The observed reduction in current with each inhibitor indicates the presence of all the major subtypes of Ca2+ currents in Kiss1ARH neurons. Among the inhibitors used, the largest components of the whole-cell calcium current was found to be sensitive to nifedipine (26.1%), conoGVIA (25.1%), and SNX-482 (31.1%) (Figures 2A, B, D and 2F). Subsequently, we documented contributions from TTA-P2-sensitive channels, accounting for approximately 6.7% of the total Ca2+ current, and AgaIVA-sensitive channels, which constituted approximately 3.9% (Figures 2E, C). These findings indicate that high voltage-activated L-, N-, and R-type channels constitute the largest components of the voltage-activated Ca2+ current in Kiss1ARH neurons that not only increase the overall excitability but also greatly facilitate TRPC5 channel opening (Blair et al., 2009), which is the major downstream target of TACR3 activation by NKB (Qiu J. et al., 2016).

Relative contribution of voltage-gated calcium currents in Kiss1ARH neurons from OVX mice.

A-E, representative current –voltage relationships showing that Cd2+ (non-selective blocker of calcium channels) sensitive peak currents were inhibited by different calcium channel blockers: A, nifedipine; B, ω-conotoxin GIVA; C, ω-agatoxin IVA; D, SNX-482; E, TTA-P2. F, the maximum peak currents were measured at -10 mV. The proportions of Ca2+ currents inhibited by nifidipine (L type), ω-conotoxin GVIA (N type), ω-agatoxin IVA (P/Q), SNX-482 (R type) and TTA-P2 (T type). Data are expressed as mean ± SEM, n = cell numbers.

Voltage-activated Ca2+ channels contribute to generation of slow-EPSP in Kiss1ARH neurons

Our previous study utilizing optogenetics demonstrated that high-frequency photostimulation of Kiss1ARH neurons releases NKB. This release of NKB induces slow excitatory postsynaptic potentials (EPSPs) and facilitates the recruitment of other Kiss1ARH neurons, resulting in synchronous firing of the Kiss1ARH neuronal population (Qiu J. et al., 2016). Additionally, chelating intracellular calcium with the fast chelator BAPTA abolishes the slow EPSP and persistent firing in Kiss1ARH neurons, highlighting the role of calcium signaling in these processes (Qiu Jian et al., 2021). To access the involvement of HVA channels in the generation of the slow EPSP, we conducted experiments where we blocked the L-type Ca2+ channels with nifedipine (10 µM) and the N- and P/Q-type Ca2+ channels with ω-conotoxin MVIIC (1 µM). We then measured the slow EPSP. Indeed, both nifedipine and ω-conotoxin MVIIC significantly inhibited the slow EPSP by 42.9% and 60.4%, respectively (Figure 3). In addition, we used SNX (100 nM), which selectively blocks R-type Ca2+ channels, and the slow EPSP was reduced to 28.7% of its control value (Figure 3C). The selective T-channel blocker TTA-P2 (5 µM) inhibited the slow EPSP by 28.6% (data not shown). Therefore, it appears that all of the calcium channels contribute to maintaining the sustained depolarization underlying the slow EPSP.

Blockade of HVA Ca2+ channels decreases the slow EPSP in Kiss1ARH neurons.

A-C, representative traces showing that the slow EPSPs were abolished by perfusing the blocker of the L-type calcium channel, nifedipine (A) or N- and P/Q -type calcium channels, ω-conotoxin MVIIC (B), or the R-type calcium channel, SNX 482 (C), respectively. The arrows indicate the measurements of slow EPSP amplitude, denoted as R1 and R2, after low-pass filtering. D. Bar graphs summarizing the effects of drugs on the R2/R1 ratios. The slow EPSP was generated in OVX Kiss1-Cre::Ai32 mice. Comparisons between different treatments were performed using a one-way ANOVA analysis (F (3, 44) = 19.72, p<0.0001) with the Bonferroni’s post hoc test. **, **** indicates p<0.01, 0.001, respectively vs. control.

E2 increases the mRNA expression and the whole-cell current of voltage-activated Ca2+ channels

The neuropeptides NKB (tachykinin2, Tac2) and kisspeptin (Kiss1), which are expressed in Kiss1ARH neurons, are crucial for the pulsatile release of gonadotropin-releasing hormone (GnRH) and reproductive processes. E2 decreases the expression of Kiss1 and Tac2 mRNA in Kiss1ARH neurons but enhances the excitability of Kiss1ARH neurons by amplifying the expression of Cacna1g, Hcn1, and Hcn2 mRNA, as well as increasing T-type calcium currents and h-currents (Qiu, 2018, 20613}. Moreover, E2 drives Slc17a6 mRNA expression and enhances glutamatergic synaptic input to arcuate neurons and Kiss1AVPV neurons (Qiu J. et al., 2018). As a result, the E2-driven increase in Kiss1ARH neuronal excitability and glutamate neurotransmission may play a crucial role in triggering the surge of GnRH, ultimately leading to the LH surge.

To assess the impact of E2 on the modulation of voltage-activated calcium channels and Kiss1ARH neuronal excitability, we employed real-time PCR (qPCR) to measure the relative expression levels of ion channel subtypes in Kiss1ARH neurons. We compared the expression in E2-treated females to those treated with oil, using the specific primers listed in Table 1. The quantification was conducted on pools of 5 or 10 neurons, as indicated in the Methods. In both oil- and E2-treated females, we quantified the expression of Cav1.2 (L-type), Cav2.1 (P/Q-type), Cav2.2 (N-type) and Cav2.3 (R-type) mRNAs. Remarkably, all of these mRNA transcripts exhibited increased expression levels in response to E2 treatment (Figure 4A and B). Congruent with our previous findings (Qiu J. et al., 2018), we observed that the mRNA expression of Cav3.1, Hcn1, and Hcn2 was also upregulated in response to E2 treatment (Figure 4C). These results suggested that E2 has a regulatory effect on the expression and function of all of these ion channels in Kiss1ARH neurons. To determine whether the increased mRNA expression translated into functional changes at the cellular level, we measured the whole-cell calcium current in Kiss1ARH neurons obtained from ovariectomized mice treated with either vehicle or E2. We discovered that E2 treatment led to a significant increase in the peak calcium current density in Kiss1ARH neurons, which was recapitulated as predicted by our computational modeling (Figure 5A-D). These findings indicate that the upregulation of the mRNA expression of calcium channels by E2 translated to an augmented peak calcium current in Kiss1ARH neurons.

A, E2 increases the expression of low and high voltage-activated calcium channels in Kiss1ARH neurons.

Kiss1ARH neurons (three to four 10-cell pools) were harvested from each of 5 vehicle- and 5 E2-treated, OVX females to quantify ion channel mRNA expression of low and high voltage activated calcium channels as described in the Methods. The analysis included: T-type (Cav3.1) low voltage-activated, as well as the following high-voltage activated channels: R-type (Cav 2.3), L-type (Cav 1.2), N-type (Cav 2.2) and P/Q-type (Cav 2.1) calcium channels. Interestingly, all of these channels were upregulated with E2 treatment, which significantly increased the whole-cell calcium current (see Figure 5). B, E2 also increased the expression of hyperpolarization-activated, cyclic-nucleotide gated HCN1 and HCN2 channels in Kiss1ARH neurons. The same Kiss1ARH neuronal pools were also analyzed for mRNA expression of HCN1 and HCN2 ion channels. HCN1 channel mRNA expression was the most highly upregulated by E2 treatment in Kiss1ARH neurons; although, the mRNA expression of HCN2 was also significantly increased. The expression values were calculated via the ΔΔCT method, normalized to GAPDH and relative to the oil control values. Bar graphs represent the mean ± SEM. * p < 0.05, **p < 0.01, ***p < 0.005, ****p < 0.001, oil versus E2. C, in addition, we have shown previously that E2-treatment also increases the associated T- and h-currents (not shown), as well as the neuronal excitability (measured as rebound excitation) in Kiss1ARH neurons. The left panel illustrates an example of rebound burst firing in Kiss1ARH neurons, while the right panel displays bar graphs representing the mean ± SEM. ***p < 0.005 (Qiu J. et al., 2018).

E2 treatment (positive-feedback regimen) increases the Ca2+ currents in Kiss1ARH neurons.

A-B, Ca2+ currents in Kiss1ARH neurons with the same membrane capacitance from oil-treated (A) or E2-treated (B) animals. C, the maximum peak currents were measured at -10 mV. The current amplitudes were normalized to the cell capacitance in all cases to calculate current density. The bar graphs summarized the density of Ca2+ current in Kiss1ARH neurons from oil-treated and E2-treated animals. The mean density was significantly greater in E2-treated (13.4 ± 0.9 pA/pF, n = 11) than in oil-treated OVX females (7.2 ± 0.5 pA/pF, n = 40) (unpaired two-tailed t-test, t(49) = 5.75, ****p < 0.0001). D. The modeling predicts that E2-treated, OVX females exhibit a significantly greater inward Ca2+ current (red trace) than the vehicle-treated females (black trace). The green arrow in the red trace indicates the T-channel “inflection.” E. Relative contribution of voltage-gated calcium currents in Kiss1ARH neurons from OVX, E2-treated mice. The maximum peak currents were measured at -10 mV. The proportions of Ca2+ currents inhibited by nifidipine (L type), ω-Conotoxin GVIA (N type), ω-agatoxin IVA (P/Q), SNX-482 (R type) and TTA-P2 (T type). Data are expressed as mean ± SEM, n = cell numbers.

The largest components of the calcium currents in Kiss1ARH neurons from the E2-treated, ovariectomized females were found to be sensitive to nifedipine, conoGVIA, and SNX-482, accounting for approximately 24.9%, 24.6%, and 27.0% of the total current across cells, respectively, which is very similar to their contributions to the whole-cell current from vehicle-treated, ovariectomized females (Figure 3). In addition, contributions from TTA-P2-sensitive channels accounted for approximately 11.1% of the total Ca2+ current, while agaIVA-sensitive channels contributed to approximately 11.0% (Figure 5E). These results highlight the prevalence of L-, N-, and R-type calcium channels as the major contributors to the whole-cell calcium current in Kiss1ARH neurons from E2-treated, ovariectomized females, but also the involvement of T-type and P/Q-type channels.

E2 does not alter the kinetics of calcium channel activation or de-inactivation

In order to determine whether the increase in peak voltage-activated calcium current density was the result of E2 regulating calcium channel kinetics, mRNA expression or both, we examined the voltage dependence of activation and inactivation. By measuring the voltage dependence of activation, we assessed how E2 affects the ability of calcium channels to open in response to membrane potential changes. Similarly, by examining the voltage dependence of inactivation, we determined how E2 influences the inactivation kinetics of calcium channels. Based on our results, there was no difference in the voltage dependence of activation between cells from the vehicle-treated control group (V1/2 = -32.3 ± 2.1 mV; n = 13) and cells from estrogen-treated females (V1/2 = -33.6 ± 2.5 mV; n = 11). Similarly, there was not a significant difference in the voltage dependence of inactivation between control cells (V1/2 = -48.9 ± 4.8 mV; n = 6) and estrogen-treated cells (V1/2 = -44.1 ± 1.9 mV; n = 5) (Figure 6). Therefore, although E2 increased the mRNA expression of HVA calcium channels, it did not affect the channel kinetics in Kiss1ARH neurons. Furthermore, our previous studies established that there is no difference in the voltage dependence of activation and inactivation of T-type calcium channels in hypothalamic arcuate neurons between the vehicle-treated and E2-treated, ovariectomized females (Qiu J. et al., 2006). Also, E2 downregulated the expression of Kcnd2 mRNA encoding Kv4.2, which is expressed in Kiss1ARH neurons (Mendonça et al., 2018) and has similar kinetics of activation as the T-type calcium channels (Oil-treated, ovariectomized females relative mRNA expression: 1.053, n = 5 animals versus E2-treated expression: 0.5643, n=5; t-test p = 0.0061). This opposing K+ current would dampen the inward calcium current. Therefore, it appears that E2 does not modulate calcium channel kinetics directly but rather alters the mRNA expression to increase the conductance.

Voltage dependence of ICa in Kiss1ARH neurons from OVX and OVX+E2 mice.

A-B, top panels: activation and inactivation protocol. Bottom: representative traces. C-D, the mean V1/2 values for calcium channel activation were not significantly different for cells from controls versus cells from estrogen-treated females. Similarly, the V1/2 values for channel steady-state inactivation were similar for both groups.

BK, SK and KCNQ channels are involved in modulating excitability of Kiss1ARH neurons

In the brain, calcium plays a crucial role in sculpting neuronal firing by activating potassium channels, which subsequently influence neuronal behavior (Nicoll, 1988; Storm, 1990). Since HVA and LVA calcium channels were expressed in Kiss1ARH neurons, all of which contribute to the elevation of intracellular calcium concentration ([Ca2+]i) that facilitates TRPC5 channel opening (Blair et al., 2009), our next step involved measuring the changes in Ca2+-activated K+ channel conductances and assessing their mRNA expression. In various cell types increases in cytosolic calcium levels, whether resulting from extracellular influx or intracellular release, lead to the activation of plasma membrane calcium-dependent potassium channels (Sah Pankaj and Louise Faber, 2002). Similarly, in Kiss1ARH neurons, these channels would be activated by calcium influx through all four types of high voltage-gated calcium channels, as well as the low voltage-activated calcium channel, which are all active during action potential firing. The activity of Ca2+-activated K+ channels play a crucial role in numerous physiological processes, including secretion and the regulation of neuronal firing properties. Two main families of Ca2+-activated K+ channel channels have been characterized, distinguished by their biophysical and pharmacological properties. These families are known as BK (Big Conductance K+) and SK (Small Conductance K+) channels in the CNS (Kshatri et al., 2018). BK channels are known for their high potassium selectivity and large single channel conductance, typically ranging from 100 to 300 pS. Activation of BK channels requires both calcium binding and membrane depolarization (Blatz and Magleby, 1987; Marty, 1989; Sah P., 1996; Storm, 1990). On the other hand, SK channels are simply activated by increases in cytosolic calcium levels, with their half-maximal activation at 0.3 µM (Bond et al., 1999).

To investigate K+ currents, the cells were maintained at a holding potential of -70 mV while being exposed to blockers CNQX, AP5, picrotoxin and TTX. Subsequently, the membrane potential was stepped by depolarizing voltages, ranging from -60 mV to +40 mV in 10 mV increments, for a duration of 500 ms (Brereton et al., 2013). This protocol was employed to activate K+ currents (Figure 7A). First, we examined SK currents. The mean current density was determined at the end of the voltage pulses. In vehicle-treated, OVX females the application of the SK channel blocker apamin (100 nM) (Spergel, 2007) led to a significant reduction in whole-cell currents in the +20 to +40 mV range (Figures 7A, B). The mean outward current density at +40 mV in the control group was 125.5 ± 13.1 pA/pF (n = 3) with the apamin-sensitive component contributing 56.2 ± 2.3 pA/pF (n = 3) (Figures 7A, C). In contrast, in the E2-treated females, the overall mean outward current density at +40 mV was 191.8 ± 17.4 pA/pF (n = 5), which was significantly greater than the vehicle control group (Figures 7D, E). However, there was no significant difference in the apamin-sensitive component between the vehicle-treated and E2-treated females, 56.2 ± 2.3 pA/pF versus 63.8 ± 5.5 pA/pF (n = 4), respectively (Figure 7F). Our computational model was calibrated so that SK channels contributed ∼50 pA/pF to the whole-cell outward K+ current in E2-treated females (Figure 7H).

Small conductance, calcium-activated K+ (SK) channel is involved in repolarization of burst firing Kiss1ARH neurons in OVX and E2-treated, OVX mice.

A. Representative traces of the inhibition of outward currents before (left, control) and after the specific SK blocker Apamin (500 nM, middle). Apamin sensitive currents were calculated from the subtraction of control and apamin at depolarized potentials (right). Cells were clamped at -70 mV and given 500 ms voltage pulses from - 60 mV to +40 mV in 10 mV steps at 0.2 Hz, as shown in A at the bottom. B. Mean current density-voltage relationships measured at the end of the 500 ms voltage step ranging from -60 mV to +40 mV were obtained in the absence and presence of apamin (two-way ANOVA: main effect of treatment (F(1, 4) = 7.697, p = 0.0501), main effect of time (F(10, 40) = 99.3, p < 0.0001) and interaction (F(10, 40) = 7.645, p < 0.0001); mean ± SEM; n = 3; post hoc Bonferroni test, **p < 0.01, ***p < 0.005, ****p < 0.001). C. Apamin sensitive current densities were obtained from C (mean ± SEM, n = 3). D. Representative traces of the inhibition of outward currents before (left, control) and after the specific SK blocker Apamin (500 nM, middle). Apamin sensitive currents were resulted from the subtraction of control and apamin at depolarized potentials (right). E. Mean current density-voltage relationships measured at the end of the 500 ms voltage step ranging from -60 mV to +40 mV were obtained in the absence and presence of apamin (two-way ANOVA: main effect of treatment (F(1, 8) = 9.433, p = 0.0153), main effect of time (F(10, 80) = 184.9, p < 0.0001) and interaction (F(10, 80) = 8.791, p < 0.0001); mean ± SEM, n = 4; post hoc Bonferroni test, *p < 0.05, ***p < 0.005, ****p < 0.001). F. Apamin sensitive current densities were obtained from C and E (ns; Two-way ANOVA followed by Bonferroni post hoc test; mean ± SEM; OVX, n = 3; OVX+E2, n = 4). G. Kiss1ARH neurons (three to four 10-cell pools) were harvested from each of 5 vehicle- and 5 E2-treated, OVX females to quantify the mRNA expression of SK3 ion channel. E2 did not increase the mRNA expression small conductance calcium-activated K+ (SK3) channels in Kiss1ARH. The expression values were calculated via the ΔΔCT method, normalized to GAPDH and relative to the oil control values. Bar graphs represent the mean ± SEM (unpaired two-tailed t-test for SK3, ns). H. The mathematical model was calibrated on the electrophysiology data from Kiss1ARH neurons for E2-treated females before and after treatment with the specific SK blocker apamin, left panel versus middle panel respectively (see Table S1 for gSK). The modeled apamin-sensitive current (right panel) matches the electrophysiological data. For the calibration it was assumed that the applied concentration of apamin (500nM) completely blocked the SK current.

Furthermore, to investigate the expression of the mRNAs encoding SK channel subunits in Kiss1ARH neurons from vehicle-treated and E2-treated OVX females, qPCR experiments were performed on 10-cell Kiss1ARH neuronal pools (Figure 7G). We focused on SK3 channels because these channels exhibit the highest expression in the hypothalamus and E2 regulates their expression (Bosch et al., 2002). E2 treatment had no effect on the mRNA expression of the SK3 subunit. These findings support our electrophysiology results.

Additionally, following the same protocol and in the presence of the same cocktail of blockers (CNQX, AP5, picrotoxin and TTX), we investigated the contribution of BK channels to Kiss1ARH neuronal excitability. In the OVX females, the application of the BK channel blocker iberiotoxin (ibTx, 200 nM) (Niday and Bean, 2021) resulted in only a slight attenuation of the outward current (n = 5) (Figures 8A, B). The ibTx-sensitive current density measured at +40 mV was 31.1 ± 8.4 pA/pF (Figures 8A, C). However, in the E2-treated females, the application of ibTx significantly attenuated the whole-cell K+ current from +30 to +40 mV, (Figures 8D, E). Additionally, the ibTx-sensitive current was significantly larger in the +0 to +40 mV range in the E2-treated females compared to the OVX females (Figure 8F). These findings indicate that E2 treatment modulates the activity of ibTx-sensitive BK current in Kiss1ARH neurons, resulting in increased current density (100.9 ± 11.7 pA/pF versus 31.1 ± 8.4 pA/pF at +40 mV). Hence our computational model was calibrated so that BK channels contributed ∼100 pA/pF to the whole cell outward K+ current in the E2-treated females (Figures 8H).

Large conductance, calcium-activated K+ (BK) channels contributes to the repolarization of Kiss1ARH neurons in OVX and E2-treated, OVX mice.

A. Representative traces of the inhibition of outward currents before (left, control) and after the specific BK blocker iberiotoxin (IbTx; 200 nM, middle). IbTx sensitive currents were calculated from the subtraction of control and IbTx at depolarized potentials (right). Cells were clamped at -70 mV and given 500 ms voltage pulses from -60 mV to +40 mV in 10 mV steps at 0.2 Hz, as shown in A at the bottom. B. Mean current density-voltage relationships measured at the end of the 500 ms voltage step ranging from -60 mV to +40 mV were obtained in the absence and presence of IbTx (two-way ANOVA: main effect of treatment (F(1, 8) = 0.8841, p = 0.3746), main effect of time (F(10, 80) = 71.56), p < 0.0001) and interaction (F(10, 80) = 1.127, p = 0.3528); mean ± SEM, n = 5; post hoc Bonferroni test, p > 0.05). C. IbTX sensitive current densities were obtained from B (mean ± SEM, n = 5). D. Representative traces of the inhibition of outward currents before (left,control) and after the specific BK blocker iberiotoxin (IbTx; 200 nM, middle). IbTx sensitive currents were resulted from the subtraction of control and IbTx at depolarized potentials (right). E. Mean current density-voltage relationships measured at the end of the 500 ms voltage step ranging from -60 mV to +40 mV were obtained in the absence and presence of IbTX (two-way ANOVA: main effect of treatment (F(1, 6) = 3.181, p = 0.1248), main effect of time (F(10, 60) = 52.90, p < 0.0001) and interaction (F(10, 60) = 3.667, p = 0.0007); mean ± SEM, n = 4; post hoc Bonferroni test, *p < 0.05, **p < 0.01). F. IbTx sensitive current densities were obtained from C and E (two-way ANOVA: main effect of treatment (F(1, 7) = 31.63, p = 0.0008), main effect of time (F(10, 70) = 80.41, p < 0.0001) and interaction (F(10, 70) = 21.54, p <0.0001); mean ± SEM, OVX, n = 5; OVX+E2, n = 4; Bonferroni post hoc test, **p < 0.01, ****p < 0.001). G. Kiss1ARH neurons (three to four 10-cell pools) were harvested from each of 5 vehicle- and 5 E2-treated, OVX females to quantify the mRNA expression of BKα channel. E2-treatment increased the mRNA expression of BKα. The expression values were calculated via the ΔΔCT method, normalized to GAPDH and relative to the oil control values. Bar graphs represent the mean ± SEM (unpaired two-tailed t-test for BK, t(6) = 3.479, **p < 0.01). H. The mathematical model was calibrated to reproduce the current voltage relationship observed in Kiss1ARH neurons from E2-treated animals (see Table S1 for gBK) before and after treatment with IbTx. The modeled IbTx -sensitive current (right panel) matches the electrophysiological data. For the calibration it was assumed that the applied concentration of IbTx (200nM) completely blocked the BK current.

To investigate the expression of mRNA encoding BK channel subunits in Kiss1ARH neurons from vehicle-treated and E2-treated OVX females, qPCR experiments were performed on 10-cell Kiss1ARH neuronal pools (Figure 8G). E2 treatment significantly increased the mRNA expression of the BKα1 (Kcnma1) subunit. These findings support our electrophysiological findings that there is a significant increase in BK channel activity in Kiss1ARH neurons with E2 treatment (Figure 8F). In addition, E2 increased the mRNA expression of Kcnb1 encoding Kv2.1 (E2-treated relative mRNA expression: 1.672, n = 5 versus oil-treated mRNA expression: 1.086, n = 5; t-test p-value = 0.0024). The combination of the up-regulation of the two of these K+ channels would facilitate rapid repolarization of Kiss1ARH following an action potential.

Traditionally, the after hyperpolarization is divided into three distinct phases: fast (fAHP), medium (mAHP), and slow after hyperpolarization (sAHP) (Storm, 1990; Vogalis et al., 2003). The fast after hyperpolarization (fAHP) is primarily mediated by the BK family of potassium channels (Storm, 1987). The medium after hyperpolarization (mAHP) is predominantly mediated by apamin-sensitive SK2 channels (Bond et al., 2004; Peters et al., 2005). However, KCNQ family members contribute to both the mAHP and sAHP (Tzingounis A. V. et al., 2010; Tzingounis A. V. and Nicoll, 2008). Therefore, to investigate the contribution of KCNQ channels to Kiss1ARH neuronal excitability, voltage clamp experiments were conducted in the presence of TTX, CNQX, AP5, and picrotoxin, and a standard M-current protocol was run using the M-channel blocker XE-991 to isolate the M-current (Greene et al., 2017; Roepke et al., 2011) (Figure 9 A, B). The application of XE-991 resulted in the inhibition of M-current within a physiologically relevant voltage range of -60 to -30 mV in E2-treated OVX females but exhibited minimal impact in OVX females (Figure 9 C, D). Although the XE991-sensitive current was relatively small compared to other voltage-activated K+ conductances, it demonstrated a significant increase in E2-treated, OVX females (Figures 9 E). The maximum peak current density sensitive to XE-991 at -30 mV was found to be four times higher in E2-treated OVX females when compared to OVX females. This would contribute to the repolarization following burst firing. Furthermore, E2 increased the mRNA expression of Kcnq2, (Figure 9F), which suggests that KCNQ channels play a key role in repolarizing Kiss1ARH neurons following burst firing. Indeed, our modeling predicted that M-current contributed to the repolarization following burst firing (Figure 9G).

KCNQ channels (M-current) contribute to the slow AHP in Kiss1ARH neurons.

A-B. Representative current traces of the M-current inhibition caused by 40 µM XE-991 perfused for 10 min in (A) OVX-oil and (B) OVX+E2-treated female mice. Inset: M-current deactivation protocol. C-D. Current density-voltage plots from –75 to –30 mV of vehicle and XE-991 perfusion in (C) OVX-oil and (D) OVX+E2-treated mice. Two-way ANOVA for C: main effect of treatment (F(1, 17) = 1.908, p = 0.1851), main effect of time (F(9, 153) = 187.1, p < 0.0001), and interaction (F(9, 153) = 3.901, p = 0.0002); Veh, n = 11; XE-991, n = 8; Bonferroni post hoc test, p > 0.05. For D: main effect of Veh and XE-991 (F(1, 24) = 24.92, p < 0.0001), main effect of time (F(9, 216) = 174.5, p < 0.0001), and interaction (F(9, 216) = 52.75, p < 0.0001); Veh, n = 13; XE-991, n = 13; Bonferroni post hoc test, a = p < 0.05, b = p < 0.001. E. Treatment with E2 elevated, while XE-991 diminished the maximum peak current density elicited by a -30 mV step in OVX- and OVX+E2-treated mice. Two-way ANOVA: main effect of Veh and XE-991 (F(1, 41) = 47.59, p < 0.0001), main effect of OVX and OVX+E2 (F(1, 41) = 15.76, p = 0.0003), and interaction (F(1, 41) = 18.2, p = 0.0001; Veh, n = 11; XE-991, n = 8; Bonferroni post hoc test, Veh: OVX vs. OVX+E2, a = p < 0.001. XE-991: OVX vs. OVX+E2, p > 0.05. F. Kiss1ARH neurons (three to four 10-cell pools) were harvested from each of 5 vehicle- and 5 E2-treated, OVX females to quantify the mRNA expression of Kcnq2. E2 treatment increased the mRNA expression of Kcnq2. Unpaired t-test, t(8) = 4.850, **p = 0.0013. G. Percent contribution of the different K+ currents to the repolarization current during burst-type firing activity in the OVX+E2 state. At each time point, the length of each color bar denotes the percent contribution of the corresponding current to the total outward current.

E2 increases Vglut2 but down regulates Tac2, Trpc5 and Girk2 mRNA expression in Kiss1ARH neurons

Based on our electrophysiological results, Kiss1ARH neurons appear to transition from peptidergic to glutamatergic neurotransmission through E2-mediated changes in the expression of voltage-activated Ca2+ channels and K+ channels, and their respective conductances. Therefore, we asked the question is there a difference in peptide and glutamate mRNA expression mediating this transition? Therefore, we ran a comparison between Tac 2 (NKB) and Vglut2 (surrogate for glutamate) expression. The cycle threshold (CT) was compared between Tac2 and Vglut2 as well as Kiss1, TRPC5 and GIRK2 in Kiss neuronal cell pools from OVX oil-treated and OVX E2-treated animals (Figures 10A, B). It is worth noting that lower number of cycles illustrate a higher quantity of mRNA expression because the fluorescence is detected earlier, and one cycle difference represents a doubling in expression. As expected, the reference gene Gapdh did not change with E2-treatment. However, quantitative PCR results revealed that E2 treatment of OVX females significantly reduced Tac2 expression (Figure 10C), whereas Vglut2 mRNA was significantly increased in Kiss1ARH neurons (Figure 10D). Moreover, both Trpc5 and Girk2 expression were significantly reduced in E2-treated, OVX females (Figures 10E, F).

Estradiol decreases Tac 2, Trpc5 and Kcnj6 but increases Vglut 2 mRNA expression in Kiss1ARH neurons.

A. qPCR amplification curves illustrating the cycle threshold (CT) for Tac2, Gapdh, Kiss1, Trpc5, Slc17a6 (Vglut2) and Kcnj6 (GIRK2) in Kiss1ARH five cell (C,D,E) or ten cell (F) neuronal pools (three to six pools from each animal) in OVX Oil-treated, and B, in OVX E2-treated females. C. Quantitative real-time PCR analysis of Tac2 mRNA (n=5 animals), D, Vglut2 (n=7 animals), E, Trpc5 (n=5 animals), F, Kcnj6 (n=5 animals). Comparisons were made between Oil-treated and E2-treated, OVX females using the comparative 2-ΔΔCT method. Bar graphs represent the mean ± SEM (Unpaired t test for Tac2, t(6)= 6.350, p<0.001; Unpaired t test for Vglut2, t(8)= 4.522, p<0.001; Unpaired t test for Trpc5, t(6)= 4.818, p<0.01; Unpaired t test for Kcnj6, t(6)= 3.457, p<0.01). The data for C and D (Tac2 and Slc17a6) has been published previously (Qiu J. et al., 2018).

CRISPR mutagenesis of Trpc5 attenuates slow EPSP and reduces excitability of Kiss1ARH neurons

Our computational modeling suggested that TRPC5 channels play a dominant role in regulating cell excitability. Therefore, as proof of principle, we utilized a CRISPR approach to mutate TRPC5 channels in Kiss1ARH neurons similar to our previous studies (Hunker et al., 2020; Stincic et al., 2021). Hunker et al. developed a single viral vector for conditional expression of the smaller Staphylococcus aureus (SaCas9) and sgRNA that yields high-efficiency mutagenesis in specific cell types (Hunker et al., 2020). To selectively mutate Trpc5 in Kiss1ARH neurons, we generated two guide RNA’s, one targeting exon 2, which is conserved across all splice variants, and the other targeting exon 7, the pore forming domain (Figures 11A, B). A cohort of Kiss1ARH mice were given bilateral stereotaxic injections into the ARH of the two AAV1-FLEX-SaCas9-sgTrpc5’s or a control virus containing the Trpc5 guide with three base pairs in the seed region mutated (SaCas9-control) as described (Hunker et al., 2020). An additional Cre-dependent virus of the same serotype (AAV1) that drove expression of a fluorophore (YFP or mCherry) was co-administered in order to visualize injection quality and facilitate harvesting of cells (Figure 11C). After three weeks, mice underwent ovariectomy since OVX mice express the maximum slow EPSP amplitude (Qiu J. et al., 2016). Brain slices were prepared, and cells harvested as previously described (Qiu J. et al., 2018) and analyzed with qPCR. We found that the Trpc5 mutagenesis group displayed a reduction in relative expression of Trpc5 in Kiss1ARH neurons compared to the control group (Figure 11D). Hence, the qPCR data verified that in the sgTrpc5-targeted mice we can selectively reduce Trpc5 gene expression in targeted cells.

CRISPR mutagenesis of Trpc5 channels in Kiss1ARH neurons.

A, structure of AAV1-FLEX-SaCas9-U6sgTrpc5-exon2. Exon 2 of Trpc5 is denoted with guide sequence highlighted in red, the PAM is underlined. B. Structure of AAV1-FLEX-SaCas9-U6sgTrpc5-exon7. Exon 7 of Trpc5 is denoted with guide sequence highlighted in red, the PAM is underlined. C1, image of coronal section through the ARH from Kiss1-Cre::Ai32 mouse with dual co-injections of AAV-DIO-mCherry and AAV1-FLEX-SaCas9-U6-sgTrpc5. Scale = 200 µm. C2, C3, higher power overlays of epifluorescence (EYFP & mCherry) images with recording pipette patched onto Kiss1ARH-Cre:mCherry cell (C3). Scale = 40 µm. D, quantitative PCR measurements of Trpc5 transcripts in double sgRNA mutagenesis of Trpc5 (second sgRNA against pore forming region) in Kiss1ARH neurons. Primers were targeted to 1st or 2nd guide, respectively.

As predicted, mutagenesis of Trpc5 in Kiss1ARH neurons significantly attenuated the slow EPSP (Figures 12A, B, C) such that the postsynaptic excitation was reduced to a “trickle” of action potential firing. What we would not have predicted is that the double sgRNA mutagenesis of Trpc5 channels in Kiss1ARH neurons significantly hyperpolarized the resting membrane potential by 7 mV (Figure 12D). Moreover, the rheobase (minimum current required to induce firing) significantly increased by ∼20% in females bearing the sgTrpc5 double mutagenesis (Figure 12E). The firing frequency versus injected current (F-I) curve for sgTrpc5 double mutagenesis Kiss1ARH neurons was also significantly attenuated (Figure 12F). In agreement with these experimental findings, simulations of our mathematical model confirmed that TPRC5 channels should lower the rheobase and greatly enhance the firing activity of Kiss1ARH neurons (Figures 12G, H). Finally, we employed our mathematical model to further investigate the transition from synchronous firing driven by NKB release and TRPC5 channel activation to burst firing generated by E2-mediated upregulation of endogenous conductances. Our simulations suggest that synchronous firing is indeed sculpted by the interplay between TRPC5 and GIRK channels, whereas burst firing is controlled by the E2-dependent increase of calcium and calcium-activated K+ conductances (Figure 13).

Double CRISPR mutagenesis of Trpc5 attenuates slow EPSP, increases rheobase and shifts the F-I curve.

A, high-frequency photo-stimulation (20 Hz) generated slow EPSP in Kiss1ARH neuron from ovariectomized, control mouse. Red trace is slow EPSP after low-pass filtering. B, slow EPSP in Kiss1ARH neuron from OVX, double sgTrpc5 -targeted mouse. C, summary of the effects of Trpc5 mutagenesis on slow EPSP amplitude in female mice (**** p < 0.0001). D, double sgRNA mutagenesis of Trpc5 channels in Kiss1ARH neurons significantly increased the RMP (control: -64.5±1.4 mV versus double sgTrpc5 1, 2: -71.1±1.2 mV, *p = 0.0007). E, current ramp showing the increased rheobase in sgTrpc5 double mutagenesis (control: 31.1±1.2 pA, n=31, versus sgTrpc5 1&2, 35.3±1.0 pA, n=33, p = 0.0073). F, firing frequency vs. current (F-I) curves for control versus sgTrpc5 double mutagenesis (*<0.05; **<0.01 and ***p < 0.005, respectively). G, model simulations of the effects of a current ramp (50 pA/sec) for OVX (left panel) and OVX female with reduced (muted) TRPC5 conductance (right panel), and H, the associated firing frequency vs. current curves. In the latter case the TRCP5 conductance was halved, which is a conservative estimation of the CRISPR state in which the Trpc5 is much more mutated in Kiss1ARH neurons (Figure 11).

Computational modeling of a Kiss1ARH neuron in the OVX and OVX + E2 state demonstrates its distinct dynamic responses.

A model of the Kiss1ARH neuron was developed and calibrated using molecular data and electrophysiological recordings of Kiss1ARH neurons from OVX and OVX+E2 mice. A. Simulations of the OVX-parameterized model demonstrating high frequency activity in response to NKB stimulation. The balance between GIRK and TRCP5 conductance controls the response of the neuron to NKB stimulation, with neuronal response eliminated when TRPC5 conductance is low (red triangle) relative to the GIRK conductance. B. The OVX+E2 parameterized models demonstrate sustained burst firing activity. The bursting activity that is supported by elevated h- and Ca 2+-currents (red square). C. In the OVX+E2 state, burst firing activity is also supported by high conductance of HVA Ca2+ channels relative to the conductance of TRPC5 channels. Representative points in the parameter space giving rise to burst firing activity are marked with red squares, whereas red triangles are used for points resulting in regular spiking. The black line separates these two regions of activity.

Discussion

We have shown that E2 plays a critical role in transitioning the glutamatergic/peptidergic Kiss1ARH neurons from a high frequency firing mode for synchronization, which is dependent on NKB-driven activation of TRPC5 channels, to a short bursting mode that would facilitate glutamate release. E2 decreased the expression of the peptide neurotransmitters NKB (kisspeptin and dynorphin) and TRPC5 channels but increased the mRNA expression of Vglut2 and voltage-activated calcium channels that contribute to burst firing and glutamate release from the Kiss1ARH neurons. We determined that the increase in mRNA expression of the HVA calcium channels translated into a significant increase in whole-cell current with all of the calcium channels contributing proportionally. Most importantly the kinetics of activation and inactivation were unaltered with E2 treatment, which indicates that other post-translation modifications were not affecting channel activity. Surprisingly and somewhat counter intuitive, the BK α1 subunit was also upregulated, but based on our modeling the rapid repolarization of the Kiss1ARH neurons (i.e., the fast AHP) facilitates higher frequency of action potential firing. Moreover, our modeling confirmed that TRPC5 channels, which generate the slow EPSP (a.k.a., plateau potential in other CNS neurons), are vital for initiating and sustaining synchronous firing of Kiss1ARH neurons, while concurrent activation of GIRK channels repolarizes Kiss1ARH neurons. E2 treatment of ovariectomized females decreased both Trpc5 and Girk2 channel mRNA expression, which in our model correlated with the reduction in sustained high frequency firing of Kiss1ARH neurons. Therefore, the synchronous high frequency firing of Kiss1ARH neurons in a low E2 milieu correlates with the pulsatile release of GnRH (LH from the pituitary gland), whereas the transition to burst firing in the presence of high circulating levels of E2 (e.g., proestrus) facilitates the GnRH (LH) surge through its glutamatergic synaptic connection with Kiss1AVPV/PeN neurons.

Core calcium conductances underlying synchronous and burst firing of Kiss1ARH neurons

TRPC5 channels are highly expressed in Kiss1ARH neurons (Figure 11), and TRPC5 channels are essentially ligand-activated calcium channels with a high permeability to calcium (PCa/PNa = 9:1) (Venkatachalam and Montell, 2007). In general, mammalian TRPC channels are activated by both G protein-coupled receptors and receptor tyrosine kinases (Ambudkar and Ong, 2007; Clapham, 2003), and are one of the major downstream effectors activated by glutamate binding to group I metabotropic glutamate receptors (mGluR1 and mGluR5) in CNS neurons (Bengtson et al., 2004; Berg et al., 2007; Faber et al., 2006; Tozzi et al., 2003). In substantia nigra dopamine neurons mGluR1 agonists induce a current that exhibits the tell-tale double-rectifying current-voltage plot of TRPC channel activation (Tozzi et al., 2003), similar to what we see with the effects of the NKB agonist senktide in Kiss1ARH neurons (Qiu Jian et al., 2021). Both mGluR1 and TacR3 are Gq-coupled to phospholipase C (PLC) activation which leads to hydrolysis of phosphatidylinositol 4,5-bisphosphate (PIP2) to diacylglycerol (DAG) and inositol 1,4,5 triphosphate (IP3), which is involved in channel activation (Birnbaumer, 2009). TacR3 (NKB) signaling has additional consequences since many K+ (e.g., GIRK, KCNQ) channels are dependent on PIP2 for channel opening, and depletion of PIP2 by PLC leads to channel closure (Brown and Passmore, 2009) (Zhang C. et al., 2013) (Whorton and MacKinnon, 2011) (Zheng et al., 2022). Therefore, depletion of PIP2 by NKB signaling would further facilitate the sustained firing of Kiss1ARH neurons during synchronization.

A plateau potential has been characterized in hippocampal and cortical neurons (Arboit et al., 2020; Zhang Z. et al., 2011) as such these neurons express biophysical properties that allow them to continue to persistently fire even after a triggering synaptic event has subsided (Zylberberg and Strowbridge, 2017). The persistent firing activity of these neurons is linked to ICAN (Zylberberg and Strowbridge, 2017), and TRPC5 channels appear to be responsible for the ICAN (Zhang Z. et al., 2011). With TacR3 activation in Kiss1ARH neurons there is an influx of Ca+2 through TRPC5 channels leading to greater build-up of [Ca2+]i that facilitates the opening of more TRPC5 channels in a self-sustaining (autocatalytic) manner (Qiu J. et al., 2016). Using the fast intracellular calcium chelator BAPTA, which has been shown to robustly inhibit TRPC5 channel activation in heterologous cells (Blair et al., 2009), we abolished the slow EPSP and persistent firing of Kiss1ARH neurons following optogenetic stimulation (Qiu Jian et al., 2021). Moreover, HVA calcium channel blockers attenuated the generation of the slow EPSP (Figure 3) so it appears that they also contribute to the ICAN since calcium influx via both LVA and HVA calcium channels can also facilitate TRPC5 channel opening in Kiss1ARH neurons. In the ovariectomized female, treatment with E2 upregulated Cav1.2, Cav2.1, Cav2.2, and Cav2.3 mRNA by 1.5-to 2-fold and Cav3.1 mRNA expression by ∼3-fold. Hence, E2 significantly increased whole-cell calcium currents Kiss1ARH neurons, which greatly enhanced the excitability and contributed to the burst firing of Kiss1ARH neurons (present findings and (Qiu J. et al., 2018)). However, the amplitude of the slow EPSP with E2 treatment is only ∼25% of the amplitude in the ovariectomized state (Qiu J. et al., 2016). Therefore, there appears to be a physiologic transition of Kiss1ARH neurons from the slow EPSP firing mode in the OVX state to the burst firing mode in the presence of E2, which has important physiologic ramifications as discussed below.

TRPC5 and GIRK channels are vital for synchronization of Kiss1ARH neurons

Recently, Tian et al. demonstrated that TRPC4, a close homolog of TRPC5 sharing ∼64 percent homology, is a “coincidence detector” of neurotransmission by both Gq/11 and Gi/o-coupled receptors in lateral septal (LS) neurons (Tian et al., 2022). In whole-cell recordings of LS neurons, TRPC 4 channels mediate a strong depolarizing plateau potential that in contrast to TRPC5 channel activation in Kiss1ARH neurons, abrogates action potential firing as a result of a depolarization block. In many instances the plateau potential in LS neurons is followed by an AHP, which is dependent on the activation of Gi/o-coupled receptors. In contrast, we have not observed an AHP in Kiss1ARH neurons following the slow EPSP (Qiu J. et al., 2016). Tian and colleagues showed that the depolarizing plateau in LS neurons is codependent on activation of both Gq/11-coupled mGluR1 glutamate receptors and Gi/o-coupled γ-aminobutyric acid type B receptors, the latter activating GIRK channels. Moreover, the firing patterns in LS neurons encodes information about the relative strengths of these contrasting inputs (i.e., Gq/11 versus Gi/o) such that only mGluR1 produces weak depolarization accompanied by increased firing of LS neurons, whereas pure GABAB receptor activation hyperpolarizes the cells and abrogates firing activity. Coincident input of both mGluR1 and GABAB receptors results in a brief burst of action potentials followed by a pause in firing, and both the pause duration and firing recovery patterns reflect the relative strengths of Gq/11 versus Gi/o inputs. Importantly, Tian and colleagues computationally simulated these various scenarios with computational modeling, and similar to our modeling, utilized only TRPC4 and GIRK channels. A notable difference between the Kiss1ARH neurons and the LS neuronal circuitry is that the GIRK channel activity is predominately at the nerve terminal of Kiss1ARH neurons, and GIRK channels are opened via dynorphin binding to kappa-opioid receptors, which does not translate into membrane hyperpolarization of the soma membrane from which we are recording (Qiu J. et al., 2016). Therefore, although the high frequency firing activity of both LS and Kiss1ARH neurons can be modeled around TRPC and GIRK channels, the generated firing patterns are dramatically different based on the timing (co-incident activation of TRPC4 and GIRK channels in LS) and localization of the GIRK channels in the axon terminal of Kiss1ARH neurons. Moreover, E2-treated, ovariectomized females show a significant down-regulation of both Trpc5 and Girk2 mRNA expression in Kiss1ARH neurons (Figure 10), which is important for the physiological transitioning as described below.

Interestingly, the hypothalamic A12 (ARH) dopamine neurons show a rhythmic “oscillatory” firing behavior that transitions to a tonic firing mode with synaptic input from Thyrotropin-releasing hormone (TRH) neurons (Lyons D.J. et al., 2010) or feedback by circulating prolactin, released by pituitary lactotrophs (Lyons D. J. et al., 2012). A more recent paper has revealed that TRPC5 channels mediate the plateau potential and tonic firing in A12 dopamine neurons in response to prolactin (Blum et al., 2019). Similar to our findings with CRISPR deletion of Trpc5 in Kiss1ARH neurons (Figures 11 and 12), conditional knockout of Trpc5 in dopamine neurons abrogated the prolactin-induced plateau potential and tonic firing. Although Blum and colleagues did not model the oscillatory firing or tonic firing of the A12 dopamine neurons, their findings are consistent with our results showing that the activation of TRPC5 channels underlies the slow EPSP (plateau potential) and sustained firing.

Contribution of endogenous K+ channels to synchronized and burst firing

Beyond the ligand-gated (e.g., baclofen) GIRK channels, there are endogenous K+ channels that help sculpt the firing activity of kisspeptin neurons. We focused on the calcium-activated K+ channel family: the large-conductance, calcium-activated potassium (BK, also called BKCa, KCa1.1, MaxiK, Slo), small conductance, calcium-activated K+ (SK1, SK2, SK3) (Bond et al., 2005), and the K+ channels underlying the M-current (KCNQ, Kv7.1-7.5) (Brown and Passmore, 2009), which mediate the fast afterhyperpolarization (AHP), the intermediate AHP/slow AHP, respectively (Andrade et al., 2012)

BK channels are gated by both voltage and cytoplasmic calcium and sculpt action potential firing in CNS neurons (Blatz and Magleby, 1987; Marty, 1989; Sah P., 1996; Storm, 1990). Indeed, BK channels have been shown to mediate rapid spike repolarization—i.e., the fast AHP in hippocampal CA1 pyramidal neurons (Lancaster and Nicoll, 1987; Storm, 1987). Blockade of BK channels in CA1 neurons attenuates the initial discharge frequency in response to current injection, which is attributable to suppression of the BK channel-dependent rapid spike repolarization (Lancaster and Nicoll, 1987; Storm, 1987). Blockade of BK channels is thought to increase inactivation of the spike-generating transient Na+ current and activate more of the slower K+ currents, thereby enhancing refractoriness and reducing excitability during the immediate aftermath of the first action potential (Shao et al., 1999). Thus, BK channels facilitate high-frequency burst firing of CA1 neurons. Furthermore, extracellular field recordings confirmed that BK channels contribute to high-frequency burst firing in response to excitatory synaptic input to distal dendrites in CA1 neurons (Gu et al., 2007). Therefore, BK channels appear to play an important role for early high-frequency, rapidly adapting firing in hippocampal CA1 pyramidal neurons, thus promoting the type of bursting that is characteristic of these cells in vivo during behavior. Based on our in vitro electrophysiological recordings and computational modeling we see a similar physiological phenomenon in Kiss1ARH neurons (Figure 8). Not only does E2 increase the mRNA expression of Kcnma1 (Figure 8G), but also the maximum BK (IbTx-sensitive) current by 4-fold. In addition, Kcnb1 mRNA was also up-regulated, and the combination of these two K+ conductances would facilitate rapid repolarization during burst firing and promote glutamate release similar to hippocampal CA1 neurons.

In contrast to BK channel expression, E2 did not affect the mRNA expression of SK3 channel mRNA. SK channels underlie the apamin-sensitive component of the medium duration AHP and are responsible for repolarization following a burst of action potentials (Andrade et al., 2012; Bond et al., 2005). The activation of SK channels is voltage-independent, but SK channels have a higher affinity for Ca2+ than BK channels (Andrade et al., 2012). SK channels are tightly coupled (within 100 nm) to L-type Ca2+ channels, and BK channels (within 30 nm) are tightly coupled to N-type Ca2+ channels in hippocampal CA1 pyramidal neurons (Marrion and Tavalin, 1998). The determination of the proximity of SK and BK channels to HVA calcium channels in kisspeptin neurons will require cell-attached patch recordings. However, in Kiss1ARH neurons the SK channels may come into play during a short burst of action potentials but would become overwhelmed with the higher frequency synchronized, sustained firing as a result of NKB stimulation (Qiu J. et al., 2016). As discussed above what limits the synchronized firing of Kiss1ARH neurons is the activation of GIRK channels.

Finally, the calcium-activated slow AHP probably plays a critical role in the repolarization of Kiss1ARH after burst firing. The molecular identification of the channels mediating the slow AHP has long been an area of intense investigation (Andrade et al., 2012; Vogalis et al., 2003). A critical feature of the slow AHP is that it activates very slowly (hundreds of milliseconds) long after the rise in cytoplasmic Ca2+ (Sah P. and Clements, 1999) so an intermediate Ca2+ signaling molecule has long been thought to be involved. Indeed, Tzingounis and colleagues (Tzingounis A.V. et al., 2007) have provided compelling evidence that the diffusible calcium sensor hippocalcin is the critical intermediate molecule involved in Ca2+ sensing. The slow AHP is abrogated in hippocalcin KO mice, and transfection of hippocalcin into cultured hippocampal neurons generates a pronounced slow AHP in response to a depolarizing stimulus (Tzingounis A.V. et al., 2007). Importantly, the slow AHP is activated by Ca2+ with an EC50 ≈ 300 nM, which is well within the operational range of hippocalcin but well below that of calmodulin (Andrade et al., 2012). Finally, two seminal papers from Tzingounis and colleagues demonstrate that KCNQ 2, 3 channels are responsible for the slow AHP in hippocampal dentate neurons (Tzingounis A. V. and Nicoll, 2008), and KCNQ 5 channels are responsible for the slow AHP in CA3 neurons (Tzingounis A. V. et al., 2010). Moreover, the KCNQ channel blocker XE991 attenuates the slow AHP in CA3 neurons (Tzingounis A. V. and Nicoll, 2008). Based on these seminal findings we investigated the role of the KCNQ channels, which “classically” underlie the M-current in Kiss1ARH. The M-current was first identified in Kiss1ARH neurons by Conde and Roepke (Conde and Roepke, 2019), and presently we found that Kcnq2 mRNA is expressed in Kiss1ARH neurons and up-regulated by E2, which translated into a greater M-current in Kiss1ARH neurons (Figure 9). Incorporating the M-current into our computational model indeed supports our hypothesis that this is a critical K+ conductance, along with SK and BK, for membrane repolarization after burst firing (Figure 9G). Importantly the slow AHP, as opposed to the fast AHP (BK) and medium AHP (SK) is highly regulated by multiple neurotransmitters (Andrade et al., 2012), which sets the stage for further modulation of the slow EPSP in Kiss1ARH neurons.

Importance of E2-driven physiological transitioning

Since the expression of the peptide neurotransmitters in Kiss1ARH neurons are down-regulated by E2, the Kiss1ARH neurons are believed to be under “inhibitory” control by E2 and are important for “negative-feedback” regulation of GnRH release (Lehman et al., 2013; Navarro V. M. et al., 2009; Smith et al., 2005) (Rance Naomi E. and Young, 1991) (Rance N.E., 2009). However, our past (Gottsch et al., 2011) and current findings document that these Kiss1ARH neurons express HVA and LVA calcium and HCN (pacemaker) channels and are excited by co-released glutamate from neighboring Kiss1ARH neurons, which indicates that these neurons have pacemaker electrophysiological properties similar to other CNS neurons (Bal and McCormick, 1993; Lüthi and McCormick, 1998). Additionally, in contrast to the neuropeptides, E2 increases Slc17a6 (Vglut2) mRNA expression, in addition to mRNA for the voltage-activated calcium and HCN channels, and increases glutamate release onto Kiss1AVPV/PeN neurons (Qiu J. et al., 2018). Interestingly, Slc17a6 mRNA expression in Kiss1ARH neurons and the probability of glutamate release are decreased along with the neuropeptides in intact versus castrated males (Nestor et al., 2016), which indicates a profound sex difference in the glutamate signaling by Kiss1ARH neurons (Nestor et al., 2016; Qiu J. et al., 2018). Obviously, in the male there is no preovulatory LH surge so there is no need for excitatory glutamatergic input to the few Kiss1AVPV neurons in the male.

In females, conditional knockout of Slc17a6 in Kiss1 neurons abrogates glutamate release from Kiss1ARH neurons (Qiu J. et al., 2018). Kiss1AVPV/PeN neurons do not express Slc17a6 and do not release glutamate. Within the Kiss1ARH neurocircuitry the lack of glutamate transmission does not diminish the slow EPSP in ovariectomized females (Qiu J. et al., 2018). Indeed, a recent publication from the Herbison lab demonstrates that glutamate generates small “synchronizing” events that are dependent on the ionotropic receptors (Han et al., 2023), but the fast neurotransmitter is unable to support the sustained firing (i.e., slow EPSP) that is necessary for peptide release and synchronization of the KNDy network. Rather, we believe that glutamate neurotransmission is more important for excitation of Kiss1AVPV/PeN neurons and facilitating the GnRH (LH) surge with high circulating levels of E2, when peptide neurotransmitters are at a nadir, but glutamate levels are high in female Kiss1ARH neurons. Indeed, low frequency (5 Hz) optogenetic stimulation of Kiss1ARH neurons, which only releases glutamate in E2-treated, ovariectomized females (Qiu J. et al., 2016), generates a surge-like increase in LH release during periods of optical stimulation (Lin et al., 2021; Voliotis et al., 2021). Therefore, there appears to be a clear role for glutamatergic transmission from the Kiss1ARH to Kiss1AVPV/PeN neurons in amplifying the LH surge in the female mouse. Finally it is important to keep in mind that even in the presence of high physiological levels of E2, the mRNA expression of Tac2 is many-fold higher than Kiss1 (Figure 10), which is essential for NKB maintaining synchronous firing of Kiss1ARH neurons, albeit at a lower frequency, across all physiological states (Qiu J. et al., 2016). Indeed, there is a progressive change from a strictly pulsatile pattern of GnRH in the hypophyseal portal circulation to one containing both pulsatile and non-pulsatile components during the development of the GnRH surge in the ewe (Evans, Dahl, Mauger, & Karsch, 1995; Evans, Dahl, Mauger, Padmanabhan, et al., 1995), and a pulsatile mode of LH secretion during the preovulatory LH surge is also evident in other species including humans (Rossmanith et al., 1990). Therefore, we believe that our cellular molecular and electrophysiological findings in combination with our computational modelling provide a foundation for understanding the complex role of Kiss1ARH neurons in controlling fertility in the mammal. Finally, our model provides the first comprehensive biophysical description of the conductances underlying the neuronal activity of KissARH neurons, which can serve as a basis for future computational modelling of the KissARH neuronal network and its interactions with other brain regions involved in the complex regulation of mammalian female reproduction.

Methods and Materials

Animals

All the animal procedures described in this study were performed in accordance with institutional guidelines based on National Institutes of Health standards and approved by the Institutional Animal Care and Use Committee at Oregon Health and Science University (OHSU) or in accordance with the United Kingdom Animals (Scientific Procedures) Act 1986 and were approved by King’s College London (KCL) Ethical Review Committee.

Mice

Kiss1Cre transgenic female mice version 2 (Padilla et al., 2018) were selectively bred at OHSU and KCL. They also were crossed with heterozygous Ai32 mice (RRID:IMSR_JAX:024109, C57BL/6 background) which carry ChR2 (H134R)–EYFP gene in their Gt(ROSA)26Sor locus (Madisen et al., 2012). All animals were maintained under controlled temperature and photoperiod (lights on at 0600h and off at 1800h at OHSU or 0700h and off at 1900h at KCL) and given free access to food (Lab Diets 5L0D) and water. Where specified, Kiss1Cre mice received viral injections to express channelrhodopsin 2 (ChR2) in Kiss1ARH neurons, fourteen to twenty-one days prior to each experiment as described (Qiu J. et al., 2016). Some of the females were ovariectomized seven days prior to an experiment. Each animal was injected on day 5 following OVX with 0.25 μg E2 or vehicle, followed on day 6 with 1.50 μg E2 or vehicle and used for experiments on day 7 (Bosch et al., 2013).

AAV delivery to Kiss1Cre mice

Fourteen to twenty-one days prior to each experiment, the Kiss1Cre mice (>60 d old) received bilateral ARH injections of a Cre-dependent adeno-associated viral (AAV; serotype 1) vector encoding mCherry (AAV1-Ef1α-DIO-mCherry) or AAV1 vectors designed to encode SaCas9 and single-guide RNAs (sgRNAs) (See the SaCas9 section for specifics on the sgRNA design). Using aseptic techniques, anesthetized female mice (1.5% isoflurane/O2) received a medial skin incision to expose the surface of the skull. The glass pipette with a beveled tip (diameter = 45 μm) was filled with mineral oil, loaded with an aliquot of AAV using a Nanoject II (Drummond Scientific). ARH injection coordinates were anteroposterior (AP): −1.20 mm, mediolateral (ML): ± 0.30 mm, dorsoventral (DV): −5.80 mm (surface of brain z = 0.0 mm); 500 nl of the AAV (2.0×1012 particles/ ml) was injected (100 nl/min) into each position, and the pipette left in place for 10 min post-injection, then slowly retracted from the brain. The skin incision was closed using Vetbond (3M) and each mouse received analgesia (Rimadyl, 4-5 mg/kg, s.c.).

Generation of AAV1-FLEX-SaCas9-U6-sgTrpc5

The generation of AAV1-FLEX-SaCas9-U6-sgTrpc5 viruses were done at the University of Washington using published methods (Gore et al., 2013; Hunker et al., 2020). The constructs of sgRNAs for Trpc5 (AAV1-FLEX-SaCas9-U6sgTrpc5) were designed to target exon2 and exon7, respectively (Figure 11A and B). To achieve Trpc5 mutagenesis in Kiss1ARH neurons, Kiss1Cre mice were co-injected with AAV1-DIO-mCherry, AAV1-FLEX-SaCas9-U6-sgTrpc5-exon2, and AAV1-FLEX-SaCas9-U6-sgTrpc5-exon7 at a ratio of 10%, 45%, and 45%, respectively. Control animals were co-injected with AAV1-FLEX-SaCas9-U6-sgRosa26 and AAV1-DIO-mCherry at a ratio of 90% and 10%, respectively. AAV1-DIO-mCherry was co-injected with Cas9 vectors to confirm the targeting of injections and visualize the infected Kiss1ARH neurons.

Visualized whole-cell patch recording

Electrophysiological and optogenetic studies were made in coronal brain slices (250 μm) containing the ARH from AAV1-EF1α-DIO-mCherry injected Kiss1Cre:GFP or Kiss1-Cre:EYFP::AI32 mice, which were vehicle-treated OVX, and E2-treated OVX females 10 weeks and older as previously described (Qiu J. et al., 2016; Qiu J. et al., 2018). Whole-cell patch recordings were performed in voltage-clamp and current-clamp as previously described (Qiu J. et al., 2018) using an Olympus BX51 W1 fixed stage scope out-fitted with epifluorescence and IR-DIC video microscopy. Patch pipettes (A-M Systems; 1.5 μm outer diameter borosilicate glass) were pulled on a Brown/Flaming puller (Sutter Instrument, model P-97) and filled with the following solution: 128 mM potassium gluconate, 10 mM NaCl, 1 mM MgCl2, 11 mM EGTA, 10 mM HEPES, 2 mM ATP, and 0.25 mM GTP adjusted to pH 7.3 with KOH; 295 mOsm. Pipette resistances ranged from 3.5–4 MΩ. In whole-cell configuration, access resistance was less than 30 MΩ; the access resistance was 80% compensated. The input resistance was calculated by measuring the slope of the I-V relationship curve between −70 and −50 mV. Standard whole-cell patch recording procedures and pharmacological testing were performed as previously described (Qiu J. et al., 2003; Qiu J. et al., 2014). Electrophysiological signals were digitized with a Digidata 1322A (Axon Instruments) and the data were analyzed using p-Clamp software (Molecular Devices, Foster City, CA). The liquid junction potential was corrected for all data analysis.

For optogenetic stimulation, a light-induced response was evoked using a light-emitting diode (LED) 470 nm blue light source controlled by a variable 2A driver (ThorLabs, Newton, NJ) with the light path directly delivered through an Olympus 40x water-immersion lens. For high-frequency (20 Hz) stimulation the length of stimulation was 10 seconds (Qiu J. et al., 2016).

For studying the activation/inactivation characteristics of the Ca2+ current, the electrodes were filled with an internal solution as described (Qiu J. et al., 2003) consisting of the following (in mM): 100 Cs+ gluconate, 20 TEA-Cl, 10 NaCl, 1 MgCl2, 10 HEPES, 11 EGTA, 1 ATP, 0.25 GTP, the pH was adjusted to 7.3 with CsOH at 300 mOsm. The bath solution as described (Zhang X. B. and Spergel, 2012) consisted of (in mM) 117.5 NaCl, 25 NaHCO3, 1.25 NaH2PO4, 10 TEA-Cl, 2 CaCl2, 1 MgCl2, 20 sucrose and 5 glucose, gassed with 95% O2 /5% CO2 (pH 7.4, 309 mOsm) and supplemented with 1 µM TTX, 10 µM CNQX, 50 µM AP5 and 100 µM picrotoxin. The effects of estrogen treatment on the peak current, peak current density, and activation/inactivation characteristics of calcium current were measured. Activation curves were fitted by the Boltzmann equation: I/Imax=1/{1+exp[V1/2-Vs)/k]}, where I is the peak current at the step potential Vs, Imax is the peak current amplitude, V1/2 is the step potential yielding half-maximum current, and k is the slope factor. Inactivation curves were fit with the Boltzmann equation: I/Imax = 1 - 1/{1 + exp [(VH - V1/2)/k]}, where I is the peak current at the step potential VH, Imax is the peak current amplitude, V1/2 is the step potential at which half the current is inactivated, and k is the slope factor.

To record M-currents, pipettes were filled with an internal solution consisting of 10 mM NaCl, 128 mM K-gluconate, 1 mM MgCl, 10 mM HEPES, 1 mM ATP, 1.1 mM EGTA, and 0.25 mM GTP (pH 7.3; 290 mOsm). During voltage-clamp, we employed a standard deactivation protocol (Conde and Roepke, 2019; Roepke et al., 2011) to measure potassium currents. This involved 500-ms voltage steps ranging from –30 to –75 mV in 5-mV increments, following a 300-millisecond prepulse to –20 mV. The amplitude of the M-current relaxation or deactivation was quantified as the difference between the initial (<10 ms) and sustained current (>475 ms) of the current trace.

The bath solution for whole-cell recording of BK, SK and M currents was aCSF supplemented with 1 µM TTX, 10 µM CNQX, 50 µM AP5 and 100 µM picrotoxin.

Electrophysiological solutions/drugs

A standard artificial cerebrospinal fluid (aCSF) was used (Qiu J. et al., 2003; Qiu J. et al., 2010). All drugs were purchased from Tocris Bioscience unless otherwise specified. 1 mM TTX (Alomone Labs), 50 mM DL-2-Amino-5-phosphonopentanoic acid sodium salt (AP5), 10 mM 6-Cyano-7-nitroquinoxaline-2,3-dione disodium (CNQX), 100 mM picrotoxin, 10 mM Nifedipine (Sigma), 2 mM ω-conotoxin GVIA (ConoGVIA; Alomone Labs), 0.5 mM ω-conotoxin MVIIC (ConoMVIIC; Alomone Labs, 50 µM ω-agatoxin IVA (AgaIVA; Alomone Labs), 50 µM SNX-482 (Alomone Labs), 10 mM TTA-P2 (TTAP2; Alomone Labs), 200 mM CdCl2 (Cd2+; Sigma), 40 mM XE991 (Alomone Labs), 100 µM Iberiotoxin (Alomone Labs), 500 µM Apamin (Alomone Labs) and 100 mM NiCl2 (Ni2+; Sigma). Stocks (1000×) were prepared in dimethylsulfoxide DMSO (picrotoxin, TTAP2) or water (TTX, AP5, CNQX, ConoGVIA, ConoMVIIC, AgaIVA, SNX-482, Cd2+, Ni2+) and stored at −20°C. Aliquots of the stock solutions were stored as appropriate until needed.

Cell harvesting of dispersed Kiss1Cre neurons and real-time quantitative PCR (qPCR)

Cell harvesting and qPCR was conducted as previously described (Bosch et al., 2013). The ARH was microdissected from basal hypothalamic coronal slices obtained from female Kiss1Cre version 2 mice (Padilla et al., 2018) (n = 5-7 animals/group). The dispersed cells were visualized, patched, and then harvested (5 or 10 cells/tube) as described previously (Bosch et al., 2013). Briefly, ARH tissue was incubated in papain (7mg/ml in oxygenated aCSF) for 50 min at 37° C then washed 4 times in low Ca2+ aCSF and two times in aCSF. For cell dispersion, Pasteur pipettes were flame polished to decreasing tip sizes and gentle trituration used to disperse the neurons onto a glass bottom dish. The plated cells were bathed in oxygenated aCSF using a peristaltic pump to keep the cells viable and clear of debris. Healthy cells with processes and a smooth cell membrane were harvested. Pipettes (World Precision Instruments; 1.5 μm outer diameter borosilicate glass) were pulled on a Brown/Flaming puller (Sutter Instrument, model P-87) to a 10 µm diameter tip. The cells were harvested using the XenoWorks Microinjector System (Sutter Instruments, Navato, CA) which provides negative pressure in the pipette and fine control to draw the cell up into the pipette. Cell pools were harvested and stored at -80°C. All cell pools were DNAse-treated using DNase1. cDNA synthesis was performed as previously described (Bosch et al., 2013).

Primers for the genes that encode for low and high voltage-gated calcium channels, TRPC5, Vglut2, large conductance calcium-activated K+ (BKα) channels, small conductance calcium-activated K+ (SK3) channels and GAPDH were designed using Clone Manager software (Sci Ed Software) to cross at least one intron-exon boundary and optimized as previously described using Power Sybr Green method (Bosch et al., 2013). We have already published primer sequences for the low voltage-activated calcium channels, Vglut2 and GAPDH (see Table 1) (Qiu J. et al., 2018). Real-time qPCR controls included neuronal pools without reverse transcriptase (-RT), hypothalamic RNA with RT (+) and without RT (-), as well as water blanks. Standard curves using ARH cDNA were utilized to determine the real-time PCR efficiency (E = 10(−1/m) – 1) (Biosystems, 2006; Pfaffl, 2001). Only primers resulting in efficiencies of 90-100% were used for analysis. Primer sequences, qPCR parameters and efficiency calculations are provided in Table 1.

mRNA expression analysis

qPCR was performed on a Quantstudio 7 Flex Real-Time PCR System (Applied Biosystems) using Power SYBR Green PCR Master Mix (Applied Biosystems) according to established protocols (Bosch et al., 2013). The comparative ΔΔCT method (Livak and Schmittgen, 2001; Pfaffl, 2001; Schmittgen and Livak, 2008) was used to determine values from duplicate samples of 4 µl for the target genes and 2 µl for the reference gene GAPDH in a 20 µl reaction volume containing 1x Power SYBR Green PCR Master Mix and 0.5 µM forward and reverse primers. Three to four 5-cell or 10-cell pools per animal were analyzed from 5-7 animals per group. The relative linear quantity was determined using the 2-ΔΔCT equation (Livak and Schmittgen, 2001; Pfaffl, 2001; Schmittgen and Livak, 2008). Relative mRNA expression level of target genes in Kiss1Cre neurons was obtained by comparing OVX Oil-treated controls to OVX E2-treated animals. The mean Δ CT for the target genes from the OVX Oil-treated control samples was used as the calibrator. The data were expressed as n-fold change in gene expression normalized to the reference gene GAPDH and relative to the calibrator.

Experimental design and Statistical analysis

For the visualized, whole-cell patch recording experiments, only one cell was recorded per slice. Two to three slices were analyzed from each Kiss1Cre mouse, with at least 3-5 mice contributing to each group. For cell harvesting of dispersed Kiss1Cre-YFP neurons and qPCR measurements, 10 cells per pool and 3-6 pools from each animal were used, unless otherwise specified. Statistical comparisons between two groups were performed using an unpaired two-tailed Student’s t-test. Comparisons between more than two groups were performed using the repeated measures, multifactorial ANOVA. If a significant interaction was encountered, we then moved to the one-way ANOVA, followed by the multiple range tests as specified in the appropriate figure legends. All data were analyzed using GraphPad Prism version 6. All data are presented as mean ± standard error of the mean (SEM). Differences were considered statistically significant if the probability of error was less than 5%.

Mathematical model and simulation of neuronal behavior

A mathematical model of the Kiss1ARH neuron was developed and calibrated based on our physiological findings. Here, we employed the Hodgkin-Huxley modelling approach (Hodgkin and Huxley, 1952). Accordingly, the equation describing the membrane potential Vm of the Kiss1ARH neuron is given by where Cm is the membrane capacitance and I is the sum of 12 ionic currents:

INaT and INaP are the transient and persistent sodium currents, respectively; IA represents the A current; IM represents the M current; ISK and IBK are the potassium currents through the SK and BK channels respectively; Ih the HCN current; IT the T-type calcium current; ICa represents other calcium currents (L-, N-, P/Q-, and R-type); ITRPC5 represents Calcium current throught the TRPC5 channel; IGIRK potassium current through the GIRK channels. Finally, Ileak represents the contribution of leak currents. We use the Hodgkin-Huxley formalism to model current dynamics and their dependence on the membrane potential. A complete specification of all currents along with the complete table of model parameters can be found in the supplementary material. Model simulations were conducted in Matlab using the built-in numerical solver (ode45; based on an explicit Runge-Kutta (4, 5) formula). The Matlab code is available at https://git.exeter.ac.uk/mv286/kiss1-arcuate-neuron-model.

eLife Assessment

This study addresses the effects of estrogen on the kisspeptin1 subset of neurons in the arcuate nucleus of the hypothalamus of female mice after ovaries were surgically removed. The authors repeat some of their prior work and provide new and interesting findings about the effects of estrogen on currents mediated by calcium and potassium channels, suggest a neurotransmitter "switch", and suggest Trpc5 regulates Kisspeptin 1 neuron excitability. While useful in its significance, there are concerns that the evidence for some conclusions is incomplete. This study will be of interest to endocrinologists and reproductive biologists.

Reviewer #1 (Public Review):

Summary:

In this work, Qiu and colleagues examined the effects of preovulatory (i.e., proestrous or late follicular phase) levels of circulating estradiol on multiple calcium and potassium channel conductances in arcuate nucleus kisspeptin neurons. Although these cells are strongly linked to a role as the "GnRH pulse generator," the goal here was to examine the physiological properties of these cells in a hormonal milieu mimicking late proestrus, the time of the preovulatory GnRH-LH surge. Computational modeling is used to manipulate multiple conductances simultaneously and support a role for certain calcium channels in facilitating a switch in firing mode from tonic to bursting. CRISPR knockdown of the TRPC5 channel reduced overall excitability, but this was only examined in cells from ovariectomized mice without estradiol treatment. The patch clamp experiments are comprehensive and overall solid but a direct demonstration of the role of these conductances in being necessary for surge generation (or at least having a direct physiological consequence on surge properties) is lacking, substantially reducing the impact of the findings.

Strengths:

(1) Examination of multiple types of calcium and potassium currents, both through electrophysiology and molecular biology.

(2) Focus on arcuate kisspeptin neurons during the surge is relatively conceptually novel as the anteroventral periventricular nucleus (AVPV) kisspeptin neurons have received much more attention as the "surge generator" population.

(3) The modeling studies allow for direct examination of manipulation of single and multiple conductances, whereas the electrophysiology studies necessarily require examination of each current in isolation. The construction of an arcuate kisspeptin neuron model promises to be of value to the reproductive neuroendocrinology field.

Weaknesses:

(1) The novelty of some of the experiments needs to be clarified. This reviewer's understanding is that prior experiments largely used a different OVX+E2 treatment paradigm mimicking periods of low estradiol levels, whereas the present work used a "high E2" treatment model. However, Figures 10C and D are repeated from a previous publication by the same group, according to the figure legend. Findings from "high" vs. "low" E2 treatment regimens should be labeled and clearly separated in the text. It would also help to have direct comparisons between results from low E2 and high E2 treatment conditions.

(2) In multiple places, links are made between the changes in conductances and the transition from peptidergic to glutamatergic neurotransmission. However, this relationship is never directly assessed. The data that come closest are the qPCR results showing reduced Tac2 and increased Vglut2 mRNA, but in the figure legend, it appears that these results are from a prior publication using a different E2 treatment regimen.

(3) Similarly, no recordings of arcuate-AVPV glutamatergic transmission are made so the statements that Kiss1ARH neurons facilitate the GnRH surge via this connection are still only conjecture and not supported by the present experiments.

(4) Figure 1 is not described in the Results section, and is only tenuously connected to the statement in the introduction in which it is cited. The relevance of panels C and D is not clear. In this regard, much is made of the burst firing pattern that arises after E2 treatment in the model, but this burst firing pattern is not demonstrated directly in the slice electrophysiology examples.

(5) In Figure 3, it would be preferable to see the raw values for R1 and R2 in each cell, to confirm that all cells were starting from a similar baseline. In addition, it is unclear why the data for TTA-P2 is not shown, or how many cells were recorded to provide this finding.

(6) In Figure 5, panel C lists 11 cells in the E2 condition but panel E lists data from 37 cells. The reason for this discrepancy is not clear.

(7) In all histogram figures, it would be preferable to have the data for individual cells superimposed on the mean and SEM.

(8) The CRISPR experiments were only performed in OVX mice, substantially limiting interpretation with respect to potential roles for TRPC5 in shaping arcuate kisspeptin neuron function during the preovulatory surge.

(9) Furthermore, there are no demonstrations that the CRISPR manipulations impair or alter the LH surge.

(10) The time of day of slice preparation and recording needs to be specified in the Methods.

Reviewer #2 (Public Review):

Summary:

Kisspeptin neurons of the arcuate nucleus (ARC) are thought to be responsible for the pulsatile GnRH secretory pattern and to mediate feedback regulation of GnRH secretion by estradiol (E2). Evidence in the literature, including the work of the authors, indicates that ARC kisspeptin coordinate their activity through reciprocal synaptic interactions and the release of glutamate and of neuropeptide neurokinin B (NKB), which they co-express. The authors show here that E2 regulates the expression of genes encoding different voltage-dependent calcium channels, calcium-dependent potassium channels, and canonical transient receptor potential (TRPC5) channels and of the corresponding ionic currents in ARC kisspeptin neurons. Using computer simulations of the electrical activity of ARC kisspeptin neurons, the authors also provide evidence of what these changes translate into in terms of these cells' firing patterns. The experiments reveal that E2 upregulates various voltage-gated calcium currents as well as 2 subtypes of calcium-dependent potassium currents while decreasing TRPC5 expression (an ion channel downstream of NKB receptor activation), the slow excitatory synaptic potentials (slow EPSP) elicited in ARC kisspeptin neurons by NKB release and expression of the G protein-associated inward-rectifying potassium channel (GIRK). Based on these results, and on those of computer simulations, the authors propose that E2 promotes a functional transition of ARC kisspeptin neurons from neuropeptide-mediated sustained firing that supports coordinated activity for pulsatile GnRH secretion to a less intense firing in glutamatergic burst-like firing pattern that could favor glutamate release from ARC kisspeptin. The authors suggest that the latter might be important for the generation of the preovulatory surge in females.

Strengths:

The authors combined multiple approaches in vitro and in silico to gain insights into the impact of E2 on the electrical activity of ARC kisspeptin neurons. These include patch-clamp electrophysiology combined with selective optogenetic stimulation of ARC kisspeptin neurons, reverse transcriptase quantitative PCR, pharmacology, and CRIPR-Cas9-mediated knockdown of the Trpc5 gene. The addition of computer simulations for understanding the impact of E2 on the electrical activity of ARC kisspeptin cells is also a strength.

The authors add interesting information on the complement of ionic currents in ARC kisspeptin neurons and on their regulation by E2 to what was already known in the literature. Pharmacological and electrophysiological experiments appear of the highest standards. Robust statistical analyses are provided throughout, although some experiments (illustrated in Figures 7 and 8) do have rather low sample numbers.

The impact of E2 on calcium and potassium currents is compelling. Likewise, the results of Trpc5 gene knockdown do provide good evidence that the TRPC5 channel plays a key role in mediating the NKB-mediated slow EPSP. Surprisingly, this also revealed an unsuspected role for this channel in regulating the membrane potential and excitability of ARC kisspeptin neurons.

Weaknesses:

The manuscript also has weaknesses that obscure some of the conclusions drawn by the authors.

One has to do with the fact that "burst-like" firing that the authors postulate ARC kisspeptin neurons transition to after E2 replacement is only seen in computer simulations, and not in slice patch-clamp recordings. A more direct demonstration of the existence of this firing pattern, and of its prominence over neuropeptide-dependent sustained firing under conditions of high E2 would make a more convincing case for the authors' hypothesis.

In addition, and quite importantly, the authors compare here two conditions, OVX versus OVX replaced with high E2, that may not reflect the physiological conditions (the diestrous [low E2] and proestrous [high E2] stages of the estrous cycle) under which the proposed transition between neuropeptide-dependent sustained firing and less intense burst firing might take place. This is an important caveat to keep in mind when interpreting the authors' findings. Indeed, that E2 alters certain ionic currents when added back to OVX females, does not mean that the magnitude of these ionic currents will vary during the estrous cycle.

Lastly, the results of some of the pharmacological and genetic experiments may be difficult to interpret as presented. For example, in Figure 3, although it is possible that blockade of individual calcium channel subtypes suppresses the slow EPSP through decreased calcium entry at the somato-dendritic compartment to sustain TRPC5 activation and the slow depolarization (as the authors imply), a reasonable alternative interpretation would be that at least some of the effects on the amplitude of the slow EPSP result from suppression of presynaptic calcium influx and, thus, decreased neurotransmitter and neuropeptide secretion. Along the same lines, in Figure 12, one possible interpretation of the observed smaller slow EPSPs seen in mice with mutant TRPC5 could be that at least some of the effect is due to decreased neurotransmitter and neuropeptide release due to the decreased excitability associated with TRPC5 knockdown.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

In this work, Qiu and colleagues examined the effects of preovulatory (i.e., proestrous or late follicular phase) levels of circulating estradiol on multiple calcium and potassium channel conductances in arcuate nucleus kisspeptin neurons. Although these cells are strongly linked to a role as the "GnRH pulse generator," the goal here was to examine the physiological properties of these cells in a hormonal milieu mimicking late proestrus, the time of the preovulatory GnRH-LH surge. Computational modeling is used to manipulate multiple conductances simultaneously and support a role for certain calcium channels in facilitating a switch in firing mode from tonic to bursting. CRISPR knockdown of the TRPC5 channel reduced overall excitability, but this was only examined in cells from ovariectomized mice without estradiol treatment. The patch clamp experiments are comprehensive and overall solid but a direct demonstration of the role of these conductances in being necessary for surge generation (or at least having a direct physiological consequence on surge properties) is lacking, substantially reducing the impact of the findings.

Strengths:

(1) Examination of multiple types of calcium and potassium currents, both through electrophysiology and molecular biology.

(2) Focus on arcuate kisspeptin neurons during the surge is relatively conceptually novel as the anteroventral periventricular nucleus (AVPV) kisspeptin neurons have received much more attention as the "surge generator" population.

(3) The modeling studies allow for direct examination of manipulation of single and multiple conductances, whereas the electrophysiology studies necessarily require examination of each current in isolation. The construction of an arcuate kisspeptin neuron model promises to be of value to the reproductive neuroendocrinology field.

We thank the reviewer for recognizing our comprehensive examination of Kiss-ARH neurons through electrophysiological, molecular and computational modeling of their activity during the preovulatory surge, which as the reviewer pointed out is “conceptually novel.” We have bolstered our argument that Kiss1-ARH neurons transition from synchronized firing to burst firing with the E2-mediated regulation of channel expression with the addition of new experiments. We have addressed the recommendations as follows:

Weaknesses:

(1) The novelty of some of the experiments needs to be clarified. This reviewer's understanding is that prior experiments largely used a different OVX+E2 treatment paradigm mimicking periods of low estradiol levels, whereas the present work used a "high E2" treatment model. However, Figures 10C and D are repeated from a previous publication by the same group, according to the figure legend. Findings from "high" vs. "low" E2 treatment regimens should be labeled and clearly separated in the text. It would also help to have direct comparisons between results from low E2 and high E2 treatment conditions.

We have revised Figures 10C and 10D to include new findings (only) on Tac2 and Vglut2 expression in OVX and E2-treated Kiss1ARH. Most importantly, our E2 treatment regime is clearly stated in the Methods and is exactly the same that was used previously (Qiu, eLife 2016 and Qiu, eLife 2018) for the induction of the LH surge in OVX mice (Bosch, Molecular and Cellular Endocrinology 2013) .

(2) In multiple places, links are made between the changes in conductances and the transition from peptidergic to glutamatergic neurotransmission. However, this relationship is never directly assessed. The data that come closest are the qPCR results showing reduced Tac2 and increased Vglut2 mRNA, but in the figure legend, it appears that these results are from a prior publication using a different E2 treatment regimen.

In the revised Figure 1, we have now included a clear depiction of the transition from synchronized firing driven by NKB signaling in OVX females to burst firing driven by glutamate in E2-treated females. All of the qPCR results in the revised manuscript are new. We have used the same E2 treatment paradigm as previously published (Qiu, eLife 2018).

(3) Similarly, no recordings of arcuate-AVPV glutamatergic transmission are made so the statements that Kiss1ARH neurons facilitate the GnRH surge via this connection are still only conjecture and not supported by the present experiments.

Using a horizontal hypothalamic slice preparation, we have shown that Kiss1-ARH neurons excite GnRH neurons via Kiss1ARH glutaminergic input to Kiss1AvPV/Pen neurons (summarized in Fig. 12, Qiu, eLife 2016). We did not think that it was necessary to repeat these experiments for the current manuscript.

(4) Figure 1 is not described in the Results section and is only tenuously connected to the statement in the introduction in which it is cited. The relevance of panels C and D is not clear. In this regard, much is made of the burst firing pattern that arises after E2 treatment in the model, but this burst firing pattern is not demonstrated directly in the slice electrophysiology examples.

We have extensively revised Figure 1 to include new whole-cell, current clamp recordings that document burst firing in E2-treated, OVX females, which is now cited in the Results.

(5) In Figure 3, it would be preferable to see the raw values for R1 and R2 in each cell, to confirm that all cells were starting from a similar baseline. In addition, it is unclear why the data for TTA-P2 is not shown, or how many cells were recorded to provide this finding.

Before initiating photo-stimulation for each Kiss1-ARH neuron, we adjust the resting membrane potential to -70 mV, as noted in each panel in Figure 3, through current injections. We have now included new findings on the effects of the T-channel blocker TTA-P2 on slow EPSP in the revised Figure 3. The number of cells tested with each calcium channel blocker is depicted in each of the bar graphs summarizing the effects of the blockers (Figure 3E).

(6) In Figure 5, panel C lists 11 cells in the E2 condition but panel E lists data from 37 cells. The reason for this discrepancy is not clear.

In Figure 5D, we measured the L-, N-, P/Q and R channel currents after pretreatment with TTA-P2 to block the T-type current, whereas in Figure 5C, we measured the total current without TTA-P2.

(7) In all histogram figures, it would be preferable to have the data for individual cells superimposed on the mean and SEM.

In the revised Figures we have included the individual data points for the individual neurons and animals (qPCR).

(8) The CRISPR experiments were only performed in OVX mice, substantially limiting interpretation with respect to potential roles for TRPC5 in shaping arcuate kisspeptin neuron function during the preovulatory surge.

The TRPC5 channels are most important for generating slow EPSPs when expression of NKB is high in the OVX state. Conversely, the glutamatergic response becomes more significant when the expression of NKB and TRPC5 channel are muted in the E2-treated state. Therefore, the CRISPR experiments were specifically conducted in OVX mice to maximize the effects.

(9) Furthermore, there are no demonstrations that the CRISPR manipulations impair or alter the LH surge.

In this manuscript, our focus is on the cellular electrophysiological activity of the Kiss1ARH neurons in OVX and E2-treated OVX females. Exploration of CRISPR manipulations related to the LH surge is certainly slated for future experiments, but these in vivo experiments are beyond the scope of these comprehensive cellular electrophysiological and molecular studies.

(10) The time of day of slice preparation and recording needs to be specified in the Methods.

We have provided the times of slice preparation and recordings in the revised Methods and Materials.

Reviewer #2 (Public Review):

Summary:

Kisspeptin neurons of the arcuate nucleus (ARC) are thought to be responsible for the pulsatile GnRH secretory pattern and to mediate feedback regulation of GnRH secretion by estradiol (E2). Evidence in the literature, including the work of the authors, indicates that ARC kisspeptin coordinate their activity through reciprocal synaptic interactions and the release of glutamate and of neuropeptide neurokinin B (NKB), which they co-express. The authors show here that E2 regulates the expression of genes encoding different voltage-dependent calcium channels, calcium-dependent potassium channels, and canonical transient receptor potential (TRPC5) channels and of the corresponding ionic currents in ARC kisspeptin neurons. Using computer simulations of the electrical activity of ARC kisspeptin neurons, the authors also provide evidence of what these changes translate into in terms of these cells' firing patterns. The experiments reveal that E2 upregulates various voltage-gated calcium currents as well as 2 subtypes of calcium-dependent potassium currents while decreasing TRPC5 expression (an ion channel downstream of NKB receptor activation), the slow excitatory synaptic potentials (slow EPSP) elicited in ARC kisspeptin neurons by NKB release and expression of the G protein-associated inward-rectifying potassium channel (GIRK). Based on these results, and on those of computer simulations, the authors propose that E2 promotes a functional transition of ARC kisspeptin neurons from neuropeptide-mediated sustained firing that supports coordinated activity for pulsatile GnRH secretion to a less intense firing in glutamatergic burst-like firing pattern that could favor glutamate release from ARC kisspeptin. The authors suggest that the latter might be important for the generation of the preovulatory surge in females.

Strengths:

The authors combined multiple approaches in vitro and in silico to gain insights into the impact of E2 on the electrical activity of ARC kisspeptin neurons. These include patch-clamp electrophysiology combined with selective optogenetic stimulation of ARC kisspeptin neurons, reverse transcriptase quantitative PCR, pharmacology, and CRIPR-Cas9-mediated knockdown of the Trpc5 gene. The addition of computer simulations for understanding the impact of E2 on the electrical activity of ARC kisspeptin cells is also a strength.

The authors add interesting information on the complement of ionic currents in ARC kisspeptin neurons and on their regulation by E2 to what was already known in the literature. Pharmacological and electrophysiological experiments appear of the highest standards. Robust statistical analyses are provided throughout, although some experiments (illustrated in Figures 7 and 8) do have rather low sample numbers.

The impact of E2 on calcium and potassium currents is compelling. Likewise, the results of Trpc5 gene knockdown do provide good evidence that the TRPC5 channel plays a key role in mediating the NKB-mediated slow EPSP. Surprisingly, this also revealed an unsuspected role for this channel in regulating the membrane potential and excitability of ARC kisspeptin neurons.

We thank the reviewer for recognizing that the “pharmacological and electrophysiological experiments appear of the highest standards” and “the addition of the computer modeling for understanding the impact of E2 on the electrical activity of ARC kisspeptin cells is also a strength. However, we agree with the reviewer that we needed to provide a direct demonstration of “burst-like” firing of Kiss1-ARH neurons, which we have provided in Figure 1. We have addressed the other recommendations as follows:

Weaknesses:

The manuscript also has weaknesses that obscure some of the conclusions drawn by the authors.

One has to do with the fact that "burst-like" firing that the authors postulate ARC kisspeptin neurons transition to after E2 replacement is only seen in computer simulations, and not in slice patch-clamp recordings. A more direct demonstration of the existence of this firing pattern, and of its prominence over neuropeptide-dependent sustained firing under conditions of high E2 would make a more convincing case for the authors' hypothesis.

We have provided a more direct demonstration of the existence of this firing pattern in the whole-cell current clamp experiments in the revised Figure 1.

In addition, and quite importantly, the authors compare here two conditions, OVX versus OVX replaced with high E2, that may not reflect the physiological conditions (the diestrous [low E2] and proestrous [high E2] stages of the estrous cycle) under which the proposed transition between neuropeptide-dependent sustained firing and less intense burst firing might take place. This is an important caveat to keep in mind when interpreting the authors' findings. Indeed, that E2 alters certain ionic currents when added back to OVX females, does not mean that the magnitude of these ionic currents will vary during the estrous cycle.

We have published that the magnitude of the slow EPSP, which is TRPC5 channel mediated, varies throughout the estrous cycle with the slow EPSP reaching a maximal amplitude during diestrus, which was significantly reduced during proestrus, similar to that found in OVX compared to E2-treated, OVX females (Figure 2, Qiu, eLife 2016). Moreover, TRPC5 channel mRNA expression, similar to the peptides, is downregulated by an E2 treatment (Figure 10 this manuscript) that mimics proestrus levels of the steroid (Bosch et al., Mol Cell Endocrinology 2013). Furthermore, the magnitude of ionic currents is directly proportional to the number of ion channels expressed in the plasma membrane, which we have found correlates with mRNA expression. Therefore, it is likely that the magnitude of these ionic currents will vary during the estrous cycle.

Lastly, the results of some of the pharmacological and genetic experiments may be difficult to interpret as presented. For example, in Figure 3, although it is possible that blockade of individual calcium channel subtypes suppresses the slow EPSP through decreased calcium entry at the somato-dendritic compartment to sustain TRPC5 activation and the slow depolarization (as the authors imply), a reasonable alternative interpretation would be that at least some of the effects on the amplitude of the slow EPSP result from suppression of presynaptic calcium influx and, thus, decreased neurotransmitter and neuropeptide secretion. Along the same lines, in Figure 12, one possible interpretation of the observed smaller slow EPSPs seen in mice with mutant TRPC5 could be that at least some of the effect is due to decreased neurotransmitter and neuropeptide release due to the decreased excitability associated with TRPC5 knockdown.

The reviewer raises a good point, but our previous findings clearly demonstrated that chelating intracellular calcium with BAPTA in whole-cell current clamp recordings abolishes the slow EPSP and persistent firing (Qiu et al., J. Neurosci 2021), which we have noted is the rationale for dissecting out the contribution of T, R, N, L and P/Q calcium channels to the slow EPSP in our current studies. The revised Figure 3 also includes the effects of T-channel blocker.

However, to further bolster the argument for the post-synaptic contribution of the calcium channels to the slow EPSP and eliminate the potential presynaptic effects of the calcium channel blockers on the postsynaptic slow EPSP amplitude, which may result from reduced presynaptic calcium influx and subsequently decreased neurotransmitter release, we have utilized an additional strategy. Specifically, we have measured the response to the externally administered TACR3 agonist senktide under conditions in which the extracellular calcium influx, as well as neurotransmitter and neuropeptide release, are blocked (revised Figure 3).

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

(1) The use of optogenetics in Figure 3 to trigger the slow EPSP could be better clarified in the text.

We have clarified in the Methods the optogenetic protocol for generating the slow EPSP, which we have published previously (Qiu et al., eLife 2016; eLife 2018, J. Neurosci 2021).

(2) The citation for Figure 4C in the text does not match what is shown in the figure.

Figure 4C has been removed in the revised manuscript.

(3) Figure 5 - it would be clearer to have panel D labeled as "model results" or similar to distinguish it from the slice recording data.

Panel D has been labeled as "Model results”.

(4) The text in lines 191-197 in the Results may be better suited to the Discussion.

We have modified the text in order to present the new findings without the discussion points.

(5) It is somewhat confusing to have figure panels cited out of order in the main text (e.g., 7H before 7G and 8H before 8G).

We have edited the text to report the findings in the proper order of the panels in Figures 7 and 8.

Reviewer #2 (Recommendations For The Authors):

- The observations that E2 treatment of OVX mice has an effect on the magnitude of a number of ionic currents does not necessarily mean that these changes will be seen during the estrous cycle, in response to fluctuations in circulating E2 concentrations. Experiments comparing either different estrous cycle stages or OVX mice treated with low or high E2 would be required to gain insight into this question. As such, the relevance of the authors' findings (however interesting these are as they stand) to any potential physiological endocrine/reproductive state transition is questionable, in the reviewer's opinion. The authors should acknowledge this important caveat and moderate the interpretations of their findings and the conclusions of their manuscript accordingly.

We have published that the magnitude of the slow EPSP, which is TRPC5 channel mediated, varies throughout the estrous cycle with the slow EPSP being large during diestrus and significantly reduced during proestrus, similar to that found in OVX compared to E2-treated, OVX females (Figure 2, Qiu, eLife 2016). Moreover, TRPC5 channel mRNA expression, similar to the peptides, is downregulated by an E2 treatment (Figure 10 this manuscript) that mimics proestrus levels of the steroid (Bosch et al., Mol Cell Endocrinology 2013). Furthermore, the magnitude of ionic currents is directly proportional to the number of ion channels expressed in the plasma membrane, which we have found correlates with mRNA expression. Therefore, it is likely that the magnitude of these ionic currents will vary during the estrous cycle.

- The bursting firing pattern that the authors refer to and postulate will favor glutamate release under high E2 conditions is only seen in the computer simulations, not in patch-clamp recordings in brain slices (see also comment below). This substantially weakens some of the conclusions of the manuscript. Unless the authors can convincingly demonstrate a change in ARC kisspeptin firing pattern in response to increasing E2 using electrophysiology, these conclusions should be moderated.

We now include examples of burst firing activity under E2-treatment conditions in Figure 1 and have included summary figure (pie chart) documenting that a significant percentage of cells exhibit this activity with E2 treatment.

Other comments:

- Title: "E2 elicits distinct firing patterns" is not shown in this work. As such, the title needs to be revised.

We now show these distinct firing patterns in Figure 1, so we think the wording in the title is an accurate reflection of our findings.

- Abstract: some of the interpretations are overstated, in the reviewer's opinion.

Line 23, "... elevating the whole-cell calcium current and contributing to high-frequency firing" should be moderated, as what is shown by the authors is that blockade of calcium channel subtypes suppresses the slow EPSP and associated firing, the frequency of which is not reported (see also a later comment).

We now include examples of burst firing activity under E2-treatment conditions in Figure 1 and have modified the abstract to state “high frequency burst firing.”

Lines 26-28, that "mathematical modeling confirmed the importance of TRPC5 channels for initiating and sustaining synchronous firing, while GIRK channels, activated by Dyn binding to kappa opioid receptors, were responsible for repolarization" is simply not what the simulations show, in the reviewer's opinion. Indeed, there is no consideration of synchronous activity in the model, which simulates the firing of a single ARC kisspeptin neuron. Further, the model shows that TRPC5 can contribute to overall excitability (firing in response to current injection, Figure 12G) and that increasing TRPC5 conductance increases firing in response to NKB while this is decreased by adding GIRK conductance to the model (Figure 13A). Therefore, considerations of the importance of TRPC5 channels in initiating synchronous firing and the role of Dyn A-induced GIRK activity should not be included in the interpretations of the mathematical simulations.

The significance of synchronization lies in the fact that when neuronal networks synchronize, the behavior of each neuron within the network becomes identical. In such scenarios, the firing of a single neuron mirrors the activity of the entire neuronal network. Consequently, our model simulations, based on a single-cell neuronal model, can be utilized to make reliable inferences about synchronized neuronal activity.

Lines 31-33 (also lines 92-95), that "the transition to burst firing with high, preovulatory levels of E2 facilitates the GnRH surge through its glutamatergic synaptic connection to preoptic Kiss1 neurons" is not supported by the experiments (physiologic or computational) described in the manuscript, and is, therefore, only speculative. These statements should be removed throughout the manuscript.

Previously, we (Qiu et al., (eLife 2016) documented a direct glutamatergic projection from Kiss1-ARH neurons to Kiss1-AVPV/PeN neurons. Moreover, Lin et al. (Frontiers Endocrinology 2021) demonstrated that low frequency stimulation of Kiss1-ARH:ChR2 neurons, that is known to only release glutamate, boosts the LH surge, and in a follow-up paper the O’Byrne lab blocked this stimulation with ionotropic glutamate antagonists (Shen et al., Frontiers in Endocrinology 2022). We have included these references in the Introduction and Discussion, but we did not think that it was necessary to cite these papers in the Abstract. However, we have re-worded this final statement in the Abstract to: “the transition to burst firing with high, preovulatory levels of E2 would facilitate the GnRH surge….”

- Introduction: the usefulness of Figure 1 is questionable. From reading the figure legend, it is the reviewer's understanding that panels A and B are published elsewhere (there is no description of methods or results in the manuscript). Further, panels C and D are meant to illustrate that ARC kisspeptin neurons display different types of firing in OVX vs E2-treated OVX mice. The legend to C indicates that the trace illustrates "synchronous firing" but shows one cell (how can this be claimed as synchronous?) - the legend to D indicates that the trace "demonstrates" burst firing in ARC kisspeptin neurons. This part of the figure is, in the reviewer's opinion, misleading because these are only two examples (no quantifications or replicates are provided) obtained by stimulating firing in two different endocrine conditions by two different agonists. The "demonstration" of differential firing patterns would require a thorough examination of firing patterns in response to current injections (as in Figure 12 E-F) or in response to the two agonists, under the different hormonal conditions.

Figure 1 has now been completely revised to include new data documenting the different firing patterns. The methods detailing these experiments can be found in the Material and Methods section.

The introduction presents a rather incomplete picture of what is known regarding how ARC kisspeptin neurons might coordinate their activity to drive episodic GnRH secretion, and it omits published work showing that blockade of glutamate receptors (in particular AMPA receptors) decreases ARC kisspeptin neuron coordinated activity in the brain slices and in vivo and suppresses pulsatile GnRH/LH secretion in mice.

If we are not mistaken, the reviewer is referring to fiber photometry recordings of GCaMP activity, which we cite in the Discussion. However, for the Introduction we tried to “set the stage” for our studies on measuring the individual channels underlying the different firing patterns and how they are regulated by E2.

The introduction is also quite long with extensive descriptions of previous work by the authors and in other brain areas that would be better suited for the discussion.

Again, we are trying to rationalize why we focused on particular ion channels based on the literature.

- Results: lines 129-132 should be moderated, as whether calcium channels increase excitability or facilitate TRPC5 channel opening has not been directly assessed here.

High frequency optogenetic stimulation of Kiss1-ARH neurons and NKB through its cognate receptor (TACR3) activates TRPC 5 channels (Qiu et al., eLife 2016; J. Neurosci 2021). BAPTA prevents the opening of TRPC5 channels and abrogates the slow EPSP following high frequency stimulation. Figure 3 documents that inhibition of voltage-activated calcium channels attenuates the slow EPSP, which results in a decrease in excitability.

Lines 145-146, one limitation of this experiment is that blockade of calcium channel subtypes will not only affect calcium entry and subsequent actions of calcium on TRPC5 channels but also impair the release of neurotransmitters and neuropeptides from kisspeptin neurons. The interpretation that "calcium channels contribute to maintaining the sustained depolarization underlying the slow EPSP" needs, therefore, to be moderated as it is not possible to extract the direct contribution of calcium channels to the activation of TRPC5 channels from these experiments.

We cited our previous findings documenting that chelating intracellular calcium with BAPTA abolishes the slow EPSP and persistent firing (Qiu et al., J Neurosci 2021). However, to eliminate the potential effects of calcium channel blockers on the slow EPSP amplitude, which may result from reduced presynaptic calcium influx and subsequently decreased neurotransmitter and neuropeptide secretion, we adopted a different strategy by comparing responses between Senktide and Cd2+ plus Senktide. Our findings revealed that the non-selective Ca2+ channel blocker Cd2+ significantly inhibited Senk-induced inward current (Figures 3F-H).

Panel C should be removed from Figure 4, as it is published elsewhere.

Figure 4C has been removed.

Lines 168-169, "...E2 treatment led to a significant increase in the peak calcium current density in Kiss1ARH neurons, which was recapitulated as predicted by our computational modeling..." How did the model "predict" this increase in calcium current density? As no information is provided in the methods or supplementary information as to how the effect of E2 was integrated into the model, the authors will need to provide additional narration in the text to explain this statement. The "T-channel inflection" referred to in the figure legend will also need to be explained. Lastly, in Figure 5C, the current density unit should be pA/pF.

We have added text in the supplementary information to explain how we used the qPCR and electrophysiological data to inform the model regarding the effect that E2 has on the various ionic currents and noted in the Figure 13 legend that the increase/decrease in the conductances is physiologically mediated by E2. We have eliminated the T-channel inflection point (Figure 5D) and corrected the current density label (Figure 5C).

Lines 198-199, please clarify "E2 does not modulate calcium channel kinetics directly but rather alters the mRNA expression to increase the conductance".

We have clarified that “that long-term E2 treatment does not modulate calcium channel kinetics but rather alters the mRNA expression to increase the calcium channel conductance” by referring to the specific figures (i.e., Figures 4, 6) in a previous sentence.

Figures 7 and 8 titles do not accurately reflect the contents: there is nothing about repolarization in the experiments illustrated in Figure 7 or Figure 8. The sample sizes (3 to 4 cells) are also quite small for these experiments.

We have modified the Figure titles per the reviewer’s comments and increased the cell numbers.

The title of Figure 9 also does not fully reflect the figure's contents. Although panel G does suggest that the M current contributes to regulating the membrane potential, the reviewer's reading of this figure panel is that the fractional contribution of the M current does not vary during a short burst of action potentials. The suggestion that "KCNQ channels play a key role in repolarizing Kiss1ARH neurons following burst firing" (line 272) and the statement that "our modeling predicted that M-current contributed to the repolarization following burst firing" (line 273) should be revised accordingly.

The point is that the M-current contributes, albeit a small fraction, to the repolarization during burst firing.

Line 288, please indicate what figure informs this statement.

We have revised the statement since the modeling (Figure 13) comes later in the Results.

Line 311-313, this sentence only superficially describes the simulation, in the reviewer's opinion. Does the model inform on how TRPC5 channels/currents do that? The supplementary information indicates that there is a tone of extracellular neurokinin B embedded in the model. This is important information that should be clearly stated in the manuscript. The authors should also consider discussing the influence of this neurokinin B tone on the contribution of TRPC5 to cell excitability. As a neurokinin B tone in the extracellular space will likely alter the firing of kisspeptin neurons in the model, readers will likely need more information about all this.

In our current ramp simulations of the model (Fig 12 G&H) there is no involvement of neurokinin B (i.e., the NKB parameter is set to zero), and the effect on the rheobase is solely due to the decrease of the TRPC5 conductance. In the model, TRPC5 channels are activated by intracellular calcium levels and are therefore contributing to cell excitability even in the absence of extracellular NKB. The NKB tone is used for the simulations presented in Figure 13 where we vary the TRPC5 conductance under saturating levels of extracellular NKB.

Lines 316-318 also read as quite superficial. More explanations of what is illustrated in Figure 13 are needed. In particular, it is unclear from the methods and supplementary information what the different ratios of conductances in OVX+E2 vs in OVX are and how they were varied in the model. Furthermore, it is unclear to the reviewer how the outcome of these simulations matches the authors' postulate that E2 enables a transition to a burst firing pattern that favors glutamate release. Looking at simulated firing in Figure 13B, E2 (by increasing calcium conductances) would tend to enable high-frequency firing within bursts (nearing 50 Hz by eye) and high burst rates (approximately 4 bursts per second), which the reviewer would argue might be expected to cause significant neuropeptide release in addition to that of glutamate.

We have added to the text: “Furthermore, the burst firing of the OVX+E2 parameterized model was supported by elevated h- and Ca 2+-currents (Figure 13B) as well as by the high conductance of Ca2+ channels relative to the conductance of TRPC5 channels (Figure 13C).” We have also provided in the Supplemental Information (Table of Model Parameters) the specific conductances in the OVX and OVX+E2 state and how they are varied to produce the model simulations.

Granted the high frequency firing during a burst could release peptide, but in the E2-treated, OVX females the expression of the peptides are at “rock bottom.” Therefore, the sustained high frequency firing during the slow EPSP in the OVX state would generate maximum peptide release.

In Figure 13C, the reviewer is unclear on the ranges of TRPC5 conductances shown. The in vitro experiments suggest that E2 suppresses Trpc5 gene expression and might suppress TRPC5 currents. The ratio of gTRPC5(OVX+E2)/gTRPC5(OVX) should, thus, be <1.0. This is not represented in the parameter space provided, making the interpretation of this simulation difficult. Please clarify what the effect of decreasing gTRPC5 will be on firing patterns in the model.

Thank you for pointing this typographical error. The ratio should be gTRPC5 (OVX)/TRPC5(OVX + E2) for the X-axis.

- Discussion: many statements and conclusions are overreaching and need to be revised; for example lines 320-322, 329-330, 335-338, 369, 371-373, 391-394, 463-464, and 489-494;

We have tempered these statements, so they are not “overreaching.”

Lines 489-494: the authors should integrate published observations that i) ablation of ARC kisspeptin neurons results in increased LH surges in mice and rats and that ii) optogenetic stimulation of ARC kisspeptin fibers in the POA is only effective at increasing LH secretion in a surge-like manner when done at high frequencies (20 Hz), in their discussion of the role of ARC kisspeptin neurons and their firing patterns in the preovulatory surge.

We have included the paper from the O’Byrne lab (Shen et al. Frontiers in Endocrinology 2022) in the Discussion. However, the Mittleman-Smith paper (Endocrinology, 2016) ablating KNDy neurons using NK3-saporin not only targeted KNDy neurons but other arcuate neurons that express NK3 receptors. Therefore, we have not cited it in the Discussion.

Version 2

Abstract

Abstract

Hypothalamic kisspeptin (Kiss1) neurons are vital for pubertal development and reproduction. Arcuate nucleus Kiss1 (Kiss1ARH) neurons are responsible for the pulsatile release of Gonadotropin-releasing Hormone (GnRH). In females, the behavior of Kiss1ARH neurons, expressing Kiss1, Neurokinin B (NKB), and Dynorphin (Dyn), varies throughout the ovarian cycle. Studies indicate that 17β-estradiol (E2) reduces peptide expression but increases Vglut2 mRNA and glutamate neurotransmission in these neurons, suggesting a shift from peptidergic to glutamatergic signaling. To investigate this shift, we combined transcriptomics, electrophysiology, and mathematical modeling. Our results demonstrate that E2 treatment upregulates the mRNA expression of voltage-activated calcium channels, elevating the whole-cell calcium current and that contribute to high-frequency burst firing. Additionally, E2 treatment decreased the mRNA levels of Canonical Transient Receptor Potential (TPRC) 5 and G protein-coupled K+ (GIRK) channels. When TRPC5 channels in Kiss1ARH neurons were deleted using CRISPR, the slow excitatory postsynaptic potential (sEPSP) was eliminated. Our data enabled us to formulate a biophysically realistic mathematical model of the Kiss1ARH neuron, suggesting that E2 modifies ionic conductances in Kiss1ARH neurons, enabling the transition from high frequency synchronous firing through NKB-driven activation of TRPC5 channels to a short bursting mode facilitating glutamate release. In a low E2 milieu, synchronous firing of Kiss1ARH neurons drives pulsatile release of GnRH, while the transition to burst firing with high, preovulatory levels of E2 would facilitate the GnRH surge through its glutamatergic synaptic connection to preoptic Kiss1 neurons.

Introduction

Hypothalamic kisspeptin (Kiss1) neurons and its cognate receptor (GPR 54 or Kiss1 R) are essential for pubertal development and reproduction, and may also be involved in the control of energy homeostasis (Kotani et al., 2001) (De Roux et al., 2003) (Seminara et al., 2003) (Messager et al., 2005) (Shahab et al., 2005) (d’Anglemont de Tassigny et al., 2008) (Qiu J. et al., 2018) (Rønnekleiv et al., 2022). Kisspeptin neurons within the arcuate nucleus of the hypothalamus (Kiss1ARH) co-express Kiss1, Neurokinin B (NKB) and Dynorphin (Dyn), which are all down-regulated by 17β-estradiol (E2) (Goodman et al., 2007; Navarro V. M. et al., 2009). However, Kiss1ARH neurons also express vesicular glutamate transporter 2 (vGlut2) and release glutamate, and both vGlut2 expression and glutamate release are upregulated by E2 in females (Qiu J. et al., 2018). This indicates that peptides and glutamate in Kiss1ARH neurons are differently modulated by E2 and suggests that there is a complex E2 regulation in these neurons such that they transition from predominantly peptidergic to glutamatergic neurotransmission and hence from a “pulsatile” to “surge” mode. It has been known for decades that neurons located within the arcuate nucleus are responsible for the pulsatile release of GnRH and subsequently pulsatile release of LH from the pituitary gland (O’Byrne et al., 1991) (Moenter et al., 1993). In this respect, it is now generally accepted that Kiss1ARH neurons are the main neurons responsible for the generation of pulsatile LH release, but the underlying cellular conductances generating this activity have not been elucidated.

Morphological studies have provided evidence that Kiss1ARH neurons can communicate directly with each other (Lehman et al., 2010) (Navarro V. M. et al., 2009) (Navarro V.M. et al., 2011). Furthermore, Kiss1ARH neurons express the NKB receptor, Tacr3, as well as the kappa (κ) opioid receptor (KOR), whereas GPR54 is not expressed in Kiss1ARH neurons, rendering them unresponsive to kisspeptin (d’Anglemont de Tassigny et al., 2008; Navarro V. M. et al., 2009; Wakabayashi et al., 2010). Using optogenetics and whole-cell recordings we demonstrated that high-frequency photoactivation of Kiss1ARH neurons induces an NKB-mediated slow excitatory postsynaptic potential (EPSP), which is mediated by the recruitment of canonical transient receptor potential (TRCP5) channels. The release of NKB is limited by co-released dynorphin, which acts presynaptically to inhibit further release. Together the two peptides cause synchronized firing of Kiss1ARH neurons (Kelly et al., 2018; Qiu J. et al., 2016; Qiu Jian et al., 2021), whereas kisspeptin and glutamate appear to be the main output signals from Kiss1ARH neurons (Qiu J. et al., 2016; Qiu J. et al., 2018) (Voliotis et al., 2021) (Liu et al., 2021).

Although single action potential-generated calcium influx is sufficient to trigger the release of classical neurotransmitters such as glutamate, high frequency (10-20 Hz) is required for the release of neuropeptides such as kisspeptin, NKB and dynorphin (Qiu J. et al., 2016). Indeed, the slow EPSP, which underlies the synchronization, is similar to the “plateau potential” that has been described in hippocampal and cortical neurons (Arboit et al., 2020; Zhang Z. et al., 2011). Many neurons, including Kiss1ARH neurons, express the biophysical properties that allow them to continue to persistently fire even after a triggering synaptic event has subsided (Zylberberg and Strowbridge, 2017) (Qiu J. et al., 2016). Moreover, the intrinsic bi-stability of neurons that generates persistent firing activity has been linked to a calcium-activated, non-selective cation current (ICAN) (Zylberberg and Strowbridge, 2017), and TRPC channels, specifically TRPC5 channels, are thought to be responsible for the ICAN in cortical neurons (Zhang Z. et al., 2011). Therefore, we postulated that TacR3 activation via NKB drives influx of Ca+2 through TRPC5 channels leading to greater build-up of [Ca2+]i that facilitates the opening of more TRPC5 channels in a self-sustaining manner. Indeed, using the fast intracellular calcium chelator BAPTA, which has been shown to robustly inhibit TRPC5 channel activation in heterologous cells (Blair et al., 2009), we have been able to abolish the slow EPSP and persistent firing in Kiss1ARH neurons following optogenetic stimulation in female mice (Qiu Jian et al., 2021).

Although the expression of peptides in Kiss1ARH neurons are downregulated by high circulating levels (late follicular levels) of E2, the intrinsic excitability of and the glutamate release by Kiss1ARH neurons are increased by Cacna1g (Cav3.1, T-type calcium channel), Hcn1 and Hcn2 (Hyperpolarization-activated, Cyclic Nucleotide Gated channels) mRNA expression and Vglut2 mRNA expression, respectively (Qiu J. et al., 2018). Burst firing in CNS neurons, which efficiently releases fast amino acid transmitters like glutamate, is generated primarily by the T-type calcium channel current (IT) (e.g., in thalamic relay neurons), and the rhythmicity of this burst firing is dependent on the h-current (Ih) (for review, see (Lüthi and McCormick, 1998; Zagotta and Siegelbaum, 1996)). Ih is mediated by the HCN channel family, which includes channel subtypes 1-4, of which Hcn1 and Hcn2 are the main channels in Kiss1ARH neurons. Ih depolarizes neurons from hyperpolarized states, raising the membrane potential into the range of IT activation (Erickson et al., 1993a, 1993b; Kelly and Rønnekleiv, 1994; Lüthi and McCormick, 1998; Zhang C. et al., 2009). IT is mediated by the low-threshold voltage-gated calcium channels, CaV3.1-3.3 (for review see (Perez-Reyes, 2003)). IT initiates a transient Ca2+-driven depolarization above the threshold for action potential initiation (i.e., a low threshold spike) (Llinás, 1988; Tsien et al., 1987). This depolarization then drives neurons to fire an ensemble (burst) of Na+-driven action potentials.

Based on the above compelling evidence we postulated that Kiss1ARH neurons transition from peptidergic neurotransmission, driving the pulsatile release of GnRH via kisspeptin release into the median eminence, to glutamatergic transmission that facilitates in the preovulatory surge of GnRH (Lin et al., 2021) (Shen et al., 2022) through their projection to the anteroventral periventricular/periventricular nucleus Kiss1 (Kiss1AVPV/PeN ) neurons (Qiu J. et al., 2016). Therefore, we initiated studies to thoroughly characterize effects of high circulating (late follicular) levels of E2 on the expression of the full complement of voltage-activated calcium channels (and currents) and the opposing K+ channels, involved in the repolarization, on the excitability of Kiss1ARH neurons. We performed whole-cell recordings and single cell RT-PCR analysis of Kiss1ARH neurons to determine which channels are involved in the physiological transition from peptidergic to glutamatergic neurotransmission. Our physiological findings were incorporated into a mathematical model that accounts for the E2 effects on the firing activity of Kiss1ARH neurons and validates our hypothesis that high levels of E2 facilitate the transition from peptidergic to glutamatergic neurotransmission.

Results

Voltage-activated Ca2+ channels contribute to burst firing of Kiss1ARH neurons

Our whole-cell current clamp recordings of Kiss1ARH neurons revealed that there is a significant transition of these Kiss1 neurons from a silent/tonic firing mode in the ovariectomized state to bursting/irregular firing with E2 treatment (Figure 1). Therefore, we sought to elucidate the cationic channels contributing to this physiological transition. Previously, we have shown that an increase in the intracellular calcium concentrations can potentiate TRPC5 channel current in POMC neurons (Qiu J. et al., 2010). Additionally, chelating intracellular calcium with BAPTA abolishes the slow excitatory postsynaptic potential (EPSP) and persistent firing in Kiss1ARH neurons (Qiu Jian et al., 2021). Hence, to investigate the contributions of voltage-activated calcium channels (VGCCs) to the increase in intracellular calcium, we measured the peak calcium current contributed by both the low and high voltage-activated calcium channels. The inward currents evoked by the voltage pulses (150-ms pulses starting from a holding potential of -80 mV with 10-mV increments) were identified as calcium (Ca2+) currents, as they were blocked by the universal VGCC inhibitor Cd2+ (200 μM) (McNally et al., 2020) (Figures 2A-E). The maximum total inward and Cd2+-sensitive currents reached their peak amplitudes at -10 mV. To differentiate between various calcium channel subtypes present in Kiss1ARH neurons, we applied selective antagonists individually, allowing us to isolate the drug-sensitive current for each cell. When we individually applied specific antagonists, we observed partial inhibition of whole-cell Ca2+ currents. Treatment with 10 μM nifedipine (Hiraizumi et al., 2008; Kato et al., 2003; Lee et al., 2002), an L-type Ca2+ channel inhibitor, resulted in a partial inhibition (26.1%) of the whole-cell calcium current. Additionally, application of 2 μM ω-conotoxin GVIA (conoGVIA, targeting N-type channels) (Lee et al., 2002), 200 nM ω-agatoxin IVA (AgaIVA, targeting P/Q-type channels) (Kato et al., 2003; McNally et al., 2020), 100 nM SNX-482 (targeting R-type channels) (Hiraizumi et al., 2008), or 1 μM TTA-P2 (TTAP2, targeting T-type channels) (McNally et al., 2020) also led to partial inhibition of the Ca2+ current (Figure 2F). The observed reduction in current with each inhibitor indicates the presence of all the major subtypes of Ca2+ currents in Kiss1ARH neurons. Among the inhibitors used, the largest components of the whole-cell calcium current were sensitive to nifedipine (26.1%), conoGVIA (25.1%), and SNX-482 (31.1%) (Figures 2A, B, D and 2F). Subsequently, we documented contributions from TTA-P2-sensitive channels, accounting for approximately 6.7% of the total Ca2+ current, and AgaIVA-sensitive channels, which constituted approximately 3.9% (Figures 2E, C). These findings indicate that high voltage-activated L-, N-, and R-type channels constitute the largest component of the voltage-activated Ca2+ current in Kiss1ARH neurons that not only increases the overall excitability but also greatly facilitate TRPC5 channel opening (Blair et al., 2009), which is the major downstream target of TACR3 activation by NKB (Qiu J. et al., 2016).

Properties of the firing of Kiss1ARH neurons from OVX and E2-treated OVX mice.

A–E. Representative whole cell, current clamp recordings of spontaneous phasic burst firing (A, which is only seen in E2-treated mice with clear up and down states), irregular burst firing (B), tonic firing (C), irregular firing (D) and silent (E) in Kiss1ARH neurons from E2-treated OVX mice. F-G. Summary pie chart illustrating the distribution of firing patterns in Kiss1ARH neurons from OVX+E2 (F, from 6 mice) or OVX (G, from 8 mice) mice (OVX versus OVX+E2, Chi-square test, X2 = 16.05, df = 3, **p = 0.0011). H. Current clamp recording in a Kiss1ARH neuron from an OVX+E2 female demonstrating the response to glutamate (100 μM). RMP = -70 mV. Similar responses were observed in all recorded female Kiss1ARH neurons (n = 7 from 4 mice). I. The spiking activity above the bar in G was expanded to highlight the pronounced effects of glutamate on burst firing activity of Kiss1ARH neurons, characterized by an ensemble of spikes riding on top of low-threshold spikes (arrows). Drugs were rapidly perfused into the bath as a 4-μl bolus. J. Spontaneous AP frequency of Kiss1ARH neurons before, during, and after glutamate application (n = 7 cells). Data are presented as mean ± SEM. Statistical comparisons between different treatments were performed using one-way ANOVA (F(2, 18) = 14.60, p = 0.0002) followed by Bonferroni’s post hoc test. *** indicates p < 0.005. K-L. Representative traces showing the amplitude of slow EPSPs induced by high frequency photostimulation in Kiss1ARH neurons in the presence of ionotropic glutamate receptor antagonists CNQX and AP5 from OVX (K) and OVX+E2 (L) females. The arrows indicate the measurements of slow EPSP amplitude after low-pass filtering (shown in K). M. Bar graphs summarizing the slow EPSPs in the Kiss1ARH neurons from OVX and E2-treated OVX mice in the presence of CNQX and AP5. Statistical comparisons between the two groups were performed using an unpaired t-test (t(12) = 5.181, p =0.0002). Data are expressed as mean ± SEM , with data points representing the number of cells.

Relative contribution of voltage-gated calcium currents in Kiss1ARH neurons from OVX mice.

A-E, representative current –voltage relationships showing that Cd2+ (non-selective blocker of calcium channels) sensitive peak currents were inhibited by different calcium channel blockers: A, nifedipine; B, ω-conotoxin GIVA; C, ω-agatoxin IVA; D, SNX-482; E, TTA-P2. F, the maximum peak currents were measured at -10 mV. The proportions of Ca2+ currents inhibited by nifidipine (L type), ω-conotoxin GVIA (N type), ω-agatoxin IVA (P/Q), SNX-482 (R type) and TTA-P2 (T type). Data are expressed as mean ± SEM , with data points representing the number of cells.

Voltage-activated Ca2+ channels contribute to generation of slow-EPSP in Kiss1ARH neurons

Our previous study utilizing optogenetics demonstrated that high frequency photostimulation of Kiss1ARH neurons releases NKB. This release of NKB induces slow excitatory postsynaptic potentials (EPSPs) that facilitate the recruitment of other Kiss1ARH neurons, resulting in synchronous firing of the Kiss1ARH neuronal population (Qiu J. et al., 2016). Additionally, chelating intracellular calcium with the fast chelator BAPTA abolishes the slow EPSP and persistent firing in Kiss1ARH neurons, highlighting the role of calcium signaling in these processes (Qiu Jian et al., 2021). To access the involvement of HVA channels in the generation of the slow EPSP, we conducted experiments where we blocked the L-type Ca2+ channels with nifedipine (10 μM) and the N- and P/Q-type Ca2+ channels with ω-conotoxin MVIIC (1 μM) (Nunemaker et al., 2003). We then measured the slow EPSP. Indeed, both nifedipine and ω-conotoxin MVIIC significantly inhibited the slow EPSP by 35.2% and 60.4%, respectively (Figure 3A, B, E). In addition, we used SNX (100 nM), which selectively blocks R-type Ca2+ channels, and the slow EPSP was reduced to 28.7% of its control value (Figure 3C, E). The selective T-channel blocker TTA-P2 (5 µM) inhibited the slow EPSP by 28.6% (Figure 3D, E). Since the pharmacological blockade of the various calcium channels could also affect the presynaptic release of neurotransmitters (i.e., NKB), we measured the direct response of Kiss1ARH neurons to the TACR3 agonist senktide in the presence or absence of cadmium, a universal calcium channel blocker (Figures 3F-H). Indeed, Cd2+ significantly inhibited the inward current induced by senktide, which supports our posit that the plasma membrane calcium channels contribute to the TRPC5 channels activation. Therefore, it appears that calcium channels contribute to maintaining the sustained depolarization underlying the slow EPSP.

Blockade of HVA Ca2+ channels decreases the slow EPSP in Kiss1ARH neurons.

A-D, representative traces showing that the slow EPSPs induced by high frequency photostimulation were abolished by perfusing the blocker of the L-type calcium channel, nifedipine (A) or N- and P/Q -type calcium channels, ω-conotoxin MVIIC (B), or the R-type calcium channel, SNX 482 (C), or the T-type calcium channel, TTA-P2 (D), respectively. The arrows indicate the measurements of slow EPSP amplitude, denoted as R1 and R2, after low-pass filtering (shown in C). E. Bar graphs summarizing the effects of drugs on the R2/R1 ratios. The slow EPSP was generated in OVX Kiss1-Cre::Ai32 mice. Comparisons between different treatments were performed using a one-way ANOVA analysis (F (3, 51) = 14.36, p<0.0001) with the Bonferroni’s post hoc test. *, **, **** indicates p < 0.05, 0.01, 0.001, respectively vs. control. F-G. Representative traces show that in voltage clamp, senktide induced a significant inward current in Kiss1ARH neurons (F) in the presence of TTX. This current was inhibited by the calcium channel blocker CdCl2 (200 μM)(G) in another cell. H. Bar graphs summarize the effect of the calcium channel blocker CdCl2 on the senktide-induced inward currents. Un-paired t-test for F versus G: t(20) = 2.575, p = 0.0181; * p < 0.05. Data are expressed as mean ± SEM, with data points representing the number of cells.

E2 increases the mRNA expression and the whole-cell current of voltage-activated Ca2+ channels

The neuropeptides NKB (tachykinin2, Tac2) and kisspeptin (Kiss1), which are expressed in Kiss1ARH neurons, are crucial for the pulsatile release of gonadotropin-releasing hormone (GnRH) and reproductive processes. E2 decreases the expression of Kiss1 and Tac2 mRNA in Kiss1ARH neurons but enhances the excitability of Kiss1ARH neurons by amplifying the expression of Cacna1g, Hcn1, and Hcn2 mRNA, as well as increasing T-type calcium currents and h-currents, respectively (Qiu, 2018, 20613}. Moreover, E2 drives Slc17a6 mRNA expression and enhances glutamatergic synaptic input to arcuate neurons and Kiss1AVPV/PeN neurons (Qiu J. et al., 2018). As a result, the E2-driven increase in Kiss1ARH neuronal excitability and glutamate neurotransmission may play a crucial role in triggering the surge of GnRH, ultimately leading to the LH surge (Lin et al., 2021) (Shen et al., 2022).

To assess the impact of E2 on the modulation of voltage-activated calcium channels and Kiss1ARH neuronal excitability, we employed real-time PCR (qPCR) to measure the relative expression levels of ion channel subtypes in Kiss1ARH neurons. We compared the expression in E2-treated females to those treated with oil, using the specific primers listed in Table 1. The quantification was conducted on pools of 10 neurons, as indicated in the Methods and Materials. In both oil- and E2-treated females, we quantified the expression of Cav1.2 (L-type), Cav2.1 (P/Q-type), Cav2.2 (N-type) and Cav2.3 (R-type) mRNAs. Remarkably, all of these mRNA transcripts exhibited increased expression levels in response to E2 treatment (Figure 4). Congruent with our previous findings (Qiu J. et al., 2018), we observed that the mRNA expression of Cav3.1 and Hcn1 was also upregulated in response to E2 treatment (Figure 4). These results suggested that E2 has a regulatory effect on the expression and function of all of these ion channels in Kiss1ARH neurons. To determine whether the increased mRNA expression translated into functional changes at the cellular level, we measured the whole-cell calcium current in Kiss1ARH neurons obtained from ovariectomized mice treated with either vehicle or E2. We discovered that E2 treatment led to a significant increase in the peak calcium current density in Kiss1ARH neurons (Figure 5A-D), which was recapitulated by our computational modeling (Figure 5E). These findings indicate that the upregulation of the mRNA expression of calcium channels by E2 translated to an augmented peak calcium current in Kiss1ARH neurons.

A, E2 increases the expression of low and high voltage-activated calcium channels in Kiss1ARH neurons.

Kiss1ARH neurons (three 10-cell pools) were harvested from each of 5 vehicle- and 5 E2-treated, OVX females to quantify ion channel mRNA expression of low and high voltage activated calcium channels as described in the Methods. The analysis included: T-type (Cav3.1) low voltage-activated, as well as the following high-voltage activated channels: R-type (Cav 2.3), L-type (Cav 1.2), N-type (Cav 2.2) and P/Q-type (Cav 2.1) calcium channels. Interestingly, all of these channels were upregulated with E2 treatment, which significantly increased the whole-cell calcium current (see Figure 5). Bar graphs represent the mean ± SEM with data points representing individual animals (oil versus E2, Unpaired t test for Cav 3.1, t(8)= 7.105, ***p=0.0001; for Cav 2.3, t(8) = 3.007, *p = 0.0169; for Cav 1.2, t(8)= 2.721, *p =0.0262; for Cav 2.2, t(8) = 4.001, **p 0.0039; for Cav 2.1, t(8) = 4.225, **p =0.0028).B, The same Kiss1ARH neuronal pools were also analyzed for mRNA expression of HCN1 ion channels, and E2 also increased the expression of hyperpolarization-activated, cyclic-nucleotide gated HCN1 channels in Kiss1ARH neurons. HCN1 channel mRNA expression was the most highly upregulated by E2 treatment in Kiss1ARH neurons. The expression values were calculated via the ΔΔCT method, normalized to GAPDH and relative to the oil control values. (oil versus E2, Unpaired t test, t(8)= 11.450, ****p < 0.0001).

E2 treatment (positive-feedback regimen) increases the Ca2+ currents in Kiss1ARH neurons.

A-B, Ca2+ currents in Kiss1ARH neurons with the same membrane capacitance from oil-treated (A) or E2-treated (B) animals. C, the maximum peak currents were measured at -10 mV. The current amplitudes were normalized to the cell capacitance in all cases to calculate current density. The bar graphs summarized the density of Ca2+ current in Kiss1ARHneurons from oil-treated and E2-treated animals. The mean density was significantly greater in E2-treated (13.4 ± 0.9 pA/pF, n = 11) than in oil-treated OVX females (7.2 ± 0.5 pA/pF, n = 40) (unpaired t-test, t(49) = 5.75, ****p < 0.0001). D. Relative contribution of voltage-gated calcium currents in Kiss1ARH neurons from OVX, E2-treated mice. The maximum peak currents were measured at -10 mV. The proportions of Ca2+ currents inhibited by nifidipine (L type), ω-Conotoxin GVIA (N type), ω-agatoxin IVA (P/Q), SNX-482 (R type) and TTA-P2 (T type). Data are expressed as mean ± SEM, with data points representing the number of cells. E. The modeling predicts that E2-treated, OVX females exhibit a significantly greater inward Ca2+ current (red trace) than the vehicle-treated females (black trace). The coductance of the modelled voltage-gated calcium current (L-, N-, P/Q-, and R-type) is set to 2.1 nS in the OVX state and 2.8 nS in OVX+E2 state, while for the T-type is set to 0.66 nS in the OVX state and 5 nS in OVX+E2 state.

The largest components of the calcium currents in Kiss1ARH neurons from the E2-treated, ovariectomized females were found to be sensitive to nifedipine, conoGVIA, and SNX-482, accounting for approximately 24.9%, 24.6%, and 27.0% of the total current across cells, respectively, which is very similar to their contributions to the whole-cell current from vehicle-treated, ovariectomized females (Figure 3). In addition, contributions from TTA-P2-sensitive channels accounted for approximately 11.1% of the total Ca2+ current, while agaIVA-sensitive channels contributed to approximately 11.0% (Figure 5D). These results highlight the prevalence of L-, N-, and R-type calcium channels as the major contributors to the whole-cell calcium current in Kiss1ARH neurons from E2-treated, ovariectomized females, but also the involvement of T-type and P/Q-type channels.

E2 does not alter the kinetics of calcium channel activation or de-inactivation

In order to determine whether the increase in peak voltage-activated calcium current density was the result of E2 regulating calcium channel kinetics, mRNA expression or both, we examined the voltage dependence of activation and inactivation. By measuring the voltage dependence of activation, we assessed how E2 affects the ability of calcium channels to open in response to membrane potential changes. Similarly, by examining the voltage dependence of inactivation, we determined how E2 influences the inactivation kinetics of calcium channels. Based on our results, there was no difference in the voltage dependence of activation between cells from the vehicle-treated control group (V1/2 = -32.3 ± 2.1 mV; n = 13) and cells from estrogen-treated females (V1/2 = -33.6 ± 2.5 mV; n = 11). Similarly, there was not a significant difference in the voltage dependence of inactivation between control cells (V1/2 = -48.9 ± 4.8 mV; n = 6) and estrogen-treated cells (V1/2 = -44.1 ± 1.9 mV; n = 5) (Figure 6). Therefore, although long-term E2 treatment increased the mRNA expression of HVA calcium channels (Figure 4), it did not affect the channel kinetics in Kiss1ARH neurons (Figure 6). Also, E2 downregulated the expression of Kcnd2 mRNA encoding Kv4.2, which is expressed in Kiss1ARH neurons (Mendonça et al., 2018) and has similar kinetics of activation as the T-type calcium channels (Oil-treated, ovariectomized females relative mRNA expression: 1.053, n = 5 animals versus E2-treated expression: 0.5643, n=5; t-test p = 0.0061). This opposing K+ current would dampen the inward calcium current. Therefore, it appears that long-term E2 treatment does not modulate calcium channel kinetics but rather alters the mRNA expression to increase the calcium channel conductance.

Voltage dependence of ICa in Kiss1ARHneurons from OVX and OVX+E2 mice.

A-B, top panels: activation and inactivation protocol. Bottom: representative traces. C-D, the mean V1/2 values for calcium channel activation were not significantly different for cells from controls versus cells from E2-treated females. Similarly, the V1/2 values for channel steady-state inactivation were similar for both groups. Data are expressed as mean ± SEM.

BK, SK and KCNQ channels are involved in modulating excitability of Kiss1ARH neurons

In the brain, calcium plays a crucial role in sculpting neuronal firing by activating potassium channels, which subsequently influence neuronal behavior (Nicoll, 1988; Storm, 1990). Since HVA and LVA calcium channels were expressed in Kiss1ARH neurons, all of which contribute to the elevation of intracellular calcium concentration ([Ca2+]i) that facilitates TRPC5 channel opening (Blair et al., 2009), our next step involved measuring the changes in Ca2+-activated K+ channel conductances and assessing their mRNA expression. In various cell types increases in cytosolic calcium levels, whether resulting from extracellular influx or intracellular release, lead to the activation of plasma membrane calcium-dependent potassium channels (Sah Pankaj and Louise Faber, 2002). Similarly, in Kiss1ARH neurons, these channels would be activated by calcium influx through all four types of high voltage-gated calcium channels, as well as the low voltage-activated calcium channel, which are all active during action potential firing. The activity of Ca2+-activated K+ channels play a crucial role in numerous physiological processes, including secretion and the regulation of neuronal firing properties. Two main families of Ca2+-activated K+ channel channels have been characterized, distinguished by their biophysical and pharmacological properties. These families are known as BK (Big Conductance K+) and SK (Small Conductance K+) channels in the CNS (Kshatri et al., 2018). BK channels are known for their high potassium selectivity and large single channel conductance, typically ranging from 100 to 300 pS. Activation of BK channels requires both calcium binding and membrane depolarization (Blatz and Magleby, 1987; Marty, 1989; Sah P., 1996; Storm, 1990). On the other hand, SK channels are simply activated by increases in cytosolic calcium levels, with their half-maximal activation at 0.3 μM (Bond et al., 1999).

To investigate K+ currents, the cells were maintained at a holding potential of -70 mV while being exposed to blockers CNQX, AP5, picrotoxin and TTX. Subsequently, the membrane potential was stepped by depolarizing voltages, ranging from -60 mV to +40 mV in 10 mV increments, for a duration of 500 ms (Brereton et al., 2013). This protocol was employed to activate K+ currents (Figure 7A). First, we examined SK currents. The mean current density was determined at the end of the voltage pulses. In vehicle-treated, OVX females the application of the SK channel blocker apamin (100 nM) (Spergel, 2007) led to a significant reduction in whole-cell currents in the +10 to +40 mV range (Figures 7A, B). The mean outward current density at +40 mV in the control group was 125.5 ± 7.2 pA/pF (n = 5) with the apamin-sensitive component contributing 60.2 ± 3.3 pA/pF (n = 5) (Figures 7A, C). In contrast, in the E2-treated females, the overall mean outward current density at +40 mV was 195.4 ± 14.6 pA/pF (n = 6), which was significantly greater than the vehicle control group (Figures 7D, E). However, there was no significant difference in the apamin-sensitive component between the vehicle-treated and E2-treated females, 60.2 ± 3.3 pA/pF (n = 5) versus 128.1 ± 13 pA/pF (n = 6), respectively (Figure 7F). Furthermore, to investigate the expression of the mRNAs encoding SK channel subunits in Kiss1ARH neurons from vehicle-treated and E2-treated OVX females, qPCR experiments were performed on 10-cell Kiss1ARH neuronal pools (Figure 7G). We focused on SK3 channels because these channels exhibit the highest expression in the hypothalamus and E2 regulates their expression (Bosch et al., 2002). E2 treatment had no effect on the mRNA expression of the SK3 subunit. These molecular findings support our electrophysiology results. Our computational model was calibrated so that SK channels contributed ∼50 pA/pF to the whole-cell outward K+ current in E2-treated females (Figure 7H).

Contribution of Small conductance, calcium-activated K+ (SK) channel to Kiss1ARH neuronal excitability.

A. Representative traces of the inhibition of outward currents before (left, control) and after the specific SK blocker Apamin (500 nM, middle). Apamin sensitive currents were calculated from the subtraction of control and apamin at depolarized potentials (right). Cells were clamped at -70 mV and given 500 ms voltage pulses from -60 mV to +40 mV in 10 mV steps at 0.2 Hz, as shown in A at the bottom. B. Mean current density-voltage relationships measured at the end of the 500 ms voltage step ranging from -60 mV to +40 mV were obtained in the absence and presence of apamin (two-way ANOVA: main effect of treatment (F(1, 8) = 22.69, p = 0.0014), main effect of time (F(10, 80) = 306.0, p < 0.0001) and interaction (F(10, 80) = 24.76, p < 0.0001); mean ± SEM; n = 5; post hoc Bonferroni test, ***p < 0.005, ****p < 0.001). C. Apamin sensitive current densities were obtained from B (mean ± SEM, n = 5). D. Representative traces of the inhibition of outward currents before (left, control) and after the specific SK blocker Apamin (500 nM, middle). Apamin sensitive currents were resulted from the subtraction of control and apamin at depolarized potentials (right). E. Mean current density-voltage relationships measured at the end of the 500 ms voltage step ranging from -60 mV to +40 mV were obtained in the absence and presence of apamin (two-way ANOVA: main effect of treatment (F(1, 10) = 12.85, p = 0.0050), main effect of time (F(10,100) = 264.1, p < 0.0001) and interaction (F(10, 100) = 11.93, p < 0.0001); mean ± SEM, n = 6; post hoc Bonferroni test, **p < 0.01, ****p < 0.001). F. Apamin sensitive current densities were obtained from C and E (ns; Two-way ANOVA followed by Bonferroni post hoc test; mean ± SEM; OVX, n = 5; OVX+E2, n = 6). G. Kiss1ARH neurons (three 10-cell pools) were harvested from each of 5 vehicle- and 5 E2-treated, OVX females to quantify the mRNA expression of SK3 ion channel. E2 did not increase the mRNA expression of small conductance calcium-activated K+ (SK3) channels in Kiss1ARH. Bar graphs represent the mean ± SEM, with data points representing the number of animals, oil versus E2, Unpaired t test, t(8)= 0.551, p = 0.5967. The expression values were calculated via the ΔΔCT method, normalized to GAPDH and relative to the oil control values. H. The mathematical model was calibrated on the electrophysiology data from Kiss1ARH neurons before and after treatment with the specific SK blocker apamin, left panel versus middle panel respectively. For the calibration it was assumed that the applied concentration of apamin (500nM) completely blocked the SK current. The modeled apamin-sensitive current with gSK=28.1 nS (right panel) matches the electrophysiological data from OVX animals.

Additionally, following the same protocol and in the presence of the same cocktail of blockers (CNQX, AP5, picrotoxin and TTX), we investigated the contribution of BK channels to Kiss1ARH neuronal excitability. In the OVX females, the application of the BK channel blocker iberiotoxin (ibTx, 200 nM) (Niday and Bean, 2021) resulted in only a slight attenuation of the outward current (n = 5) (Figures 8A, B). The ibTx-sensitive current density measured at +40 mV was 32.8 ± 12.0 pA/pF (Figures 8A, C). However, in the E2-treated females, the application of ibTx significantly attenuated the whole-cell K+ current from +20 to +40 mV, (Figures 8D, E). Additionally, the ibTx-sensitive current was significantly larger in the +0 to +40 mV range in the E2-treated females compared to the OVX females (Figure 8F). These findings indicate that E2 treatment modulates the activity of ibTx-sensitive BK current in Kiss1ARH neurons, resulting in increased current density (90.0 ± 10.1 pA/pF (n = 6) versus 31.1 ± 8.4 pA/pF (n = 5) at +40 mV). To investigate the expression of mRNA encoding BK channel subunits in Kiss1ARH neurons from vehicle-treated and E2-treated OVX females, qPCR experiments were performed on 10-cell Kiss1ARH neuronal pools (Figure 8G). E2 treatment significantly increased the mRNA expression of the BKα1 (Kcnma1) subunit. These findings support our electrophysiological findings that there is a significant increase in BK channel activity in Kiss1ARH neurons with E2 treatment (Figure 8F). In addition, E2 increased the mRNA expression of Kcnb1 encoding delayed rectifier Kv2.1 (E2-treated relative mRNA expression: 1.672, n = 5 versus oil-treated mRNA expression: 1.086, n = 5; t-test p-value = 0.0024). The up-regulation of the two of these K+ channels (BKα1 and Kv2.1) would facilitate rapid repolarization of Kiss1ARH following an action potential. Our computational model was calibrated so that BK channels contributed ∼100 pA/pF to the whole cell outward K+ current in the E2-treated females (Figures 8H).

E2 up-regulates Kcnma1 mRNA and BK current in Kiss1ARH neurons.

A. Representative traces of the inhibition of outward currents before (left, control) and after the specific BK blocker iberiotoxin (IbTx; 200 nM, middle). IbTx sensitive currents were calculated from the subtraction of control and IbTx at depolarized potentials (right). Cells were clamped at -70 mV and given 500 ms voltage pulses from -60 mV to +40 mV in 10 mV steps at 0.2 Hz, as shown in A at the bottom. B. Mean current density-voltage relationships measured at the end of the 500 ms voltage step ranging from -60 mV to +40 mV were obtained in the absence and presence of IbTx (two-way ANOVA: main effect of treatment (F(1, 8) = 0.8841, p = 0.3746), main effect of time (F(10, 80) = 71.56), p < 0.0001) and interaction (F(10, 80) = 1.127, p = 0.3528); mean ± SEM, n = 5; post hoc Bonferroni test, p > 0.05). C. IbTX sensitive current densities were obtained from B (mean ± SEM, n = 5). D. Representative traces of the inhibition of outward currents before (left, control) and after the specific BK blocker iberiotoxin (IbTx; 200 nM, middle). IbTx sensitive currents were resulted from the subtraction of control and IbTx at depolarized potentials (right). E. Mean current density-voltage relationships measured at the end of the 500 ms voltage step ranging from -60 mV to +40 mV were obtained in the absence and presence of IbTX (two-way ANOVA: main effect of treatment (F(1, 10) = 4.660, p = 0.0562), main effect of time (F(10, 100) = 63.98, p < 0.0001) and interaction (F(10, 100) = 4.907, p < 0.0001); mean ± SEM, n = 6; post hoc Bonferroni test, *p < 0.05, **p < 0.01, **** < 0.001). F. IbTx sensitive current densities were obtained from C and E (two-way ANOVA: main effect of treatment (F(1, 9) = 22.04, p = 0.0011), main effect of time (F(10, 90) = 78.26, p < 0.0001) and interaction (F(10, 90) = 17.84, p <0.0001); mean ± SEM, OVX, n = 5; OVX+E2, n = 6; Bonferroni post hoc test, *p < 0.05, ***p < 0.005, ****p < 0.001). G. Kiss1ARH neurons (three 10-cell pools) were harvested from each of 5 vehicle- and 5 E2-treated, OVX females to quantify the mRNA expression of BKα channel. E2-treatment increased the mRNA expression of BKα. The expression values were calculated via the ΔΔCT method, normalized to GAPDH and relative to the oil control values. Bar graphs represent the mean ± SEM, with data points representing the number of animals (unpaired two-tailed t-test for BK, t(8) = 3.479, **p = 0.0083). H. The mathematical model was calibrated to reproduce the current voltage relationship observed in Kiss1ARH neurons before and after treatment with IbTx. For the calibration it was assumed that the applied concentration of IbTx (200 nM) completely blocked the BK current. The modeled IbTx -sensitive current with gBK=20.0 nS (right panel) matches the electrophysiological data from OVX+E2 animals.

Traditionally, the after hyperpolarization is divided into three distinct phases: fast (fAHP), medium (mAHP), and slow after hyperpolarization (sAHP) (Storm, 1990; Vogalis et al., 2003). The fast after hyperpolarization (fAHP) is primarily mediated by the BK family of potassium channels (Storm, 1987). The medium after hyperpolarization (mAHP) is predominantly mediated by apamin-sensitive SK2 channels (Bond et al., 2004; Peters et al., 2005). However, KCNQ family members contribute to both the mAHP and sAHP (Tzingounis A. V. et al., 2010; Tzingounis A. V. and Nicoll, 2008). Therefore, to investigate the contribution of KCNQ channels to Kiss1ARH neuronal excitability, voltage clamp experiments were conducted in the presence of TTX, CNQX, AP5, and picrotoxin, and a standard M-current protocol was run using the M-channel blocker XE-991 to isolate the M-current (Greene et al., 2017; Roepke et al., 2011) (Figure 9 A, B). The application of XE-991 resulted in the inhibition of M-current within a physiologically relevant voltage range of -60 to -30 mV in E2-treated OVX females but exhibited minimal impact in OVX females (Figure 9 C, D). Although the XE991-sensitive current was relatively small compared to other voltage-activated K+ conductances, it demonstrated a significant increase in E2-treated, OVX females (Figures 9 E). The maximum peak current density sensitive to XE-991 at -30 mV was found to be four times higher in E2-treated OVX females when compared to OVX females. This would contribute to the repolarization following burst firing. Furthermore, E2 increased the mRNA expression of Kcnq2, (Figure 9F), which suggests that KCNQ channels play a role in repolarizing Kiss1ARH neurons following burst firing. Indeed, our modeling predicted that M-current contributed to the repolarization following burst firing (Figure 9G).

E2 up-regulates KCNQ channels and M-current in Kiss1ARH neurons.

A-B. Representative current traces of the M-current inhibition caused by 40 µM XE-991 perfused for 10 min in (A) OVX-oil and (B) OVX+E2-treated female mice. Inset: M-current deactivation protocol. C-D. Current density-voltage plots from – 75 to –30 mV of vehicle and XE-991 perfusion in (C) OVX-oil and (D) OVX+E2-treated mice. Two-way ANOVA for C: main effect of treatment (F(1, 17) = 1.908, p = 0.1851), main effect of time (F(9, 153) = 187.1, p < 0.0001), and interaction (F(9, 153) = 3.901, p = 0.0002); Veh, n = 11; XE-991, n = 8; Bonferroni post hoc test, p > 0.05. For D: main effect of Veh and XE-991 (F(1, 24) = 24.92, p < 0.0001), main effect of time (F(9, 216) = 174.5, p < 0.0001), and interaction (F(9, 216) = 52.75, p < 0.0001); Veh, n = 13; XE-991, n = 13; Bonferroni post hoc test, a = p < 0.05, b = p < 0.001. E. Treatment with E2 elevated, while XE-991 diminished the maximum peak current density elicited by a -30 mV step in OVX- and OVX+E2-treated mice. Two-way ANOVA: main effect of Veh and XE-991 (F(1, 41) = 47.59, p < 0.0001), main effect of OVX and OVX+E2 (F(1, 41) = 15.76, p = 0.0003), and interaction (F(1, 41) = 18.2, p = 0.0001; Veh, n = 11; XE-991, n = 8; Bonferroni post hoc test, Veh: OVX vs. OVX+E2, a = p < 0.001. XE-991: OVX vs. OVX+E2, p > 0.05. Data are expressed as mean ± SEM, with data points representing the number of cells. F. Kiss1ARH neurons (three10-cell pools) were harvested from each of 5 vehicle- and 5 E2-treated, OVX females to quantify the mRNA expression of Kcnq2. E2 treatment increased the mRNA expression of Kcnq2. Unpaired t-test, t(8) = 4.850, **p = 0.0013. Data are expressed as mean ± SEM, with data points representing the number of animals. G. Percent contribution of the different K+ currents to the repolarization current during burst-type firing activity in the OVX+E2 state. At each time point, the length of each color bar denotes the percent contribution of the corresponding current to the total outward current.

E2 increases Vglut2 but down regulates Tac2, Trpc5 and Girk2 mRNA expression in Kiss1ARH neurons

Based on our electrophysiological results, Kiss1ARH neurons appear to transition from peptidergic to glutamatergic neurotransmission through E2-mediated changes in the expression of voltage-activated Ca2+ channels and K+ channels and their respective conductances. Therefore, we asked the question is there a difference in peptide and glutamate mRNA expression mediating this transition? Therefore, we ran a comparison between Tac 2 (NKB) and Vglut2 (surrogate for glutamate) expression. The cycle threshold (CT) was compared between Tac2 and Vglut2 as well as Kiss1, TRPC5 and GIRK2 in Kiss neuronal cell pools from OVX oil-treated and OVX E2-treated animals (Figures 10A, B). It is worth noting that lower number of cycles illustrate a higher quantity of mRNA expression because the fluorescence is detected earlier, and one cycle difference represents a doubling in expression. As expected, the reference gene Gapdh did not change with E2-treatment. However, quantitative PCR results revealed that E2 treatment of OVX females significantly reduced Tac2 expression (Figure 10C), whereas Vglut2 mRNA was significantly increased in Kiss1ARH neurons (Figure 10D). Moreover, both Trpc5 and Girk2 expression were significantly reduced in E2-treated, OVX females (Figures 10E, F).

Estradiol decreases Tac 2, Trpc5 and Kcnj6 but increases Vglut 2 mRNA expression in Kiss1ARH neurons.

A. qPCR amplification curves illustrating the cycle threshold (CT) for Tac2, Gapdh, Kiss1, Trpc5, Slc17a6 (Vglut2) and Kcnj6 (GIRK2) in Kiss1ARH ten cell neuronal pools (three to six pools from each animal) in OVX Oil-treated, and B, in OVX E2-treated females. C. Quantitative real-time PCR analysis of Tac2 mRNA, D, Vglut2, E, Trpc5, F, Kcnj6. Comparisons were made between Oil-treated and E2-treated, OVX females using the comparative 2-ΔΔCT method. Bar graphs represent the mean ± SEM, with data points representing the number of animals. (Unpaired t test for Tac2, t(6) = 10.670, ***p < 0.0001; for Vglut2, t(7) = 9.678, ***p < 0.0001; for Trpc5, t(6) = 5.774, **p = 0.0012; for Kcnj6, t(8) = 3.457, **p = 0.0086).

CRISPR mutagenesis of Trpc5 attenuates slow EPSP and reduces excitability of Kiss1ARH neurons

Our previous (Qiu Jian et al., 2021) and current findings suggested that TRPC5 channels play a dominant role in regulating cell excitability in ovariectomized females. Therefore, as proof of principle, we utilized a CRISPR approach to mutate TRPC5 channels in Kiss1ARH neurons similar to our previous studies (Hunker et al., 2020; Stincic et al., 2021). Hunker et al. developed a single viral vector for conditional expression of the smaller Staphylococcus aureus (SaCas9) and sgRNA that yields high-efficiency mutagenesis in specific cell types (Hunker et al., 2020). To selectively mutate Trpc5 in Kiss1ARH neurons, we generated two guide RNA’s, one targeting exon 2, which is conserved across all splice variants, and the other targeting exon 7, the pore forming domain (Figures 11A, B). A cohort of Kiss1ARH mice were given bilateral stereotaxic injections into the ARH of the two AAV1-FLEX-SaCas9-sgTrpc5’s or a control virus containing with AAV1-FLEX-SaCas9-U6-sgRosa26. An additional Cre-dependent virus of the same serotype (AAV1) that drove expression of a fluorophore (YFP or mCherry) was co-administered in order to visualize injection quality and facilitate harvesting of cells (Figure 11C). After three weeks, mice underwent ovariectomy since OVX mice express the maximum slow EPSP amplitude (Qiu J. et al., 2016). Brain slices were prepared, and cells harvested as previously described (Qiu J. et al., 2018) and analyzed with qPCR. We found that the Trpc5 mutagenesis group displayed a reduction in relative expression of Trpc5 in Kiss1ARH neurons compared to the control group (Figure 11D). Hence, the qPCR data verified that in the sgTrpc5-targeted mice we can selectively reduce Trpc5 gene expression in targeted cells.

CRISPR mutagenesis of Trpc5 channels in Kiss1ARH neurons.

A, structure of AAV1-FLEX-SaCas9-U6sgTrpc5-exon2. Exon 2 of Trpc5 is denoted with guide sequence highlighted in red, the PAM is underlined. B. Structure of AAV1-FLEX-SaCas9-U6sgTrpc5-exon7. Exon 7 of Trpc5 is denoted with guide sequence highlighted in red, the PAM is underlined. C1, image of coronal section through the ARH from Kiss1-Cre::Ai32 mouse with dual co-injections of AAV-DIO-mCherry and AAV1-FLEX-SaCas9-U6-sgTrpc5. Scale = 200 µm. C2, C3, higher power overlays of epifluorescence (EYFP & mCherry) images with recording pipette patched onto a Kiss1ARH-Cre:mCherry cell (C3). Scale = 40 µm. D, quantitative PCR measurements of Trpc5 transcripts from 10 cell neuronal pools (three pools from each animal) in double sgRNA mutagenesis of Trpc5 (second sgRNA against pore forming region) in Kiss1ARH neurons. Primers were targeted to 1st (left panel) or 2nd (right panel) guide, respectively (Unpaired t test for 1st, t(10) = 10.67, ****p < 0.0001; for 2nd, t(10) = 10.79, ****p < 0.0001). Data are expressed as mean ± SEM, with data points representing the number of animals.

As predicted, mutagenesis of Trpc5 channels in Kiss1ARH neurons significantly attenuated the slow EPSP (Figures 12A, B, C) such that the postsynaptic excitation was reduced to a “trickle” of action potential firing. What we would not have predicted is that the double sgRNA mutagenesis of Trpc5 channels in Kiss1ARH neurons significantly hyperpolarized the resting membrane potential by 7 mV (Figure 12D). Moreover, the rheobase (minimum current required to induce firing) significantly increased by ∼13 % in females bearing the sgTrpc5 double mutagenesis (Figure 12E). The firing frequency versus injected current (F-I) curve for sgTrpc5 double mutagenesis Kiss1ARH neurons was also significantly attenuated (Figure 12F). In agreement with these experimental findings, current ramp simulations of our mathematical model confirmed that a decrease of the TPRC5 conductance increased the rheobase and reduce the firing activity of Kiss1ARH neurons (Figures 12G, H). In our model, this contribution of TRPC5 channels to cell excitability is attributed to intracellular calcium levels which activate TRPC5 channels independently of NKB stimulation. Finally, we employed our mathematical model to further investigate the transition from synchronous firing driven by NKB release and TRPC5 channel activation to burst firing generated by E2-mediated upregulation of endogenous conductances. As we varied the conductance of TRCP5 channels and GIRK channels under saturating extracellular signals (e.g., NKB and Dyn), we find that balance between the two conductances controls the neuronal excitability in the OVX parameterized model, with synchronous firing eliminated when TRPC5 conductance is low (red triangle) relative to the GIRK conductance (Figure 13A). Furthermore, the burst firing of the OVX+E2 parameterized model was supported by elevated h- and Ca 2+-currents (Figure 13B) as well as by the high conductance of Ca2+ channels relative to the conductance of TRPC5 channels (Figure 13C). Importantly, the simulated burst firing was analogous to the phasic burst firing recorded ex vivo in Kiss1ARH neurons (Figure 1A).

Double CRISPR mutagenesis of Trpc5 attenuates slow EPSP, increases rheobase and shifts the F-I curve.

A, high-frequency photo-stimulation (20 Hz) generated slow EPSP in Kiss1ARH neuron from ovariectomized, control mouse. Red trace is slow EPSP after low-pass filtering. B, slow EPSP in Kiss1ARH neuron from OVX, double sgTrpc5 -targeted (Trpc5 mut) mouse. C, summary of the effects of Trpc5 mutagenesis on slow EPSP amplitude in female mice. Unpaired t-test, t(27) = 5.916, ****p < 0.0001. Data are expressed as mean ± SEM, with data points representing the number of cells. D, double sgRNA mutagenesis of Trpc5 channels in Kiss1ARH neurons significantly increased the RMP (control: -64.5 ± 1.4 mV versus Trpc5 mut, 2: -71.1 ± 1.2 mV, Unpaired t-test, t(73) = 3.524, ***p = 0.0007). E, current ramp showing the increased rheobase in Kiss1ARH neuron from Trpc5 mut mice (control: 31.1 ± 1.2 pA, n = 31, versus Trpc5 mut , 35.3 ± 1.0 pA, n = 33, Unpaired t-test, t(62) = 2.777, **p = 0.0073). F, firing frequency vs. current (F-I) curves for control versus Trpc5 mut (two-way ANOVA: main effect of treatment, F(1,52) = 13.03, p = 0.0007, main effect of time, F(8,416) = 291.3, p < 0.0001, and interaction, F(8,416) = 6.254, p < 0.0001; Control, n = 26, Trpc5 mut, n = 29; post hoc Bonferroni test, *<0.05; **<0.01 and ***p < 0.005, respectively). G, model simulations of the effects of a current ramp (50 pA/sec) for OVX (left panel) and OVX female with reduced (muted) TRPC5 conductance (right panel), and H, the associated firing frequency vs. current curves. In the latter case the TRCP5 conductance was halved, which is a conservative estimation of the CRISPR state in which the Trpc5 is much more mutated in Kiss1ARH neurons (Figure 11).

Computational modeling of a Kiss1ARH neuron in the OVX and OVX + E2 state demonstrates its distinct dynamic responses.

A model of the Kiss1ARH neuron was developed and calibrated using molecular data and electrophysiological recordings of Kiss1ARH neurons from OVX and OVX+E2 mice. A. Simulations of the OVX-parameterized model demonstrating high frequency activity in response to saturating levels of NKB stimulation. The balance between GIRK and TRCP5 conductance controls the response of the neuron to NKB stimulation, with neuronal response eliminated when TRPC5 conductance is low (red triangle) relative to the GIRK conductance. B. The OVX+E2 parameterized models demonstrate sustained burst firing activity. The bursting activity that is supported by elevated h- and Ca 2+-currents (red square) as observed in OVX+E2 mice. C. In the OVX+E2 state, burst firing activity is also supported by high conductance of HVA Ca2+ channels relative to the conductance of TRPC5 channels. Representative points in the parameter space giving rise to burst firing activity are marked with red squares, whereas red triangles are used for points resulting in regular spiking. The black line separates these two regions of activity.

Discussion

E2 appears to play a critical role in transitioning the glutamatergic/peptidergic Kiss1ARH neurons from a high frequency firing mode during synchronization, which is dependent on NKB-driven activation of TRPC5 channels, to a short bursting mode that would facilitate glutamate release. E2 decreased the expression of the peptide neurotransmitters NKB (kisspeptin and dynorphin) and TRPC5 channels but increased the mRNA expression of Vglut2, voltage-activated calcium channels and Hcn channels that contribute to burst firing and glutamate release from the Kiss1ARH neurons. We determined that the increase in mRNA expression of the HVA calcium channels translated into a significant increase in whole-cell current with all of the calcium channels contributing proportionally. Most importantly the kinetics of activation and inactivation were unaltered with E2 treatment, which indicates that other post-translation modifications were not affecting channel activity. Surprisingly and somewhat counter intuitive, the BK α1 subunit was also upregulated, but based on our modeling the rapid repolarization of the Kiss1ARH neurons (i.e., the fast AHP) would facilitate higher frequency of action potential firing. Moreover, our modeling confirmed that TRPC5 channels, which generate the slow EPSP (a.k.a., plateau potential in other CNS neurons), are vital for initiating and sustaining synchronous firing of Kiss1ARH neurons, while subsequent activation of GIRK channels repolarizes Kiss1ARH neurons. E2 treatment of ovariectomized females decreased both Trpc5 and Girk2 channel mRNA expression, which in our model correlated with the reduction in sustained high frequency firing of Kiss1ARH neurons. Therefore, the synchronous high frequency firing of Kiss1ARH neurons in a low E2 milieu correlates with the pulsatile release of GnRH (LH from the pituitary gland), whereas the transition to burst firing in the presence of high circulating levels of E2 (e.g., proestrus) would facilitate the GnRH (LH) surge through its glutamatergic synaptic connection with Kiss1AVPV/PeN neurons (Lin et al., 2021; Qiu J. et al., 2016).

Core calcium conductances underlying synchronous and burst firing of Kiss1ARH neurons

TRPC5 channels are highly expressed in Kiss1ARH neurons (Figure 11), and TRPC5 channels are essentially ligand-activated calcium channels with a high permeability to calcium (PCa/PNa = 9:1) (Venkatachalam and Montell, 2007). In general, mammalian TRPC channels are activated by both G protein-coupled receptors and receptor tyrosine kinases (Ambudkar and Ong, 2007; Clapham, 2003), and are one of the major downstream effectors activated by glutamate binding to group I metabotropic glutamate receptors (mGluR1 and mGluR5) in CNS neurons (Bengtson et al., 2004; Berg et al., 2007; Faber et al., 2006; Tozzi et al., 2003). In substantia nigra dopamine neurons mGluR1 agonists induce a current that exhibits the tell-tale double-rectifying current-voltage plot of TRPC channel activation (Tozzi et al., 2003), similar to what we see with the effects of the NKB agonist senktide in Kiss1ARH neurons (Qiu Jian et al., 2021). Both mGluR1 and TacR3 are Gq-coupled to phospholipase C (PLC) activation which leads to hydrolysis of phosphatidylinositol 4,5-bisphosphate (PIP2) to diacylglycerol (DAG) and inositol 1,4,5 triphosphate (IP3), which is involved in channel activation (Birnbaumer, 2009). TacR3 (NKB) signaling has additional consequences since many K+ (e.g., GIRK, KCNQ) channels are dependent on PIP2 for channel opening, and depletion of PIP2 by PLC leads to channel closure (Brown and Passmore, 2009) (Zhang C. et al., 2013) (Whorton and MacKinnon, 2011) (Zheng et al., 2022). Therefore, depletion of PIP2 by NKB signaling would further facilitate the sustained firing of Kiss1ARH neurons during synchronization.

A plateau potential has been characterized in hippocampal and cortical neurons (Arboit et al., 2020; Zhang Z. et al., 2011) as such these neurons express biophysical properties that allow them to continue to persistently fire even after a triggering synaptic event has subsided (Zylberberg and Strowbridge, 2017). The persistent firing activity of these neurons is linked to ICAN (Zylberberg and Strowbridge, 2017), and TRPC5 channels appear to be responsible for the ICAN (Zhang Z. et al., 2011). With TacR3 activation in Kiss1ARH neurons there is an influx of Ca+2 through TRPC5 channels leading to greater build-up of [Ca2+]i that facilitates the opening of more TRPC5 channels in a self-sustaining (autocatalytic) manner (Qiu J. et al., 2016). Using the fast intracellular calcium chelator BAPTA, which has been shown to robustly inhibit TRPC5 channel activation in heterologous cells (Blair et al., 2009), we abolished the slow EPSP and persistent firing of Kiss1ARH neurons following optogenetic stimulation (Qiu Jian et al., 2021). Moreover, HVA calcium channel blockers attenuated the generation of the slow EPSP (Figure 3) so it appears that they also contribute to the ICAN since calcium influx via both LVA and HVA calcium channels can also facilitate TRPC5 channel opening in Kiss1ARH neurons. In the ovariectomized female, treatment with E2 upregulated Cav1.2, Cav2.1, Cav2.2, and Cav2.3 mRNA by 1.5- to 2-fold and Cav3.1 mRNA expression by ∼3-fold. Hence, E2 significantly increased whole-cell calcium currents Kiss1ARH neurons, which greatly enhanced the excitability and contributed to the burst firing of Kiss1ARH neurons (Figure 1). However, the amplitude of the slow EPSP with E2 treatment is only ∼25% of the amplitude in the ovariectomized state (Qiu J. et al., 2016). Therefore, there appears to be a physiologic transition of Kiss1ARH neurons from the slow EPSP firing mode in the OVX state to the burst firing mode in the presence of E2, which has important physiologic ramifications as discussed below.

TRPC5 and GIRK channels are vital for synchronization of Kiss1ARH neurons

Recently, Tian et al. demonstrated that TRPC4, a close homolog of TRPC5 sharing ∼64 percent homology, is a “coincidence detector” of neurotransmission by both Gq/11 and Gi/o-coupled receptors in lateral septal (LS) neurons (Tian et al., 2022). In whole-cell recordings of LS neurons, TRPC 4 channels mediate a strong depolarizing plateau potential that in contrast to TRPC5 channel activation in Kiss1ARH neurons, abrogates action potential firing as a result of a depolarization block. In many instances the plateau potential in LS neurons is followed by an AHP, which is dependent on the activation of Gi/o-coupled receptors. In contrast, we have not observed an AHP in Kiss1ARH neurons following the slow EPSP (Qiu J. et al., 2016). Tian and colleagues showed that the depolarizing plateau in LS neurons is codependent on activation of both Gq/11-coupled mGluR1 glutamate receptors and Gi/o-coupled γ-aminobutyric acid type B receptors, the latter activating GIRK channels. Moreover, the firing patterns in LS neurons encodes information about the relative strengths of these contrasting inputs (i.e., Gq/11 versus Gi/o) such that only mGluR1 produces weak depolarization accompanied by increased firing of LS neurons, whereas pure GABAB receptor activation hyperpolarizes the cells and abrogates firing activity. Coincident input of both mGluR1 and GABAB receptors results in a brief burst of action potentials followed by a pause in firing, and both the pause duration and firing recovery patterns reflect the relative strengths of Gq/11 versus Gi/o inputs. Importantly, Tian and colleagues computationally simulated these various scenarios with computational modeling, and similar to our modeling, utilized only TRPC4 and GIRK channels. A notable difference between the Kiss1ARH neurons and the LS neurons is that dynorphin binds to kappa-opioid receptors to open GIRK channels in the nerve terminal of Kiss1ARH neurons (presynaptic effect) (Qiu J. et al., 2016), whereas GABA binds to some GABAB receptors to hyperpolarize LS neurons (Tian et al., 2022). Therefore, although the high frequency firing activity of both LS and Kiss1ARH neurons can be modeled around TRPC and GIRK channels, the generated firing patterns are dramatically different based on the timing (co-incident activation of TRPC4 and GIRK channels in LS) and localization of the kappa opioid-coupled GIRK channels in the axon terminal of Kiss1ARH neurons. Moreover, E2-treated, ovariectomized females show a significant down-regulation of both Trpc5 and Girk2 mRNA expression in Kiss1ARH neurons (Figure 10), which is important for the physiological transitioning as described below.

Interestingly, the hypothalamic A12 (ARH) dopamine neurons show a rhythmic “oscillatory” firing behavior that transitions to a tonic firing mode with synaptic input from Thyrotropin-releasing hormone (TRH) neurons (Lyons D.J. et al., 2010) or feedback by circulating prolactin, released by pituitary lactotrophs (Lyons D. J. et al., 2012). A more recent paper has revealed that TRPC5 channels mediate the plateau potential and tonic firing in A12 dopamine neurons in response to prolactin (Blum et al., 2019). Similar to our findings with CRISPR deletion of Trpc5 in Kiss1ARH neurons (Figures 11 and 12), conditional knockout of Trpc5 in dopamine neurons abrogated the prolactin-induced plateau potential and tonic firing. Although Blum and colleagues did not model the oscillatory firing or tonic firing of the A12 dopamine neurons, their findings are consistent with our results showing that the activation of TRPC5 channels underlies the slow EPSP (plateau potential) and sustained firing.

Contribution of endogenous K+ channels to synchronized and burst firing

Beyond the ligand-gated (e.g., baclofen) GIRK channels, there are endogenous K+ channels that help sculpt the firing activity of kisspeptin neurons. We focused on the calcium-activated K+ channel family: the large-conductance, calcium-activated potassium (BK, also called BKCa, KCa1.1, MaxiK, Slo), small conductance, calcium-activated K+ (SK1, SK2, SK3) (Bond et al., 2005), and the K+ channels underlying the M-current (KCNQ, Kv7.1-7.5) (Brown and Passmore, 2009), which mediate the fast afterhyperpolarization (AHP), the intermediate AHP/slow AHP, respectively (Andrade et al., 2012) BK channels are gated by both voltage and cytoplasmic calcium and sculpt action potential firing in CNS neurons (Blatz and Magleby, 1987; Marty, 1989; Sah P., 1996; Storm, 1990). Indeed, BK channels have been shown to mediate rapid spike repolarization—i.e., the fast AHP in hippocampal CA1 pyramidal neurons (Lancaster and Nicoll, 1987; Storm, 1987). Blockade of BK channels in CA1 neurons attenuates the initial discharge frequency in response to current injection, which is attributable to suppression of the BK channel-dependent rapid spike repolarization (Lancaster and Nicoll, 1987; Storm, 1987). Blockade of BK channels is thought to increase inactivation of the spike-generating transient Na+ current and activate more of the slower K+ currents, thereby enhancing refractoriness and reducing excitability during the immediate aftermath of the first action potential (Shao et al., 1999). Thus, BK channels facilitate high-frequency burst firing of CA1 neurons. Furthermore, extracellular field recordings confirmed that BK channels contribute to high-frequency burst firing in response to excitatory synaptic input to distal dendrites in CA1 neurons (Gu et al., 2007). Therefore, BK channels appear to play an important role for early high-frequency, rapidly adapting firing in hippocampal CA1 pyramidal neurons, thus promoting the type of bursting that is characteristic of these cells in vivo during behavior. Based on our in vitro electrophysiological recordings and computational modeling we see a similar physiological phenomenon in Kiss1ARH neurons (Figure 8). Not only does E2 increase the mRNA expression of Kcnma1 (Figure 8G), but also the maximum BK (IbTx-sensitive) current by 3-fold, which would facilitate rapid repolarization during burst firing and promote glutamate release similar to hippocampal CA1 neurons.

In contrast to BK channel expression, E2 did not affect the mRNA expression of SK3 channel mRNA. SK channels underlie the apamin-sensitive component of the medium duration AHP and are responsible for repolarization following a burst of action potentials (Andrade et al., 2012; Bond et al., 2005). The activation of SK channels is voltage-independent, but SK channels have a higher affinity for Ca2+ than BK channels (Andrade et al., 2012). SK channels are tightly coupled (within 100 nm) to L-type Ca2+ channels, and BK channels (within 30 nm) are tightly coupled to N-type Ca2+ channels in hippocampal CA1 pyramidal neurons (Marrion and Tavalin, 1998). The determination of the proximity of SK and BK channels to HVA calcium channels in kisspeptin neurons will require cell-attached patch recordings. However, in Kiss1ARH neurons the SK channels may come into play during a short burst of action potentials but would become overwhelmed with the higher frequency synchronized, sustained firing as a result of NKB stimulation (Qiu J. et al., 2016). As discussed above what limits the synchronized firing of Kiss1ARH neurons is the activation of GIRK channels.

Finally, the calcium-activated slow AHP probably plays a critical role in the repolarization of Kiss1ARH after burst firing. The molecular identification of the channels mediating the slow AHP has long been an area of intense investigation (Andrade et al., 2012; Vogalis et al., 2003). A critical feature of the slow AHP is that it activates very slowly (hundreds of milliseconds) long after the rise in cytoplasmic Ca2+ (Sah P. and Clements, 1999) so an intermediate Ca2+ signaling molecule has long been thought to be involved. Indeed, Tzingounis and colleagues (Tzingounis A.V. et al., 2007) have provided compelling evidence that the diffusible calcium sensor hippocalcin is the critical intermediate molecule involved in Ca2+ sensing. The slow AHP is abrogated in hippocalcin KO mice, and transfection of hippocalcin into cultured hippocampal neurons generates a pronounced slow AHP in response to a depolarizing stimulus (Tzingounis A.V. et al., 2007). Importantly, the slow AHP is activated by Ca2+ with an EC50 ≈ 300 nM, which is well within the operational range of hippocalcin but well below that of calmodulin (Andrade et al., 2012). Finally, two seminal papers from Tzingounis and colleagues demonstrate that KCNQ 2, 3 channels are responsible for the slow AHP in hippocampal dentate neurons (Tzingounis A. V. and Nicoll, 2008), and KCNQ 5 channels are responsible for the slow AHP in CA3 neurons (Tzingounis A. V. et al., 2010). Moreover, the KCNQ channel blocker XE991 attenuates the slow AHP in CA3 neurons (Tzingounis A. V. and Nicoll, 2008). Based on these seminal findings we investigated the role of the KCNQ channels, which “classically” underlie the M-current in Kiss1ARH. The M-current was first identified in Kiss1ARH neurons by Conde and Roepke (Conde and Roepke, 2019), and presently we found that Kcnq2 mRNA is expressed in Kiss1ARH neurons and up-regulated by E2, which translated into a greater M-current in Kiss1ARH neurons (Figure 9). Incorporating the M-current into our computational model indeed supports our hypothesis that M-current, along with ISK and IBK, contributes to membrane repolarization after burst firing (Figure 9G). Importantly the slow AHP, as opposed to the fast AHP (BK) and medium AHP (SK) is highly regulated by multiple neurotransmitters (Andrade et al., 2012), which sets the stage for further modulation of the slow EPSP in Kiss1ARH neurons.

Importance of E2-driven physiological transitioning

Since the expression of the peptide neurotransmitters in Kiss1ARH neurons are down-regulated by E2, the Kiss1ARH neurons are believed to be under “inhibitory” control by E2 and are important for “negative-feedback” regulation of GnRH release (Lehman et al., 2013; Navarro V. M. et al., 2009; Smith et al., 2005) (Rance Naomi E. and Young, 1991) (Rance N.E., 2009). However, our past (Gottsch et al., 2011) and current findings document that these Kiss1ARH neurons express HVA and LVA calcium and HCN (pacemaker) channels and are excited by co-released glutamate from neighboring Kiss1ARH neurons, which indicates that these neurons have pacemaker electrophysiological properties similar to other CNS neurons (Bal and McCormick, 1993; Lüthi and McCormick, 1998). Additionally, in contrast to the neuropeptides, E2 increases Slc17a6 (Vglut2) mRNA expression, in addition to mRNA for the voltage-activated calcium and HCN channels, and increases glutamate release onto Kiss1AVPV/PeN neurons (Qiu J. et al., 2018). Interestingly, Slc17a6 mRNA expression in Kiss1ARH neurons and the probability of glutamate release are decreased along with the neuropeptides in intact versus castrated males (Nestor et al., 2016), which indicates a profound sex difference in the glutamate signaling by Kiss1ARH neurons (Nestor et al., 2016; Qiu J. et al., 2018). Obviously, in the male there is no preovulatory LH surge so there is no need for excitatory glutamatergic input to the few Kiss1AVPV neurons in the male.

In females, conditional knockout of Slc17a6 in Kiss1 neurons abrogates glutamate release from Kiss1ARH neurons (Qiu J. et al., 2018). Kiss1AVPV/PeN neurons do not express Slc17a6 and do not release glutamate. Within the Kiss1ARH neurocircuitry the lack of glutamate transmission does not diminish the slow EPSP in ovariectomized females (Qiu J. et al., 2018). Indeed, a recent publication from the Herbison lab demonstrates that glutamate generates small “synchronizing” events that are dependent on the ionotropic receptors (Han et al., 2023), but the fast neurotransmitter is unable to support the sustained firing (i.e., slow EPSP) that is necessary for peptide release and synchronization of the KNDy network. Rather, we postulate that glutamate neurotransmission is more important for excitation of Kiss1AVPV/PeN neurons and facilitating the GnRH (LH) surge with high circulating levels of E2 when peptide neurotransmitters are at a nadir and glutamate levels are high in female Kiss1ARH neurons. Indeed, low frequency (5 Hz) optogenetic stimulation of Kiss1ARH neurons, which only releases glutamate in E2-treated, ovariectomized females (Qiu J. et al., 2016), generates a surge-like increase in LH release during periods of optical stimulation (Lin et al., 2021; Voliotis et al., 2021). In a subsequent study optical stimulation of Kiss1ARH neuron terminals in the AVPV at 20 Hz, a frequency commonly used for terminal stimulation in vivo, generated a similar surge of LH (Shen et al., 2022). Additionally, intra-AVPV infusion of glutamate antagonists, AP5+CNQX, completely blocked the LH surge induced by Kiss1ARH terminal photostimulation in the AVPV (Shen et al., 2022). Therefore, there appears to be a clear role for glutamatergic transmission from the Kiss1ARH to Kiss1AVPV/PeN neurons in amplifying the LH surge in the female mouse. Finally it is important to keep in mind that even in the presence of high physiological levels of E2, the mRNA expression of Tac2 is many-fold higher than Kiss1 (Figure 10), which is essential for NKB maintaining synchronous firing of Kiss1ARH neurons, albeit at a lower frequency, across all physiological states (Qiu J. et al., 2016). Indeed, there is a progressive change from a strictly pulsatile pattern of GnRH in the hypophyseal portal circulation to one containing both pulsatile and non-pulsatile components during the development of the GnRH surge in the ewe (Evans, Dahl, Mauger, & Karsch, 1995; Evans, Dahl, Mauger, Padmanabhan, et al., 1995), and a pulsatile mode of LH secretion during the preovulatory LH surge is also evident in other species including humans (Rossmanith et al., 1990). Therefore, we believe that our cellular molecular and electrophysiological findings in combination with our computational modelling provide a foundation for understanding the complex role of Kiss1ARH neurons in controlling fertility in the mammal. Finally, our model provides the first comprehensive biophysical description of the conductances underlying the neuronal activity of KissARH neurons, which can serve as a basis for future computational modelling of the KissARH neuronal network and its interactions with other brain regions involved in the complex regulation of mammalian female reproduction.

Methods and Materials

Animals

All the animal procedures described in this study were performed in accordance with institutional guidelines based on National Institutes of Health standards and approved by the Institutional Animal Care and Use Committee at Oregon Health and Science University (OHSU).

Mice

Kiss1Cre transgenic female mice version 2 (Padilla et al., 2018) were selectively bred at OHSU. They also were crossed with heterozygous Ai32 mice (RRID:IMSR_JAX:024109, C57BL/6 background) which carry ChR2 (H134R)–EYFP gene in their Gt(ROSA)26Sor locus (Madisen et al., 2012). All animals were maintained under controlled temperature and photoperiod (lights on at 0600h and off at 1800h) and given free access to food (Lab Diets 5L0D) and water. Where specified, Kiss1Cre mice received viral injections to express channelrhodopsin 2 (ChR2) in Kiss1ARH neurons, fourteen to twenty-one days prior to each experiment as described (Qiu J. et al., 2016). Some of the females were ovariectomized seven days prior to an experiment. Each animal was injected on day 5 following OVX with 0.25 μg E2 or vehicle, followed on day 6 with 1.50 μg E2 or vehicle and used for experiments on day 7 (Bosch et al., 2013).

AAV delivery to Kiss1Cre mice

Fourteen to twenty-one days prior to each experiment, the Kiss1Cremice (>60 d old) received bilateral ARH injections of a Cre-dependent adeno-associated viral (AAV; serotype 1) vector encoding mCherry (AAV1-Ef1α-DIO-mCherry) or AAV1 vectors designed to encode SaCas9 and single-guide RNAs (sgRNAs) (See the SaCas9 section for specifics on the sgRNA design). Using aseptic techniques, anesthetized female mice (1.5% isoflurane/O2) received a medial skin incision to expose the surface of the skull. The glass pipette with a beveled tip (diameter = 45 μm) was filled with mineral oil, loaded with an aliquot of AAV using a Nanoject II (Drummond Scientific). ARH injection coordinates were anteroposterior (AP): −1.20 mm, mediolateral (ML): ± 0.30 mm, dorsoventral (DV): −5.80 mm (surface of brain z = 0.0 mm); 500 nl of the AAV (2.0×1012 particles/ ml) was injected (100 nl/min) into each position, and the pipette left in place for 10 min post-injection, then slowly retracted from the brain. The skin incision was closed using Vetbond (3M) and each mouse received analgesia (Rimadyl, 4-5 mg/kg, s.c.).

Generation of AAV1-FLEX-SaCas9-U6-sgTrpc5

The generation of AAV1-FLEX-SaCas9-U6-sgTrpc5 viruses were done at the University of Washington using published methods (Gore et al., 2013; Hunker et al., 2020). The constructs of sgRNAs for Trpc5 (AAV1-FLEX-SaCas9-U6sgTrpc5) were designed to target exon2 and exon7, respectively (Figure 11A and B). To achieve Trpc5 mutagenesis in Kiss1ARH neurons, Kiss1Cre mice were co-injected with AAV1-DIO-mCherry, AAV1-FLEX-SaCas9-U6-sgTrpc5-exon2, and AAV1-FLEX-SaCas9-U6-sgTrpc5-exon7 at a ratio of 10%, 45%, and 45%, respectively. Control animals were co-injected with AAV1-FLEX-SaCas9-U6-sgRosa26 and AAV1-DIO-mCherry at a ratio of 90% and 10%, respectively. AAV1-DIO-mCherry was co-injected with Cas9 vectors to confirm the targeting of injections and visualize the infected Kiss1ARH neurons. Each animal was used for experiments three weeks after the viral injection, with OVX performed one week prior to the experiments.

Visualized whole-cell patch recording

Electrophysiological and optogenetic studies were made in coronal brain slices (250 μm) containing the ARH from AAV1-EF1α-DIO-mCherry injected Kiss1Cre:GFP or Kiss1-Cre:EYFP::AI32 mice, which were vehicle-treated OVX, and E2-treated OVX females 10 weeks and older as previously described (Qiu J. et al., 2016; Qiu J. et al., 2018). Whole-cell patch recordings were performed in voltage-clamp and current-clamp as previously described (Qiu J. et al., 2018) using an Olympus BX51 W1 fixed stage scope out-fitted with epifluorescence and IR-DIC video microscopy. Patch pipettes (A-M Systems; 1.5 μm outer diameter borosilicate glass) were pulled on a Brown/Flaming puller (Sutter Instrument, model P-97) and filled with the following solution (unless otherwise specified): 128 mM potassium gluconate, 10 mM NaCl, 1 mM MgCl2, 11 mM EGTA, 10 mM HEPES, 2 mM ATP, and 0.25 mM GTP adjusted to pH 7.3 with KOH; 295 mOsm. Pipette resistances ranged from 3.5–4 MΩ. In whole-cell configuration, access resistance was less than 30 MΩ; the access resistance was 80% compensated. The input resistance was calculated by measuring the slope of the I-V relationship curve between −70 and −50 mV. Standard whole-cell patch recording procedures and pharmacological testing were performed as previously described (Qiu J. et al., 2003; Qiu J. et al., 2014). Electrophysiological signals were digitized with a Digidata 1322A (Axon Instruments) and the data were analyzed using p-Clamp software (Molecular Devices, Foster City, CA). The liquid junction potential was corrected for all data analysis.

For optogenetic stimulation, a light-induced response was evoked using a light-emitting diode (LED) 470 nm blue light source controlled by a variable 2A driver (ThorLabs, Newton, NJ) with the light path directly delivered through an Olympus 40x water-immersion lens. For high-frequency (20 Hz) stimulation the length of stimulation was 10 seconds (light intensity 0.9 mW and pulse duration, 10 ms) (Qiu J. et al., 2016).

For studying the firing pattern of Kiss1ARH neurons from OVX and E2-treated OVX mice, the electrodes were filled with an internal solution consisting of the following (in mM): 120 KMeSO4, 5 Na2 phosphocreatine, 10 KCl, 2 MgSO4, 10 HEPES, 4 K2ATP, and 0.25 GTP adjusted to pH 7.3 with KOH; 290 mOsm. The bath solution as described (Ashhad and Feldman, 2020) consisted of (in mM): 124 NaCl, 3.5 KCl, 25 NaHCO3, 0.5 NaH2PO4 and 30 D-glucose,1 MgSO4, 1.5 CaCl2, gassed with 95% O2 /5% CO2 (pH 7.4, 310 mOsm).

For studying the activation/inactivation characteristics of the Ca2+ current, the electrodes were filled with an internal solution as described (Qiu J. et al., 2003) consisting of the following (in mM): 100 Cs+ gluconate, 20 TEA-Cl, 10 NaCl, 1 MgCl2, 10 HEPES, 11 EGTA, 1 ATP, 0.25 GTP, the pH was adjusted to 7.3 with CsOH at 300 mOsm. The bath solution as described (Zhang X. B. and Spergel, 2012) consisted of (in mM) 117.5 NaCl, 25 NaHCO3, 1.25 NaH2PO4, 10 TEA-Cl, 2 CaCl2, 1 MgCl2, 20 sucrose and 5 glucose, gassed with 95% O2 /5% CO2 (pH 7.4, 309 mOsm) and supplemented with 1 μM TTX, 10 μM CNQX, 50 μM AP5 and 100 μM picrotoxin. The effects of estrogen treatment on the peak current, peak current density, and activation/inactivation characteristics of calcium current were measured. Activation curves were fitted by the Boltzmann equation: I/Imax=1/{1+exp[V1/2-Vs)/k]}, where I is the peak current at the step potential Vs, Imax is the peak current amplitude, V1/2 is the step potential yielding half-maximum current, and k is the slope factor. Inactivation curves were fit with the Boltzmann equation: I/Imax = 1 - 1/ {1 + exp [(VH - V1/2)/k]}, where I is the peak current at the step potential VH, Imax is the peak current amplitude, V1/2 is the step potential at which half the current is inactivated, and k is the slope factor.

To record M-currents, pipettes were filled with an internal solution consisting of 10 mM NaCl, 128 mM K-gluconate, 1 mM MgCl, 10 mM HEPES, 1 mM ATP, 1.1 mM EGTA, and 0.25 mM GTP (pH 7.3; 290 mOsm). During voltage-clamp, we employed a standard deactivation protocol (Conde and Roepke, 2019; Roepke et al., 2011) to measure potassium currents. This involved 500-ms voltage steps ranging from –30 to –75 mV in 5-mV increments, following a 300-millisecond prepulse to –20 mV. The amplitude of the M-current relaxation or deactivation was quantified as the difference between the initial (<10 ms) and sustained current (>475 ms) of the current trace.

The bath solution for whole-cell recording of BK, SK and M currents was aCSF supplemented with 1 μM TTX, 10 μM CNQX, 50 μM AP5 and 100 μM picrotoxin.

Electrophysiological solutions/drugs

A standard artificial cerebrospinal fluid (aCSF) was used (Qiu J. et al., 2003; Qiu J. et al., 2010). All drugs were purchased from Tocris Bioscience unless otherwise specified. 1 mM TTX (Alomone Labs), 50 mM DL-2-Amino-5-phosphonopentanoic acid sodium salt (AP5), 10 mM 6-Cyano-7-nitroquinoxaline-2,3-dione disodium (CNQX), 100 mM picrotoxin, 10 mM Nifedipine (Sigma), 2 mM ω-conotoxin GVIA (ConoGVIA; Alomone Labs), 0.5 mM ω-conotoxin MVIIC (ConoMVIIC; Alomone Labs, 50 μM ω-agatoxin IVA (AgaIVA; Alomone Labs), 50 μM SNX-482 (Alomone Labs), 10 mM TTA-P2 (TTAP2; Alomone Labs), 200 mM CdCl2 (Cd2+; Sigma), 40 mM XE991 (Alomone Labs), 100 μM Iberiotoxin (Alomone Labs), 500 μM Apamin (Alomone Labs), 100 mM glutamic acid and 1 mM senktide. Stocks (1000×) were prepared in dimethylsulfoxide DMSO (picrotoxin, TTA-P2 and senktide) or water (TTX, AP5, CNQX, ConoGVIA, ConoMVIIC, AgaIVA, SNX-482, CdCl2 and glutamic acid) and stored at −20°C. Aliquots of the stock solutions were stored as appropriate until needed.

Cell harvesting of dispersed Kiss1Cre neurons and real-time quantitative PCR (qPCR)

Cell harvesting and qPCR was conducted as previously described (Bosch et al., 2013). The ARH was microdissected from basal hypothalamic coronal slices obtained from female Kiss1Cre version 2 mice (Padilla et al., 2018) (n = 4-6 animals/group). The dispersed cells were visualized, patched, and then harvested (10 cells/tube) as described previously (Bosch et al., 2013). Briefly, ARH tissue was incubated in papain (7 mg/ml in oxygenated aCSF) for 50 min at 37° C then washed 4 times in low Ca2+ aCSF and two times in aCSF. For cell dispersion, Pasteur pipettes were flame polished to decreasing tip sizes and gentle trituration used to disperse the neurons onto a glass bottom dish. The plated cells were bathed in oxygenated aCSF using a peristaltic pump to keep the cells viable and clear of debris. Healthy cells with processes and a smooth cell membrane were harvested. Pipettes (World Precision Instruments; 1.5 μm outer diameter borosilicate glass) were pulled on a Brown/Flaming puller (Sutter Instrument, model P-87) to a 10 µm diameter tip. The cells were harvested using the XenoWorks Microinjector System (Sutter Instruments, Navato, CA) which provides negative pressure in the pipette and fine control to draw the cell up into the pipette. Cell pools were harvested and stored at -80°C. All cell pools were DNAse-treated using DNase1. cDNA synthesis was performed as previously described (Bosch et al., 2013).

Primers for the genes that encode for low and high voltage-gated calcium channels, TRPC5, Vglut2, large conductance calcium-activated K+ (BKα) channels, small conductance calcium-activated K+ (SK3) channels and GAPDH were designed using Clone Manager software (Sci Ed Software) to cross at least one intron-exon boundary and optimized as previously described using Power Sybr Green method (Bosch et al., 2013). Real-time qPCR controls included neuronal pools without reverse transcriptase (-RT), hypothalamic RNA with RT (+) and without RT (-), as well as water blanks. Standard curves using ARH cDNA were utilized to determine the real-time PCR efficiency (E = 10(−1/m) – 1) (Biosystems, 2006; Pfaffl, 2001). Only primers resulting in efficiencies of 90-100% were used for analysis. Primer sequences, qPCR parameters and efficiency calculations are provided in Table 1.

mRNA expression analysis

qPCR was performed on a Quantstudio 7 Flex Real-Time PCR System (Applied Biosystems) using Power SYBR Green PCR Master Mix (Applied Biosystems) according to established protocols (Bosch et al., 2013). The comparative ΔΔCT method (Livak and Schmittgen, 2001; Pfaffl, 2001; Schmittgen and Livak, 2008) was used to determine values from duplicate samples of 4 µl for the target genes and 2 µl for the reference gene GAPDH in a 20 μl reaction volume containing 1x Power SYBR Green PCR Master Mix and 0.5 μM forward and reverse primers. Three to six 10-cell pools per animal were analyzed from 4-6 animals per group. The relative linear quantity was determined using the 2-ΔΔCT equation (Livak and Schmittgen, 2001; Pfaffl, 2001; Schmittgen and Livak, 2008). Relative mRNA expression level of target genes in Kiss1Cre neurons was obtained by comparing OVX Oil-treated controls to OVX E2-treated animals. The mean Δ CT for the target genes from the OVX Oil-treated control samples was used as the calibrator. The data were expressed as n-fold change in gene expression normalized to the reference gene GAPDH and relative to the calibrator.

Experimental design and Statistical analysis

For the whole-cell patch recording, pharmacological experiments, only one cell was recorded per slice. Two to three slices were analyzed from each Kiss1Cre mouse, with at least 3-5 mice contributing to each group. For cell harvesting of dispersed Kiss1Cre-YFP neurons and qPCR measurements, 10 cells per pool and 3-6 pools from each animal were used, unless otherwise specified. Statistical comparisons between two groups were performed using an unpaired two-tailed Student’s t-test. Comparisons between more than two groups were performed using the repeated measures, multifactorial ANOVA. If a significant interaction was encountered, we then moved to the one-way ANOVA, followed by the multiple range tests as specified in the appropriate figure legends. All data were analyzed using GraphPad Prism version 6. All data are presented as mean ± standard error of the mean (SEM). Differences were considered statistically significant if the probability of error was less than 5%.

Mathematical model and simulation of neuronal behavior

A mathematical model of the Kiss1ARH neuron was developed and calibrated based on our physiological findings. Here, we employed the Hodgkin-Huxley modelling approach (Hodgkin and Huxley, 1952). Accordingly, the equation describing the membrane potential Vm of the Kiss1ARH neuron is given by where Cm is the membrane capacitance and I is the sum of 12 ionic currents: INaT and INaP are the transient and persistent sodium currents, respectively; IA represents the A current; IM represents the M current; ISK and IBK are the potassium currents through the SK and BK channels respectively; Ih the HCN current; IT the T-type calcium current; ICa represents other calcium currents (L-, N-, P/Q-, and R-type); ITRPCS represents Calcium current throught the TRPC5 channel; IGIRK potassium current through the GIRK channels. Finally, Ileak represents the contribution of leak currents. We use the Hodgkin-Huxley formalism to model current dynamics and their dependence on the membrane potential. A complete specification of all currents along with the complete table of model parameters can be found in the supplementary material. Model simulations were conducted in Matlab using the built-in numerical solver (ode45; based on an explicit Runge-Kutta (4, 5) formula). The Matlab code is available at https://git.exeter.ac.uk/mv286/kiss1-arcuate-neuron-model.

eLife Assessment

This important study combined multiple approaches to gain insight into why rising estradiol levels, by influencing hypothalamic neurons, ultimately lead to ovulation. The experimental data were robust, but evidence for the conclusion that the findings explain how estradiol acts in the intact female was incomplete because of the lack of experimental conditions that better approximate physiological conditions. This work will be of interest to reproductive biologists working on ovarian biology and female fertility.

Reviewer #1 (Public Review):

Summary:

In this work, Qiu and colleagues examined the effects of preovulatory (i.e., proestrous or late follicular phase) levels of circulating estradiol on multiple calcium and potassium channel conductances in arcuate nucleus kisspeptin neurons. Although these cells are strongly linked to a role as the "GnRH pulse generator," the goal here was to examine the physiological properties of these cells in a hormonal milieu mimicking late proestrus, the time of the preovulatory GnRH-LH surge. Computational modeling is used to manipulate multiple conductances simultaneously and support a role for certain calcium channels in facilitating a switch in firing mode from tonic to bursting. CRISPR knockdown of the TRPC5 channel reduced overall excitability, but this was only examined in cells from ovariectomized mice without estradiol treatment. The manuscript has been substantially improved from the initial version by the addition of new experiments and clarification of important figures. Importantly, the overlap of data with previous reports from the same group has been corrected.

Strengths:

(1) Examination of multiple types of calcium and potassium currents, both through electrophysiology and molecular biology.

(2) Focus on arcuate kisspeptin neurons during the surge is relatively conceptually novel as the anteroventral periventricular nucleus (AVPV) kisspeptin neurons have received much more attention as the "surge generator" population.

(3) The modeling studies allow for direct examination of manipulation of single and multiple conductances, whereas the electrophysiology studies necessarily require examination of each current in isolation. Construction of an arcuate kisspeptin neuron model promises to be of value to the reproductive neuroendocrinology field.

Weaknesses:

A remaining weakness in this revised version of the manuscript is that the relevance of the CRISPR experiments is still rather tenuous given that the goal is to understand what happens in the estrogen-treatment condition, and these experiments were performed only in OVX mice. Similar concerns reflect that the computational model examining the effect of E2 infers multiple conductances based on qPCR data and an assumption that the conductances are directionally proportional to the level of gene expression, and then tunes these to the current recordings obtained from OVX mice, without a direct confirmation in OVX+E2 conditions that the model parameters accurately reflect the properties of these currents in the presence of estrogen.

Reviewer #2 (Public Review):

Summary:

Kisspeptin neurons of the arcuate nucleus (ARC) are thought to be responsible for the pulsatile GnRH secretory pattern and to mediate feedback regulation of GnRH secretion by estradiol (E2). Evidence in the literature, including the work of the authors, indicates that ARC kisspeptin coordinate their activity through reciprocal synaptic interactions and the release of glutamate and of neuropeptide neurokinin B (NKB), which they co-express. The authors show here that E2 regulates the expression of genes encoding different voltage-dependent calcium channels, calcium-dependent potassium channels and canonical transient receptor potential (TRPC5) channels and of the corresponding ionic currents in ARC kisspeptin neurons. Using computer simulations of the electrical activity of ARC kisspeptin neurons, the authors also provide evidence of what these changes translate into in terms of these cells' firing patterns. The experiments reveal that E2 upregulates various voltage-gated calcium currents as well as 2 subtypes of calcium-dependent potassium currents while decreasing TRPC5 expression (an ion channel downstream of NKB receptor activation), the slow excitatory synaptic potentials (slow EPSP) elicited in ARC kisspeptin neurons by NKB release and expression of the G protein-associated inward-rectifying potassium channel (GIRK). Based on these results, and on those of computer simulations, the authors propose that E2 promotes a functional transition of ARC kisspeptin neurons from neuropeptide-mediated sustained firing that supports coordinated activity for pulsatile GnRH secretion to a less intense burst-like firing pattern that could favor glutamate release from ARC kisspeptin. The authors suggest that the latter might be important for the generation of the preovulatory surge in females.

Strengths:

The authors combined multiple approaches in vitro and in silico to gain insights into the impact of E2 on the electrical activity of ARC kisspeptin neurons. These include patch-clamp electrophysiology combined with selective optogenetic stimulation of ARC kisspeptin neurons, reverse transcriptase quantitative PCR, pharmacology and CRISPR-Cas9-mediated knockdown of the Trpc5 gene. The addition of computer simulations for understanding the impact of E2 on the electrical activity of ARC kisspeptin cells is also a strength.

The authors add interesting information on the complement of ionic currents in ARC kisspeptin neurons and on their regulation by E2 to what was already known in the literature. Pharmacological and electrophysiological experiments appear of the highest standards and robust statistical analyses are provided throughout. The impact of E2 replacement on calcium and potassium currents is compelling. Likewise, the results of Trpc5 gene knockdown do provide good evidence that the TRPC5 channel plays a key role in mediating the NKB-mediated slow EPSP. Surprisingly, this also revealed an unsuspected role for this channel in regulating the membrane potential and excitability of ARC kisspeptin neurons.

Weaknesses:

The manuscript also has weaknesses that obscure some of the conclusions drawn by the authors.

One is that the authors compare here two conditions, OVX versus OVX replaced with high E2, that may not reflect the physiological conditions under which the proposed transition between neuropeptide-dependent sustained firing and less intense burst firing might take place (i.e. the diestrous [low E2] and proestrous [high E2] stages of the estrous cycle). This is an important caveat to keep in mind when interpreting the authors' findings. Indeed, that E2 alters certain ionic currents when added back to OVX females, does not mean that the magnitude of all of these ionic currents will vary during the estrous cycle.

In addition, although the computational modeling indicates a role of the various E2-modulated conductances in causing a transition in ARC kisspeptin neuron firing pattern, their role is not directly tested in physiological recordings, weakening the link between these changes and the shift in firing patterns.

Overall, the manuscript provides interesting information about the effects of E2 on specific ionic currents in ARC kisspeptin neurons and some insights into the functional impact of these changes. However, some of the conclusions of the work, with regard, in particular, to the role of these changes in ion channels and their implications for the LH surge, are not fully supported by the findings.

Author response:

The following is the authors’ response to the current reviews.

Reviewer #1 (Public review):

Summary:

In this work, Qiu and colleagues examined the effects of preovulatory (i.e., proestrous or late follicular phase) levels of circulating estradiol on multiple calcium and potassium channel conductances in arcuate nucleus kisspeptin neurons. Although these cells are strongly linked to a role as the "GnRH pulse generator," the goal here was to examine the physiological properties of these cells in a hormonal milieu mimicking late proestrus, the time of the preovulatory GnRH-LH surge. Computational modeling is used to manipulate multiple conductances simultaneously and support a role for certain calcium channels in facilitating a switch in firing mode from tonic to bursting. CRISPR knockdown of the TRPC5 channel reduced overall excitability, but this was only examined in cells from ovariectomized mice without estradiol treatment.

Comments to address most recent author response:

The concern regarding the CRISPR experiments being confined to OVX mice is that the results can only suggest that CRISPR-mediated knockdown of TRPC5 can, at best, phenocopy the OVX+E condition. A reciprocal experiment in the opposite direction (for example, that returning TRPC5 to OVX levels in OVX+E mice prevents the changes in firing activity and pattern typical of the OVX+E2 condition) would strengthen the indication that E2-sensitive changes in TRPC5 expression and function are critically important to surge function. Acknowledging this as a limitation of the studies would help to better contextualize the value of the CRISPR experiments to an understanding of surge mechanisms when done only in OVX conditions.

We have noted in the manuscript that “It would be of interest in future experiments to do the reciprocal experiment to see if overexpressing Trpc5 channels in Kiss1ARH neurons from OVX + E2 females restores the RMP and “rescues” the synchronization phenotype.”

The nature of the confusion regarding the consideration of OVX+E2 conditions in the computational model primarily arises from the methods description in the supplemental file: "The effect of E2 on ionic currents is modelled as a change in the maximum conductance parameter. For currents IM,IT, ICa and ITRPC5 this change is inferred from the qPCR data assuming that the conductance is directly proportional to the mRNA expression." If these were instead based on the whole-cell recordings as the authors now indicate in their response, then this description needs to be edited and clarified accordingly. Furthermore, the section states, "For ISK, IBK, Ileak, the OVX and OVX+E2 conductances are obtained from current-voltage relationships recorded from Kiss1ARH neurons in the absence/presence of iberiotoxin (BK blocker) and apamin (SK blocker). All other currents were assumed to be unaffected by E2." This section thus does not directly indicate that the recordings in the stated figures were used in the model, and moreover suggests that currents besides ISK, IBK, and Ileak were not different in OVX+E2 conditions.

The prior evidence stated for correlation of mRNA and channel conductance is not explicitly cited in the manuscript. It is well known that post-translational modifications, physiological modulation of individual channel biophysical properties, and many other factors can influence the end output of a membrane conductance. Therefore, the authors should, at minimum, provide a literature citation supporting the assumption used here.

We have re-written the paragraph on “Modelling the effects of E2” in the Supplemental Information (now Appendix 1) to clarify the that the modeling was based on a combination of electrophysiological recordings and the qPCR data presented in this and previous publications. The statement that “all other currents were assumed to be unaffected by E2” was a misstatement and has been deleted. As per the reviewer’s request, we have listed seven publications that document the correlation between the mRNA expression and channel conductance for the various channels. We thank the reviewer for the suggestion.

Reviewer #2 (Public review):

Summary:

Kisspeptin neurons of the arcuate nucleus (ARC) are thought to be responsible for the pulsatile GnRH secretory pattern and to mediate feedback regulation of GnRH secretion by estradiol (E2). Evidence in the literature, including the work of the authors, indicates that ARC kisspeptin coordinate their activity through reciprocal synaptic interactions and the release of glutamate and of neuropeptide neurokinin B (NKB), which they co-express. The authors show here that E2 regulates the expression of genes encoding different voltage-dependent calcium channels, calcium-dependent potassium channels and canonical transient receptor potential (TRPC5) channels and of the corresponding ionic currents in ARC kisspeptin neurons. Using computer simulations of the electrical activity of ARC kisspeptin neurons, the authors also provide evidence of what these changes translate into in terms of these cells' firing patterns. The experiments reveal that E2 upregulates various voltage-gated calcium currents as well as 2 subtypes of calcium-dependent potassium currents, while decreasing TRPC5 expression (an ion channel downstream of NKB receptor activation), the slow excitatory synaptic potentials (slow EPSP) elicited in ARC kisspeptin neurons by NKB release and expression of the G protein-associated inward-rectifying potassium channel (GIRK). Based on these results, and on those of computer simulations, the authors propose that E2 promotes a functional transition of ARC kisspeptin neurons from neuropeptide-mediated sustained firing that supports coordinated activity for pulsatile GnRH secretion to a less intense burst-like firing pattern that could favor glutamate release from ARC kisspeptin. The authors suggest that the latter might be important for the generation of the preovulatory surge in females.

Strengths:

The authors combined multiple approaches in vitro and in silico to gain insights into the impact of E2 on the electrical activity of ARC kisspeptin neurons. These include patch-clamp electrophysiology combined with selective optogenetic stimulation of ARC kisspeptin neurons, reverse transcriptase quantitative PCR, pharmacology and CRISPR-Cas9-mediated knockdown of the Trpc5 gene. The addition of computer simulations for understanding the impact of E2 on the electrical activity of ARC kisspeptin cells is also a strength.

The authors add interesting information on the complement of ionic currents in ARC kisspeptin neurons and on their regulation by E2 to what was already known in the literature. Pharmacological and electrophysiological experiments appear of the highest standards and robust statistical analyses are provided throughout. The impact of E2 replacement on calcium and potassium currents is compelling. Likewise, the results of Trpc5 gene knockdown do provide good evidence that the TRPC5 channel plays a key role in mediating the NKB-mediated slow EPSP. Surprisingly, this also revealed an unsuspected role for this channel in regulating the membrane potential and excitability of ARC kisspeptin neurons.

Weaknesses:

The manuscript also has weaknesses that obscure some of the conclusions drawn by the authors.

One is that the authors compare here two conditions, OVX versus OVX replaced with high E2, that may not reflect the physiological conditions under which the proposed transition between neuropeptide-dependent sustained firing and less intense burst firing might take place (i.e. the diestrous [low E2] and proestrous [high E2] stages of the estrous cycle). This is an important caveat to keep in mind when interpreting the authors' findings. Indeed, that E2 alters certain ionic currents when added back to OVX females, does not mean that the magnitude of all of these ionic currents will vary during the estrous cycle.

We do know that the slow EPSP, which is generated by TRPC5 channels, tracks beautifully with the steroid state of female mice. Using our E2 treatment paradigm that generates a LH surge in OVX females (left panel in Author response image 1), there is no difference in the amplitude of the slow EPSP in proestrous versus OVX + E2 females (right panel in Author response image 1).

In addition, although the computational modeling indicates a role of the various E2-modulated conductances in causing a transition in ARC kisspeptin neuron firing pattern, their role is not directly tested in physiological recordings, weakening the link between these changes and the shift in firing patterns.

In future experiments we will test directly the physiological contribution of the other E2-modulated conductances in causing the transition in the firing pattern of arcuate Kiss1 neurons using CRISPR/SaCas9 technology as we have documented for the TRPC5 channel (e.g., Figures 11 and 12).

Overall, the manuscript provides interesting information about the effects of E2 on specific ionic currents in ARC kisspeptin neurons and some insights into the functional impact of these changes. However, some of the conclusions of the work, with regard, in particular, to the role of these changes in ion channels and to their implications for the LH surge, are not fully supported by the findings.

---------

The following is the authors’ response to the previous reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

In this work, Qiu and colleagues examined the effects of preovulatory (i.e., proestrous or late follicular phase) levels of circulating estradiol on multiple calcium and potassium channel conductances in arcuate nucleus kisspeptin neurons. Although these cells are strongly linked to a role as the "GnRH pulse generator," the goal here was to examine the physiological properties of these cells in a hormonal milieu mimicking late proestrus, the time of the preovulatory GnRH-LH surge. Computational modeling is used to manipulate multiple conductances simultaneously and support a role for certain calcium channels in facilitating a switch in firing mode from tonic to bursting. CRISPR knockdown of the TRPC5 channel reduced overall excitability, but this was only examined in cells from ovariectomized mice without estradiol treatment. The manuscript has been substantially improved from the initial version by the addition of new experiments and clarification of important figures. Importantly, the overlap of data with previous reports from the same group has been corrected.

Strengths:

(1) Examination of multiple types of calcium and potassium currents, both through electrophysiology and molecular biology.

(2) Focus on arcuate kisspeptin neurons during the surge is relatively conceptually novel as the anteroventral periventricular nucleus (AVPV) kisspeptin neurons have received much more attention as the "surge generator" population.

(3) The modeling studies allow for direct examination of manipulation of single and multiple conductances, whereas the electrophysiology studies necessarily require examination of each current in isolation. Construction of an arcuate kisspeptin neuron model promises to be of value to the reproductive neuroendocrinology field.

Weaknesses:

A remaining weakness in this revised version of the manuscript is that the relevance of the CRISPR experiments is still rather tenuous given that the goal is to understand what happens in the estrogen-treatment condition, and these experiments were performed only in OVX mice. Similar concerns reflect that the computational model examining the effect of E2 infers multiple conductances based on qPCR data and an assumption that the conductances are directionally proportional to the level of gene expression, and then tunes these to the current recordings obtained from OVX mice, without a direct confirmation in OVX+E2 conditions that the model parameters accurately reflect the properties of these currents in the presence of estrogen.

We are still puzzled by Reviewer’s concerns about doing the CRISPRing of Trpc5 in the OVX+E2 females. The Trpc5 channel expression is significantly reduced with the E2 treatment (Figure 10E) which we know translates into a minimal slow EPSP (Figure 2, Qiu eLife 2016) and is essentially equivalent to the slow EPSP amplitude in the Trpc5 mutagenesis in the ovariectomized females (Figure 12). TRPC5 channel conductance is already at “rock bottom.” The modeling informs us that such a low TRPC5 conductance will not support a long lasting slow EPSP and sustained firing (Figure 13A).

Also, we respectively point out that we have published a score of papers over the past 20 years showing that the channel conductance does correlate with the mRNA expression (e.g., Qiu et al., eLife 2018). Secondly, the model does take into consideration the OVX + E2 conditions (Figure 13B,C) which is based on the extensive whole-cell recordings presented in Figures 4,5,6,7,8 and 9.

Reviewer #2 (Public Review):

Summary:

Kisspeptin neurons of the arcuate nucleus (ARC) are thought to be responsible for the pulsatile GnRH secretory pattern and to mediate feedback regulation of GnRH secretion by estradiol (E2). Evidence in the literature, including the work of the authors, indicates that ARC kisspeptin coordinate their activity through reciprocal synaptic interactions and the release of glutamate and of neuropeptide neurokinin B (NKB), which they co-express. The authors show here that E2 regulates the expression of genes encoding different voltage-dependent calcium channels, calcium-dependent potassium channels and canonical transient receptor potential (TRPC5) channels and of the corresponding ionic currents in ARC kisspeptin neurons. Using computer simulations of the electrical activity of ARC kisspeptin neurons, the authors also provide evidence of what these changes translate into in terms of these cells' firing patterns. The experiments reveal that E2 upregulates various voltage-gated calcium currents as well as 2 subtypes of calcium-dependent potassium currents while decreasing TRPC5 expression (an ion channel downstream of NKB receptor activation), the slow excitatory synaptic potentials (slow EPSP) elicited in ARC kisspeptin neurons by NKB release and expression of the G protein-associated inward-rectifying potassium channel (GIRK). Based on these results, and on those of computer simulations, the authors propose that E2 promotes a functional transition of ARC kisspeptin neurons from neuropeptide-mediated sustained firing that supports coordinated activity for pulsatile GnRH secretion to a less intense burst-like firing pattern that could favor glutamate release from ARC kisspeptin. The authors suggest that the latter might be important for the generation of the preovulatory surge in females.

Strengths:

The authors combined multiple approaches in vitro and in silico to gain insights into the impact of E2 on the electrical activity of ARC kisspeptin neurons. These include patch-clamp electrophysiology combined with selective optogenetic stimulation of ARC kisspeptin neurons, reverse transcriptase quantitative PCR, pharmacology and CRISPR-Cas9-mediated knockdown of the Trpc5 gene. The addition of computer simulations for understanding the impact of E2 on the electrical activity of ARC kisspeptin cells is also a strength.

The authors add interesting information on the complement of ionic currents in ARC kisspeptin neurons and on their regulation by E2 to what was already known in the literature. Pharmacological and electrophysiological experiments appear of the highest standards and robust statistical analyses are provided throughout. The impact of E2 replacement on calcium and potassium currents is compelling. Likewise, the results of Trpc5 gene knockdown do provide good evidence that the TRPC5 channel plays a key role in mediating the NKB-mediated slow EPSP. Surprisingly, this also revealed an unsuspected role for this channel in regulating the membrane potential and excitability of ARC kisspeptin neurons.

Weaknesses:

The manuscript also has weaknesses that obscure some of the conclusions drawn by the authors.

One is that the authors compare here two conditions, OVX versus OVX replaced with high E2, that may not reflect the physiological conditions under which the proposed transition between neuropeptide-dependent sustained firing and less intense burst firing might take place (i.e. the diestrous [low E2] and proestrous [high E2] stages of the estrous cycle). This is an important caveat to keep in mind when interpreting the authors' findings. Indeed, that E2 alters certain ionic currents when added back to OVX females, does not mean that the magnitude of all of these ionic currents will vary during the estrous cycle.

Unfortunately, mice are a poor reproductive model since female mice do not have a clear follicular (estradiol-driven) phase distinctive from the luteal (progesterone-driven) phase. Had we utilized a “proestrous” female, we could not with certainty distinguish between the effects of estradiol versus progesterone on the expression of the calcium and potassium channels that were the focus of this study. Therefore, using our physiological model we can state with confidence that “estradiol elicits distinct firing patterns in arcuate nucleus kisspeptin neurons….”

Overall, the manuscript provides interesting information about the effects of E2 on specific ionic currents in ARC kisspeptin neurons and some insights into the functional impact of these changes. However, some of the conclusions of the work, with regard, in particular, to the role of these changes in ion channels and their implications for the LH surge, are not fully supported by the findings.

As we pointed out in the Discussion, the O’Byrne lab has clearly shown the relevance of Kiss1ARH neuronal burst firing and the release of glutamate to its effects on the LH surge:

“Rather, we postulate that glutamate neurotransmission is more important for excitation of Kiss1AVPV/PeN neurons and facilitating the GnRH (LH) surge with high circulating levels of E2 when peptide neurotransmitters are at a nadir and glutamate levels are high in female Kiss1ARH neurons. Indeed, low frequency (5 Hz) optogenetic stimulation of Kiss1ARH neurons, which only releases glutamate in E2-treated, ovariectomized females (Qiu J. et al., 2016), generates a surge-like increase in LH release during periods of optical stimulation (Lin et al., 2021; Voliotis et al., 2021). In a subsequent study optical stimulation of Kiss1ARH neuron terminals in the AVPV at 20 Hz, a frequency commonly used for terminal stimulation in vivo, generated a similar surge of LH (Shen et al., 2022). Additionally, intra-AVPV infusion of glutamate antagonists, AP5+CNQX, completely blocked the LH surge induced by Kiss1ARH terminal photostimulation in the AVPV (Shen et al., 2022).”

Recommendations for the authors:

Reviewer #2 (Recommendations for The Authors):

The reviewer noted the following in the revised manuscript:

- page 6, the authors may consider adding that presynaptic effects of blocking calcium channels on the slow EPSP cannot be fully ruled out. Indeed, the added experiments do indicate that some of the effects can be explained by impaired regulation of TRPC5 channels by calcium influx through calcium channels; however, the senktide-induced current is not fully blocked by the broad-spectrum calcium channel inhibitor cadmium, suggesting that the effect of blocking these channels on the slow EPSP may involve other mechanisms, such as presynaptic effects.

Optogenetic stimulation of all Kiss1ARH neurons induces the release of NKB at “physiological” concentrations, which in turn generates a slow EPSP in the recorded Kiss1ARH neuron. Blocking voltage-gated calcium channels can inhibit the NKB release from presynaptic Kiss1ARH neurons, thereby reducing the amplitude of the slow EPSP. However, in whole-cell recordings of synaptically isolated Kiss1ARH neurons, senktide directly induces a large inward current (Figure 3F), which is generated by the opening of TRPC5 channels (Qiu et al. J. Neurosci 2021). Voltage-gated calcium channels are coupled to the activation of TRPC5 channels (Blair, Kaczmarek and Clapham, J. Gen Physiol 2009), so by blocking voltage-gated calcium channels, cadmium effectively abrogates the facilitating effects of these channels on TRPC5 channel activation and significantly reduces but does not abolish the inward (excitatory) current (Figures 3F-H). We have clarified in the Results (page 6) that the Kiss1ARH neurons were synaptically isolated as depicted in Figures 3F,G.

- page 8, bottom, the mean value given for the apamin-sensitive current amplitude in E2 treated females does not match that plotted on the I/V graph in Figure 7F.

Thank you for pointing out this typographical error, which we have corrected.

Version 3

Abstract

Abstract

Hypothalamic kisspeptin (Kiss1) neurons are vital for pubertal development and reproduction. Arcuate nucleus Kiss1 (Kiss1ARH) neurons are responsible for the pulsatile release of Gonadotropin-releasing Hormone (GnRH). In females, the behavior of Kiss1ARH neurons, expressing Kiss1, Neurokinin B (NKB), and Dynorphin (Dyn), varies throughout the ovarian cycle. Studies indicate that 17β-estradiol (E2) reduces peptide expression but increases Vglut2 mRNA and glutamate neurotransmission in these neurons, suggesting a shift from peptidergic to glutamatergic signaling. To investigate this shift, we combined transcriptomics, electrophysiology, and mathematical modeling. Our results demonstrate that E2 treatment upregulates the mRNA expression of voltage-activated calcium channels, elevating the whole-cell calcium current and that contribute to high-frequency burst firing. Additionally, E2 treatment decreased the mRNA levels of Canonical Transient Receptor Potential (TPRC) 5 and G protein-coupled K+ (GIRK) channels. When TRPC5 channels in Kiss1ARH neurons were deleted using CRISPR, the slow excitatory postsynaptic potential (sEPSP) was eliminated. Our data enabled us to formulate a biophysically realistic mathematical model of the Kiss1ARH neuron, suggesting that E2 modifies ionic conductances in Kiss1ARH neurons, enabling the transition from high frequency synchronous firing through NKB-driven activation of TRPC5 channels to a short bursting mode facilitating glutamate release. In a low E2 milieu, synchronous firing of Kiss1ARH neurons drives pulsatile release of GnRH, while the transition to burst firing with high, preovulatory levels of E2 would facilitate the GnRH surge through its glutamatergic synaptic connection to preoptic Kiss1 neurons.

Introduction

Hypothalamic kisspeptin (Kiss1) neurons and its cognate receptor (GPR 54 or Kiss1 R) are essential for pubertal development and reproduction, and may also be involved in the control of energy homeostasis (Kotani et al., 2001) (De Roux et al., 2003) (Seminara et al., 2003) (Messager et al., 2005) (Shahab et al., 2005) (d’Anglemont de Tassigny et al., 2008) (Qiu J. et al., 2018) (Rønnekleiv et al., 2022). Kisspeptin neurons within the arcuate nucleus of the hypothalamus (Kiss1ARH) co-express Kiss1, Neurokinin B (NKB) and Dynorphin (Dyn), which are all down-regulated by 17β-estradiol (E2) (Goodman et al., 2007; Navarro V. M. et al., 2009). However, Kiss1ARH neurons also express vesicular glutamate transporter 2 (vGlut2) and release glutamate, and both vGlut2 expression and glutamate release are upregulated by E2 in females (Qiu J. et al., 2018). This indicates that peptides and glutamate in Kiss1ARH neurons are differently modulated by E2 and suggests that there is a complex E2 regulation in these neurons such that they transition from predominantly peptidergic to glutamatergic neurotransmission and hence from a “pulsatile” to “surge” mode. It has been known for decades that neurons located within the arcuate nucleus are responsible for the pulsatile release of GnRH and subsequently pulsatile release of LH from the pituitary gland (O’Byrne et al., 1991) (Moenter et al., 1993). In this respect, it is now generally accepted that Kiss1ARH neurons are the main neurons responsible for the generation of pulsatile LH release, but the underlying cellular conductances generating this activity have not been elucidated.

Morphological studies have provided evidence that Kiss1ARH neurons can communicate directly with each other (Lehman et al., 2010) (Navarro V. M. et al., 2009) (Navarro V.M. et al., 2011). Furthermore, Kiss1ARH neurons express the NKB receptor, Tacr3, as well as the kappa (κ) opioid receptor (KOR), whereas GPR54 is not expressed in Kiss1ARH neurons, rendering them unresponsive to kisspeptin (d’Anglemont de Tassigny et al., 2008; Navarro V. M. et al., 2009; Wakabayashi et al., 2010). Using optogenetics and whole-cell recordings we demonstrated that high-frequency photoactivation of Kiss1ARH neurons induces an NKB-mediated slow excitatory postsynaptic potential (EPSP), which is mediated by the recruitment of canonical transient receptor potential (TRCP5) channels. The release of NKB is limited by co-released dynorphin, which acts presynaptically to inhibit further release. Together the two peptides cause synchronized firing of Kiss1ARH neurons (Kelly et al., 2018; Qiu J. et al., 2016; Qiu Jian et al., 2021), whereas kisspeptin and glutamate appear to be the main output signals from Kiss1ARH neurons (Qiu J. et al., 2016; 3XXX) (Voliotis et al., 2021) (Liu et al., 2021).

Although single action potential-generated calcium influx is sufficient to trigger the release of classical neurotransmitters such as glutamate, high frequency (10-20 Hz) is required for the release of neuropeptides such as kisspeptin, NKB and dynorphin (Qiu J. et al., 2016). Indeed, the slow EPSP, which underlies the synchronization, is similar to the “plateau potential” that has been described in hippocampal and cortical neurons (Arboit et al., 2020; Zhang Z. et al., 2011). Many neurons, including Kiss1ARH neurons, express the biophysical properties that allow them to continue to persistently fire even after a triggering synaptic event has subsided (Zylberberg and Strowbridge, 2017) (Qiu J. et al., 2016). Moreover, the intrinsic bi-stability of neurons that generates persistent firing activity has been linked to a calcium-activated, non-selective cation current (ICAN) (Zylberberg and Strowbridge, 2017), and TRPC channels, specifically TRPC5 channels, are thought to be responsible for the ICAN in cortical neurons (Zhang Z. et al., 2011). Therefore, we postulated that TacR3 activation via NKB drives influx of Ca+2 through TRPC5 channels leading to greater build-up of [Ca2+]i that facilitates the opening of more TRPC5 channels in a self-sustaining manner. Indeed, using the fast intracellular calcium chelator BAPTA, which has been shown to robustly inhibit TRPC5 channel activation in heterologous cells (Blair et al., 2009), we have been able to abolish the slow EPSP and persistent firing in Kiss1ARH neurons following optogenetic stimulation in female mice (Qiu Jian et al., 2021).

Although the expression of peptides in Kiss1ARH neurons are downregulated by high circulating levels (late follicular levels) of E2, the intrinsic excitability of and the glutamate release by Kiss1ARH neurons are increased by Cacna1g (Cav3.1, T-type calcium channel), Hcn1 and Hcn2 (Hyperpolarization-activated, Cyclic Nucleotide Gated channels) mRNA expression and Vglut2 mRNA expression, respectively (Qiu J. et al., 2018). Burst firing in CNS neurons, which efficiently releases fast amino acid transmitters like glutamate, is generated primarily by the T-type calcium channel current (IT) (e.g., in thalamic relay neurons), and the rhythmicity of this burst firing is dependent on the h-current (Ih) (for review, see (Lüthi and McCormick, 1998; Zagotta and Siegelbaum, 1996)). Ih is mediated by the HCN channel family, which includes channel subtypes 1-4, of which Hcn1 and Hcn2 are the main channels in Kiss1ARH neurons. Ih depolarizes neurons from hyperpolarized states, raising the membrane potential into the range of IT activation (Erickson et al., 1993a, 1993b; Kelly and Rønnekleiv, 1994; Lüthi and McCormick, 1998; Zhang C. et al., 2009). IT is mediated by the low-threshold voltage-gated calcium channels, CaV3.1-3.3 (for review see (Perez-Reyes, 2003)). IT initiates a transient Ca2+-driven depolarization above the threshold for action potential initiation (i.e., a low threshold spike) (Llinás, 1988; Tsien et al., 1987). This depolarization then drives neurons to fire an ensemble (burst) of Na+-driven action potentials.

Based on the above compelling evidence we postulated that Kiss1ARH neurons transition from peptidergic neurotransmission, driving the pulsatile release of GnRH via kisspeptin release into the median eminence, to glutamatergic transmission that facilitates in the preovulatory surge of GnRH (Lin et al., 2021) (Shen et al., 2022) through their projection to the anteroventral periventricular/periventricular nucleus Kiss1 (Kiss1AVPV/PeN) neurons (Qiu J. et al., 2016). Therefore, we initiated studies to thoroughly characterize effects of high circulating (late follicular) levels of E2 on the expression of the full complement of voltage-activated calcium channels (and currents) and the opposing K+ channels, involved in the repolarization, on the excitability of Kiss1ARH neurons. We performed whole-cell recordings and single cell RT-PCR analysis of Kiss1ARH neurons to determine which channels are involved in the physiological transition from peptidergic to glutamatergic neurotransmission. Our physiological findings were incorporated into a mathematical model that accounts for the E2 effects on the firing activity of Kiss1ARH neurons and validates our hypothesis that high levels of E2 facilitate the transition from peptidergic to glutamatergic neurotransmission.

Results

Voltage-activated Ca2+ channels contribute to burst firing of Kiss1ARH neurons

Our whole-cell current clamp recordings of Kiss1ARH neurons revealed that there is a significant transition of these Kiss1 neurons from a silent/tonic firing mode in the ovariectomized state to bursting/irregular firing with E2 treatment (Figure 1). Therefore, we sought to elucidate the cationic channels contributing to this physiological transition. Previously, we have shown that an increase in the intracellular calcium concentrations can potentiate TRPC5 channel current in POMC neurons (Qiu J. et al., 2010). Additionally, chelating intracellular calcium with BAPTA abolishes the slow excitatory postsynaptic potential (EPSP) and persistent firing in Kiss1ARH neurons (Qiu Jian et al., 2021). Hence, to investigate the contributions of voltage-activated calcium channels (VGCCs) to the increase in intracellular calcium, we measured the peak calcium current contributed by both the low and high voltage-activated calcium channels. The inward currents evoked by the voltage pulses (150-ms pulses starting from a holding potential of -80 mV with 10-mV increments) were identified as calcium (Ca2+) currents, as they were blocked by the universal VGCC inhibitor Cd2+ (200 μM) (McNally et al., 2020) (Figures 2A-E). The maximum total inward and Cd2+-sensitive currents reached their peak amplitudes at -10 mV. To differentiate between various calcium channel subtypes present in Kiss1ARH neurons, we applied selective antagonists individually, allowing us to isolate the drug-sensitive current for each cell. When we individually applied specific antagonists, we observed partial inhibition of whole-cell Ca2+ currents. Treatment with 10 μM nifedipine (Hiraizumi et al., 2008; Kato et al., 2003; Lee et al., 2002), an L-type Ca2+ channel inhibitor, resulted in a partial inhibition (26.1%) of the whole-cell calcium current. Additionally, application of 2 μM ω-conotoxin GVIA (conoGVIA, targeting N-type channels) (Lee et al., 2002), 200 nM ω-agatoxin IVA (AgaIVA, targeting P/Q-type channels) (Kato et al., 2003; McNally et al., 2020), 100 nM SNX-482 (targeting R-type channels) (Hiraizumi et al., 2008), or 1 μM TTA-P2 (TTAP2, targeting T-type channels) (McNally et al., 2020) also led to partial inhibition of the Ca2+ current (Figure 2F). The observed reduction in current with each inhibitor indicates the presence of all the major subtypes of Ca2+ currents in Kiss1ARH neurons. Among the inhibitors used, the largest components of the whole-cell calcium current were sensitive to nifedipine (26.1%), conoGVIA (25.1%), and SNX-482 (31.1%) (Figures 2A, B, D and 2F). Subsequently, we documented contributions from TTA-P2-sensitive channels, accounting for approximately 6.7% of the total Ca2+ current, and AgaIVA-sensitive channels, which constituted approximately 3.9% (Figures 2E, C). These findings indicate that high voltage-activated L-, N-, and R-type channels constitute the largest component of the voltage-activated Ca2+ current in Kiss1ARH neurons that not only increases the overall excitability but also greatly facilitate TRPC5 channel opening (Blair et al., 2009), which is the major downstream target of TACR3 activation by NKB (Qiu J. et al., 2016).

Properties of the firing of Kiss1ARH neurons from OVX and E2-treated OVX mice.

A–E. Representative whole cell, current clamp recordings of spontaneous phasic burst firing (A, which is only seen in E2-treated mice with clear up and down states), irregular burst firing (B), tonic firing (C), irregular firing (D) and silent (E) in Kiss1ARH neurons from E2-treated OVX mice. F-G. Summary pie chart illustrating the distribution of firing patterns in Kiss1ARH neurons from OVX+E2 (F, from 6 mice) or OVX (G, from 8 mice) mice (OVX versus OVX+E2, X2(3)= 6.05, p = 0.0011). H. Current clamp recording in a Kiss1ARH neuron from an OVX+E2 female demonstrating the response to glutamate (100 μM). RMP = -70 mV. Similar responses were observed in all recorded female Kiss1ARH neurons (n = 7 from 4 mice). I. The spiking activity above the bar in G was expanded to highlight the pronounced effects of glutamate on burst firing activity of Kiss1ARH neurons, characterized by an ensemble of spikes riding on top of low-threshold spikes (arrows). Drugs were rapidly perfused into the bath as a 4-μl bolus. J. Spontaneous AP frequency of Kiss1ARH neurons before, during, and after glutamate application (n = 7 cells). Data are presented as mean ± SEM. Statistical comparisons between different treatments were performed using one-way ANOVA (F(2, 18) = 14.60, p = 0.0002) followed by Bonferroni’s post hoc test. *** indicates p < 0.005. K-L. Representative traces showing the amplitude of slow EPSPs induced by high frequency photostimulation in Kiss1ARH neurons in the presence of ionotropic glutamate receptor antagonists CNQX and AP5 from OVX (K) and OVX+E2 (L) females. The arrows indicate the measurements of slow EPSP amplitude after low-pass filtering (shown in K). M. Bar graphs summarizing the slow EPSPs in the Kiss1ARH neurons from OVX and E2-treated OVX mice in the presence of CNQX and AP5. Statistical comparisons between the two groups were performed using an unpaired t-test (t(12) = 5.181, p =0.0002). Data are expressed as mean ± SEM, with data points representing the number of cells.

Relative contribution of voltage-gated calcium currents in Kiss1ARH neurons from OVX mice.

A-E, representative current –voltage relationships showing that Cd2+ (non-selective blocker of calcium channels) sensitive peak currents were inhibited by different calcium channel blockers: A, nifedipine; B, ω-conotoxin GIVA; C, ω-agatoxin IVA; D, SNX-482; E, TTA-P2. F, the maximum peak currents were measured at -10 mV. The proportions of Ca2+ currents inhibited by nifidipine (L type), ω-conotoxin GVIA (N type), ω-agatoxin IVA (P/Q), SNX-482 (R type) and TTA-P2 (T type). Data are expressed as mean ± SEM, with data points representing the number of cells.

Voltage-activated Ca2+ channels contribute to generation of slow-EPSP in Kiss1ARH neurons

Our previous study utilizing optogenetics demonstrated that high frequency photostimulation of Kiss1ARH neurons releases NKB. This release of NKB induces slow excitatory postsynaptic potentials (EPSPs) that facilitates the recruitment of other Kiss1ARH neurons, resulting in synchronous firing of the Kiss1ARH neuronal population (Qiu J. et al., 2016). Additionally, chelating intracellular calcium with the fast chelator BAPTA abolishes the slow EPSP and persistent firing in Kiss1ARH neurons, highlighting the role of calcium signaling in these processes (Qiu Jian et al., 2021). To access the involvement of HVA channels in the generation of the slow EPSP, we conducted experiments where we blocked the L-type Ca2+ channels with nifedipine (10 μM) and the N- and P/Q-type Ca2+ channels with ω-conotoxin MVIIC (1 μM) (Nunemaker et al., 2003). We then measured the slow EPSP. Indeed, both nifedipine and ω-conotoxin MVIIC significantly inhibited the slow EPSP by 35.2% and 60.4%, respectively (Figure 3A, B, E). In addition, we used SNX (100 nM), which selectively blocks R-type Ca2+ channels, and the slow EPSP was reduced to 28.7% of its control value (Figure 3C, E). The selective T-channel blocker TTA-P2 (5 µM) inhibited the slow EPSP by 28.6% (Figure 3D, E). Since the pharmacological blockade of the various calcium channels could also affect the presynaptic release of neurotransmitters (i.e., NKB), we measured the direct response of Kiss1ARH neurons to the TACR3 agonist senktide in the presence or absence of cadmium, a universal calcium channel blocker. In synaptically isolated whole-cell recordings of Kiss1ARH neurons, Cd2+ significantly inhibited the inward current induced by senktide (Figures 3F-H), which supports our posit that plasma membrane voltage-activated calcium channels contribute to the TRPC5 channel activation. Therefore, it appears that calcium channels contribute to maintaining the sustained depolarization underlying the slow EPSP.

Blockade of HVA Ca2+ channels decreases the slow EPSP in Kiss1ARH neurons.

A-D, representative traces showing that the slow EPSPs induced by high frequency photostimulation were abolished by perfusing the blocker of the L-type calcium channel, nifedipine (A) or N- and P/Q -type calcium channels, ω-conotoxin MVIIC (B), or the R-type calcium channel, SNX 482 (C), or the T-type calcium channel, TTA-P2 (D), respectively. The arrows indicate the measurements of slow EPSP amplitude, denoted as R1 and R2, after low-pass filtering (shown in C). E. Bar graphs summarizing the effects of drugs on the R2/R1 ratios. The slow EPSP was generated in OVX Kiss1-Cre::Ai32 mice. Comparisons between different treatments were performed using a one-way ANOVA analysis (F (3, 51) = 14.36, p<0.0001) with the Bonferroni’s post hoc test. *, **, **** indicates p < 0.05, 0.01, 0.001, respectively vs. control. F-G. Representative traces show that in voltage clamp, senktide induced a significant inward current in Kiss1ARH neurons (F) in the presence of TTX. This current was inhibited by the calcium channel blocker CdCl2 (200 μM)(G) in another cell. H. Bar graphs summarize the effect of the calcium channel blocker CdCl2 on the senktide-induced inward currents. Un-paired t-test for F versus G: t(20) = 2.575, p = 0.0181; * p < 0.05. Data are expressed as mean ± SEM, with data points representing the number of cells.

E2 increases the mRNA expression and the whole-cell current of voltage-activated Ca2+ channels

The neuropeptides NKB (tachykinin2, Tac2) and kisspeptin (Kiss1), which are expressed in Kiss1ARH neurons, are crucial for the pulsatile release of gonadotropin-releasing hormone (GnRH) and reproductive processes. E2 decreases the expression of Kiss1 and Tac2 mRNA in Kiss1ARH neurons but enhances the excitability of Kiss1ARH neurons by amplifying the expression of Cacna1g, Hcn1, and Hcn2 mRNA, as well as increasing T-type calcium currents and h-currents, respectively (Qiu, 2018, 20613}. Moreover, E2 drives Slc17a6 mRNA expression and enhances glutamatergic synaptic input to arcuate neurons and Kiss1AVPV/PeN neurons (Qiu J. et al., 2018). As a result, the E2-driven increase in Kiss1ARH neuronal excitability and glutamate neurotransmission may play a crucial role in triggering the surge of GnRH, ultimately leading to the LH surge (Lin et al., 2021) (Shen et al., 2022).

To assess the impact of E2 on the modulation of voltage-activated calcium channels and Kiss1ARH neuronal excitability, we employed real-time PCR (qPCR) to measure the relative expression levels of ion channel subtypes in Kiss1ARH neurons. We compared the expression in E2-treated females to those treated with oil, using the specific primers listed in Table 1. The quantification was conducted on pools of 10 neurons, as indicated in the Methods and Materials. In both oil- and E2-treated females, we quantified the expression of Cav1.2 (L-type), Cav2.1 (P/Q-type), Cav2.2 (N-type) and Cav2.3 (R-type) mRNAs. Remarkably, all of these mRNA transcripts exhibited increased expression levels in response to E2 treatment (Figure 4). Congruent with our previous findings (Qiu J. et al., 2018), we observed that the mRNA expression of Cav3.1 and Hcn1 was also upregulated in response to E2 treatment (Figure 4). These results suggested that E2 has a regulatory effect on the expression and function of all of these ion channels in Kiss1ARH neurons. To determine whether the increased mRNA expression translated into functional changes at the cellular level, we measured the whole-cell calcium current in Kiss1ARH neurons obtained from ovariectomized mice treated with either vehicle or E2. We discovered that E2 treatment led to a significant increase in the peak calcium current density in Kiss1ARH neurons (Figure 5A-D), which was recapitulated by our computational modeling (Figure 5E). These findings indicate that the upregulation of the mRNA expression of calcium channels by E2 translated to an augmented peak calcium current in Kiss1ARH neurons.

A, E2 increases the expression of low and high voltage-activated calcium channels in Kiss1ARH neurons.

Kiss1ARH neurons (three 10-cell pools) were harvested from each of 5 vehicle- and 5 E2-treated, OVX females to quantify ion channel mRNA expression of low and high voltage activated calcium channels as described in the Methods. The analysis included: T-type (Cav3.1) low voltage-activated, as well as the following high-voltage activated channels: R-type (Cav 2.3), L-type (Cav 1.2), N-type (Cav 2.2) and P/Q-type (Cav 2.1) calcium channels. Interestingly, all of these channels were upregulated with E2 treatment, which significantly increased the whole-cell calcium current (see Figure 5). Bar graphs represent the mean ± SEM with data points representing individual animals (oil versus E2, Unpaired t test for Cav 3.1, t(8)= 7.105, ***p=0.0001; for Cav 2.3, t(8) = 3.007, *p = 0.0169; for Cav 1.2, t(8)= 2.721, *p =0.0262; for Cav 2.2, t(8) = 4.001, **p 0.0039; for Cav 2.1, t(8) = 4.225, **p =0.0028).B, The same Kiss1ARH neuronal pools were also analyzed for mRNA expression of HCN1 ion channels, and E2 also increased the expression of hyperpolarization-activated, cyclic-nucleotide gated HCN1 channels in Kiss1ARH neurons. HCN1 channel mRNA expression was the most highly upregulated by E2 treatment in Kiss1ARH neurons. The expression values were calculated via the ΔΔCT method, normalized to GAPDH and relative to the oil control values. (oil versus E2, Unpaired t test, t(8)= 11.450, ****p < 0.0001).

E2 treatment (positive-feedback regimen) increases the Ca2+ currents in Kiss1ARH neurons.

A-B, Ca2+ currents in Kiss1ARH neurons with the same membrane capacitance from oil-treated (A) or E2-treated (B) animals. C, the maximum peak currents were measured at -10 mV. The current amplitudes were normalized to the cell capacitance in all cases to calculate current density. The bar graphs summarized the density of Ca2+ current in Kiss1ARH neurons from oil-treated and E2-treated animals. The mean density was significantly greater in E2-treated (13.4 ± 0.9 pA/pF, n = 11) than in oil-treated OVX females (7.2 ± 0.5 pA/pF, n = 40) (unpaired t-test, t(49) = 5.75, ****p < 0.0001). D. Relative contribution of voltage-gated calcium currents in Kiss1ARH neurons from OVX, E2-treated mice. The maximum peak currents were measured at -10 mV. The proportions of Ca2+ currents inhibited by nifidipine (L type), ω-Conotoxin GVIA (N type), ω-agatoxin IVA (P/Q), SNX-482 (R type) and TTA-P2 (T type). Data are expressed as mean ± SEM, with data points representing the number of cells. E. The modeling predicts that E2-treated, OVX females exhibit a significantly greater inward Ca2+ current (red trace) than the vehicle-treated females (black trace). The coductance of the modelled voltage-gated calcium current (L-, N-, P/Q-, and R-type) is set to 2.1 nS in the OVX state and 2.8 nS in OVX+E2 state, while for the T-type is set to 0.66 nS in the OVX state and 5 nS in OVX+E2 state.

The largest components of the calcium currents in Kiss1ARH neurons from the E2-treated, ovariectomized females were found to be sensitive to nifedipine, conoGVIA, and SNX-482, accounting for approximately 24.9%, 24.6%, and 27.0% of the total current across cells, respectively, which is very similar to their contributions to the whole-cell current from vehicle-treated, ovariectomized females (Figure 3). In addition, contributions from TTA-P2-sensitive channels accounted for approximately 11.1% of the total Ca2+ current, while agaIVA-sensitive channels contributed to approximately 11.0% (Figure 5D). These results highlight the prevalence of L-, N-, and R-type calcium channels as the major contributors to the whole-cell calcium current in Kiss1ARH neurons from E2-treated, ovariectomized females, but also the involvement of T-type and P/Q-type channels.

E2 does not alter the kinetics of calcium channel activation or de-inactivation

In order to determine whether the increase in peak voltage-activated calcium current density was the result of E2 regulating calcium channel kinetics, mRNA expression or both, we examined the voltage dependence of activation and inactivation. By measuring the voltage dependence of activation, we assessed how E2 affects the ability of calcium channels to open in response to membrane potential changes. Similarly, by examining the voltage dependence of inactivation, we determined how E2 influences the inactivation kinetics of calcium channels. Based on our results, there was no difference in the voltage dependence of activation between cells from the vehicle-treated control group (V1/2 = -32.3 ± 2.1 mV; n = 13) and cells from estrogen-treated females (V1/2 = -33.6 ± 2.5 mV; n = 11). Similarly, there was not a significant difference in the voltage dependence of inactivation between control cells (V1/2 = -48.9 ± 4.8 mV; n = 6) and estrogen-treated cells (V1/2 = -44.1 ± 1.9 mV; n = 5) (Figure 6). Therefore, although long-term E2 treatment increased the mRNA expression of HVA calcium channels (Figure 4), it did not affect the channel kinetics in Kiss1ARH neurons (Figure 6). Also, E2 downregulated the expression of Kcnd2 mRNA encoding Kv4.2, which is expressed in Kiss1ARH neurons (Mendonça et al., 2018) and has similar kinetics of activation as the T-type calcium channels (Oil-treated, ovariectomized females relative mRNA expression: 1.053, n = 5 animals versus E2-treated expression: 0.5643, n=5; t-test p = 0.0061). This opposing K+ current would dampen the inward calcium current. Therefore, it appears that long-term E2 treatment does not modulate calcium channel kinetics but rather alters the mRNA expression to increase the calcium channel conductance.

Voltage dependence of ICa in Kiss1ARH neurons from OVX and OVX+E2 mice.

A-B, top panels: activation and inactivation protocol. Bottom: representative traces. C-D, the mean V1/2 values for calcium channel activation were not significantly different for cells from controls versus cells from E2-treated females. Similarly, the V1/2 values for channel steady-state inactivation were similar for both groups. Data are expressed as mean ± SEM.

BK, SK and KCNQ channels are involved in modulating excitability of Kiss1ARH neurons

In the brain, calcium plays a crucial role in sculpting neuronal firing by activating potassium channels, which subsequently influence neuronal behavior (Nicoll, 1988; Storm, 1990). Since HVA and LVA calcium channels were expressed in Kiss1ARH neurons, all of which contribute to the elevation of intracellular calcium concentration ([Ca2+]i) that facilitates TRPC5 channel opening (Blair et al., 2009), our next step involved measuring the changes in Ca2+-activated K+ channel conductances and assessing their mRNA expression. In various cell types increases in cytosolic calcium levels, whether resulting from extracellular influx or intracellular release, lead to the activation of plasma membrane calcium-dependent potassium channels (Sah Pankaj and Louise Faber, 2002). Similarly, in Kiss1ARH neurons, these channels would be activated by calcium influx through all four types of high voltage-gated calcium channels, as well as the low voltage-activated calcium channel, which are all active during action potential firing. The activity of Ca2+-activated K+ channels play a crucial role in numerous physiological processes, including secretion and the regulation of neuronal firing properties. Two main families of Ca2+-activated K+ channel channels have been characterized, distinguished by their biophysical and pharmacological properties. These families are known as BK (Big Conductance K+) and SK (Small Conductance K+) channels in the CNS (Kshatri et al., 2018). BK channels are known for their high potassium selectivity and large single channel conductance, typically ranging from 100 to 300 pS. Activation of BK channels requires both calcium binding and membrane depolarization (Blatz and Magleby, 1987; Marty, 1989; Sah P., 1996; Storm, 1990). On the other hand, SK channels are simply activated by increases in cytosolic calcium levels, with their half-maximal activation at 0.3 μM (Bond et al., 1999).

To investigate K+ currents, the cells were maintained at a holding potential of -70 mV while being exposed to blockers CNQX, AP5, picrotoxin and TTX. Subsequently, the membrane potential was stepped by depolarizing voltages, ranging from -60 mV to +40 mV in 10 mV increments, for a duration of 500 ms (Brereton et al., 2013). This protocol was employed to activate K+ currents (Figure 7A). First, we examined SK currents. The mean current density was determined at the end of the voltage pulses. In vehicle-treated, OVX females the application of the SK channel blocker apamin (100 nM) (Spergel, 2007) led to a significant reduction in whole-cell currents in the +10 to +40 mV range (Figures 7A, B). The mean outward current density at +40 mV in the control group was 125.5 ± 7.2 pA/pF (n = 5) with the apamin-sensitive component contributing 60.2 ± 3.3 pA/pF (n = 5) (Figures 7A, C). In contrast, in the E2-treated females, the overall mean outward current density at +40 mV was 195.4 ± 14.6 pA/pF (n = 6), which was significantly greater than the vehicle control group (Figures 7D, E). However, there was no significant difference in the apamin-sensitive component between the vehicle-treated and E2-treated females, 60.2 ± 3.3 pA/pF (n = 5) versus 75.9 ± 12.3 pA/pF (n = 6), respectively (Figure 7F). Furthermore, to investigate the expression of the mRNAs encoding SK channel subunits in Kiss1ARH neurons from vehicle-treated and E2-treated OVX females, qPCR experiments were performed on 10-cell Kiss1ARH neuronal pools (Figure 7G). We focused on SK3 channels because these channels exhibit the highest expression in the hypothalamus and E2 regulates their expression (Bosch et al., 2002). E2 treatment had no effect on the mRNA expression of the SK3 subunit. These molecular findings support our electrophysiology results. Our computational model was calibrated so that SK channels contributed ∼50 pA/pF to the whole-cell outward K+ current in E2-treated females (Figure 7H).

Contribution of Small conductance, calcium-activated K+ (SK) channel to Kiss1ARH neuronal excitability.

A. Representative traces of the inhibition of outward currents before (left, control) and after the specific SK blocker Apamin (500 nM, middle). Apamin sensitive currents were calculated from the subtraction of control and apamin at depolarized potentials (right). Cells were clamped at -70 mV and given 500 ms voltage pulses from -60 mV to +40 mV in 10 mV steps at 0.2 Hz, as shown in A at the bottom. B. Mean current density-voltage relationships measured at the end of the 500 ms voltage step ranging from -60 mV to +40 mV were obtained in the absence and presence of apamin (two-way ANOVA: main effect of treatment (F(1, 8) = 22.69, p = 0.0014), main effect of time (F(10, 80) = 306.0, p < 0.0001) and interaction (F(10, 80) = 24.76, p < 0.0001); mean ± SEM; n = 5; post hoc Bonferroni test, ***p < 0.005, ****p < 0.001). C. Apamin sensitive current densities were obtained from B (mean ± SEM, n = 5). D. Representative traces of the inhibition of outward currents before (left, control) and after the specific SK blocker Apamin (500 nM, middle). Apamin sensitive currents were resulted from the subtraction of control and apamin at depolarized potentials (right). E. Mean current density-voltage relationships measured at the end of the 500 ms voltage step ranging from -60 mV to +40 mV were obtained in the absence and presence of apamin (two-way ANOVA: main effect of treatment (F(1, 10) = 12.85, p = 0.0050), main effect of time (F(10,100) = 264.1, p < 0.0001) and interaction (F(10, 100) = 11.93, p < 0.0001); mean ± SEM, n = 6; post hoc Bonferroni test, **p < 0.01, ****p < 0.001). F. Apamin sensitive current densities were obtained from C and E (ns; Two-way ANOVA followed by Bonferroni post hoc test; mean ± SEM; OVX, n = 5; OVX+E2, n = 6). G. Kiss1ARH neurons (three 10-cell pools) were harvested from each of 5 vehicle- and 5 E2-treated, OVX females to quantify the mRNA expression of SK3 ion channel. E2 did not increase the mRNA expression of small conductance calcium-activated K+ (SK3) channels in Kiss1ARH. Bar graphs represent the mean ± SEM, with data points representing the number of animals, oil versus E2, Unpaired t test, t(8)= 0.551, p = 0.5967. The expression values were calculated via the ΔΔCT method, normalized to GAPDH and relative to the oil control values. H. The mathematical model was calibrated on the electrophysiology data from Kiss1ARH neurons before and after treatment with the specific SK blocker apamin, left panel versus middle panel respectively. For the calibration it was assumed that the applied concentration of apamin (500nM) completely blocked the SK current. The modeled apamin-sensitive current with gSK=28.1 nS (right panel) matches the electrophysiological data from OVX animals.

Additionally, following the same protocol and in the presence of the same cocktail of blockers (CNQX, AP5, picrotoxin and TTX), we investigated the contribution of BK channels to Kiss1ARH neuronal excitability. In the OVX females, the application of the BK channel blocker iberiotoxin (ibTx, 200 nM) (Niday and Bean, 2021) resulted in only a slight attenuation of the outward current (n = 5) (Figures 8A, B). The ibTx-sensitive current density measured at +40 mV was 32.8 ± 12.0 pA/pF (Figures 8A, C). However, in the E2-treated females, the application of ibTx significantly attenuated the whole-cell K+ current from +20 to +40 mV, (Figures 8D, E). Additionally, the ibTx-sensitive current was significantly larger in the +0 to +40 mV range in the E2-treated females compared to the OVX females (Figure 8F). These findings indicate that E2 treatment modulates the activity of ibTx-sensitive BK current in Kiss1ARH neurons, resulting in increased current density (90.0 ± 10.1 pA/pF (n = 6) versus 31.1 ± 8.4 pA/pF (n = 5) at +40 mV). To investigate the expression of mRNA encoding BK channel subunits in Kiss1ARH neurons from vehicle-treated and E2-treated OVX females, qPCR experiments were performed on 10-cell Kiss1ARH neuronal pools (Figure 8G). E2 treatment significantly increased the mRNA expression of the BKα1 (Kcnma1) subunit. These findings support our electrophysiological findings that there is a significant increase in BK channel activity in Kiss1ARH neurons with E2 treatment (Figure 8F). In addition, E2 increased the mRNA expression of Kcnb1 encoding delayed rectifier Kv2.1 (E2-treated relative mRNA expression: 1.672, n = 5 versus oil-treated mRNA expression: 1.086, n = 5; t-test p-value = 0.0024). The up-regulation of the two of these K+ channels (BKα1 and Kv2.1) would facilitate rapid repolarization of Kiss1ARH following an action potential. Our computational model was calibrated so that BK channels contributed ∼100 pA/pF to the whole cell outward K+ current in the E2-treated females (Figures 8H).

E2 up-regulates Kcnma1 mRNA and BK current in Kiss1ARH neurons.

A. Representative traces of the inhibition of outward currents before (left, control) and after the specific BK blocker iberiotoxin (IbTx; 200 nM, middle). IbTx sensitive currents were calculated from the subtraction of control and IbTx at depolarized potentials (right). Cells were clamped at -70 mV and given 500 ms voltage pulses from -60 mV to +40 mV in 10 mV steps at 0.2 Hz, as shown in A at the bottom. B. Mean current density-voltage relationships measured at the end of the 500 ms voltage step ranging from -60 mV to +40 mV were obtained in the absence and presence of IbTx (two-way ANOVA: main effect of treatment (F(1, 8) = 0.8841, p = 0.3746), main effect of time (F(10, 80) = 71.56), p < 0.0001) and interaction (F(10, 80) = 1.127, p = 0.3528); mean ± SEM, n = 5; post hoc Bonferroni test, p > 0.05). C. IbTX sensitive current densities were obtained from B (mean ± SEM, n = 5). D. Representative traces of the inhibition of outward currents before (left, control) and after the specific BK blocker iberiotoxin (IbTx; 200 nM, middle). IbTx sensitive currents were resulted from the subtraction of control and IbTx at depolarized potentials (right). E. Mean current density-voltage relationships measured at the end of the 500 ms voltage step ranging from -60 mV to +40 mV were obtained in the absence and presence of IbTX (two-way ANOVA: main effect of treatment (F(1, 10) = 4.660, p = 0.0562), main effect of time (F(10, 100) = 63.98, p < 0.0001) and interaction (F(10, 100) = 4.907, p < 0.0001); mean ± SEM, n = 6; post hoc Bonferroni test, *p < 0.05, **p < 0.01, **** < 0.001). F. IbTx sensitive current densities were obtained from C and E (two-way ANOVA: main effect of treatment (F(1, 9) = 22.04, p = 0.0011), main effect of time (F(10, 90) = 78.26, p < 0.0001) and interaction (F(10, 90) = 17.84, p <0.0001); mean ± SEM, OVX, n = 5; OVX+E2, n = 6; Bonferroni post hoc test, *p < 0.05, ***p < 0.005, ****p < 0.001). G. Kiss1ARH neurons (three 10-cell pools) were harvested from each of 5 vehicle- and 5 E2-treated, OVX females to quantify the mRNA expression of BKα channel. E2-treatment increased the mRNA expression of BKα. The expression values were calculated via the ΔΔCT method, normalized to GAPDH and relative to the oil control values. Bar graphs represent the mean ± SEM, with data points representing the number of animals (unpaired two-tailed t-test for BK, t(8) = 3.479, **p = 0.0083). H. The mathematical model was calibrated to reproduce the current voltage relationship observed in Kiss1ARH neurons before and after treatment with IbTx. For the calibration it was assumed that the applied concentration of IbTx (200nM) completely blocked the BK current. The modeled IbTx -sensitive current with gBK=20.0 nS (right panel) matches the electrophysiological data from OVX+E2 animals.

Traditionally, the after hyperpolarization is divided into three distinct phases: fast (fAHP), medium (mAHP), and slow after hyperpolarization (sAHP) (Storm, 1990; Vogalis et al., 2003). The fast after hyperpolarization (fAHP) is primarily mediated by the BK family of potassium channels (Storm, 1987). The medium after hyperpolarization (mAHP) is predominantly mediated by apamin-sensitive SK2 channels (Bond et al., 2004; Peters et al., 2005). However, KCNQ family members contribute to both the mAHP and sAHP (Tzingounis A. V. et al., 2010; Tzingounis A. V. and Nicoll, 2008). Therefore, to investigate the contribution of KCNQ channels to Kiss1ARH neuronal excitability, voltage clamp experiments were conducted in the presence of TTX, CNQX, AP5, and picrotoxin, and a standard M-current protocol was run using the M-channel blocker XE-991 to isolate the M-current (Greene et al., 2017; Roepke et al., 2011) (Figure 9 A, B). The application of XE-991 resulted in the inhibition of M-current within a physiologically relevant voltage range of -60 to -30 mV in E2-treated OVX females but exhibited minimal impact in OVX females (Figure 9 C, D). Although the XE991-sensitive current was relatively small compared to other voltage-activated K+ conductances, it demonstrated a significant increase in E2-treated, OVX females (Figures 9 E). The maximum peak current density sensitive to XE-991 at -30 mV was found to be four times higher in E2-treated OVX females when compared to OVX females. This would contribute to the repolarization following burst firing. Furthermore, E2 increased the mRNA expression of Kcnq2, (Figure 9F), which suggests that KCNQ channels play a role in repolarizing Kiss1ARH neurons following burst firing. Indeed, our modeling predicted that M-current contributed to the repolarization following burst firing (Figure 9G).

E2 up-regulates KCNQ channels and M-current in Kiss1ARH neurons.

A-B. Representative current traces of the M-current inhibition caused by 40 µM XE-991 perfused for 10 min in (A) OVX-oil and (B) OVX+E2-treated female mice. Inset: M-current deactivation protocol. C-D. Current density-voltage plots from – 75 to –30 mV of vehicle and XE-991 perfusion in (C) OVX-oil and (D) OVX+E2-treated mice. Two-way ANOVA for C: main effect of treatment (F(1, 17) = 1.908, p = 0.1851), main effect of time (F(9, 153) = 187.1, p < 0.0001), and interaction (F(9, 153) = 3.901, p = 0.0002); Veh, n = 11; XE-991, n = 8; Bonferroni post hoc test, p > 0.05. For D: main effect of Veh and XE-991 (F(1, 24) = 24.92, p < 0.0001), main effect of time (F(9, 216) = 174.5, p < 0.0001), and interaction (F(9, 216) = 52.75, p < 0.0001); Veh, n = 13; XE-991, n = 13; Bonferroni post hoc test, a = p < 0.05, b = p < 0.001. E. Treatment with E2 elevated, while XE-991 diminished the maximum peak current density elicited by a -30 mV step in OVX- and OVX+E2-treated mice. Two-way ANOVA: main effect of Veh and XE-991 (F(1, 41) = 47.59, p < 0.0001), main effect of OVX and OVX+E2 (F(1, 41) = 15.76, p = 0.0003), and interaction (F(1, 41) = 18.2, p = 0.0001; Veh, n = 11; XE-991, n = 8; Bonferroni post hoc test, Veh: OVX vs. OVX+E2, a = p < 0.001. XE-991: OVX vs. OVX+E2, p > 0.05. Data are expressed as mean ± SEM, with data points representing the number of cells. F. Kiss1ARH neurons (three10-cell pools) were harvested from each of 5 vehicle- and 5 E2-treated, OVX females to quantify the mRNA expression of Kcnq2. E2 treatment increased the mRNA expression of Kcnq2. Unpaired t-test, t(8) = 4.850, **p = 0.0013. Data are expressed as mean ± SEM, with data points representing the number of animals. G. Percent contribution of the different K+ currents to the repolarization current during burst-type firing activity in the OVX+E2 state. At each time point, the length of each color bar denotes the percent contribution of the corresponding current to the total outward current.

E2 increases Vglut2 but down regulates Tac2, Trpc5 and Girk2 mRNA expression in Kiss1ARH neurons

Based on our electrophysiological results, Kiss1ARH neurons appear to transition from peptidergic to glutamatergic neurotransmission through E2-mediated changes in the expression of voltage-activated Ca2+ channels and K+ channels and their respective conductances. Therefore, we asked the question is there a difference in peptide and glutamate mRNA expression mediating this transition? Therefore, we ran a comparison between Tac 2 (NKB) and Vglut2 (surrogate for glutamate) expression. The cycle threshold (CT) was compared between Tac2 and Vglut2 as well as Kiss1, TRPC5 and GIRK2 in Kiss neuronal cell pools from OVX oil-treated and OVX E2-treated animals (Figures 10A, B). It is worth noting that lower number of cycles illustrate a higher quantity of mRNA expression because the fluorescence is detected earlier, and one cycle difference represents a doubling in expression. As expected, the reference gene Gapdh did not change with E2-treatment. However, quantitative PCR results revealed that E2 treatment of OVX females significantly reduced Tac2 expression (Figure 10C), whereas Vglut2 mRNA was significantly increased in Kiss1ARH neurons (Figure 10D). Moreover, both Trpc5 and Girk2 expression were significantly reduced in E2-treated, OVX females (Figures 10E, F).

Estradiol decreases Tac 2, Trpc5 and Kcnj6 but increases Vglut 2 mRNA expression in Kiss1ARH neurons.

A qPCR amplification curves illustrating the cycle threshold (CT) for Tac2, Gapdh, Kiss1, Trpc5, Slc17a6 (Vglut2) and Kcnj6 (GIRK2) in Kiss1ARH ten cell neuronal pools (three to six pools from each animal) in OVX Oil-treated, and B, in OVX E2-treated females. C. Quantitative real-time PCR analysis of Tac2 mRNA, D, Vglut2, E, Trpc5, F, Kcnj6. Comparisons were made between Oil-treated and E2-treated, OVX females using the comparative 2-ΔΔCT method. Bar graphs represent the mean ± SEM, with data points representing the number of animals. (Unpaired t test for Tac2, t(6) = 10.670, ***p < 0.0001; for Vglut2, t(7) = 9.678, ***p < 0.0001; for Trpc5, t(6) = 5.774, **p = 0.0012; Unpaired t test for Kcnj6, t(8) = 3.457, **p = 0.0086).

CRISPR mutagenesis of Trpc5 attenuates slow EPSP and reduces excitability of Kiss1ARH neurons

Our previous (Qiu Jian et al., 2021) and current findings suggested that TRPC5 channels play a dominant role in regulating cell excitability in ovariectomized females. Therefore, as proof of principle, we utilized a CRISPR approach to mutate TRPC5 channels in Kiss1ARH neurons similar to our previous studies (Hunker et al., 2020; Stincic et al., 2021). Hunker et al. developed a single viral vector for conditional expression of the smaller Staphylococcus aureus (SaCas9) and sgRNA that yields high-efficiency mutagenesis in specific cell types (Hunker et al., 2020). To selectively mutate Trpc5 in Kiss1ARH neurons, we generated two guide RNA’s, one targeting exon 2, which is conserved across all splice variants, and the other targeting exon 7, the pore forming domain (Figures 11A, B). A cohort of Kiss1ARH mice were given bilateral stereotaxic injections into the ARH of the two AAV1-FLEX-SaCas9-sgTrpc5’s or a control virus containing with AAV1-FLEX-SaCas9-U6-sgRosa26.

CRISPR mutagenesis of Trpc5 channels in Kiss1ARH neurons.

A, structure of AAV1-FLEX-SaCas9-U6sgTrpc5-exon2. Exon 2 of Trpc5 is denoted with guide sequence highlighted in red, the PAM is underlined. B. Structure of AAV1-FLEX-SaCas9-U6sgTrpc5-exon7. Exon 7 of Trpc5 is denoted with guide sequence highlighted in red, the PAM is underlined. C1, image of coronal section through the ARH from Kiss1-Cre::Ai32 mouse with dual co-injections of AAV-DIO-mCherry and AAV1-FLEX-SaCas9-U6-sgTrpc5. Scale = 200 µm. C2, C3, higher power overlays of epifluorescence (EYFP & mCherry) images with recording pipette patched onto a Kiss1ARH-Cre:mCherry cell (C3). Scale = 40 µm. D, quantitative PCR measurements of Trpc5 transcripts from 10 cell neuronal pools (three pools from each animal) in double sgRNA mutagenesis of Trpc5 (second sgRNA against pore forming region) in Kiss1ARH neurons. Primers were targeted to 1st (left panel) or 2nd (right panel) guide, respectively (Unpaired t test for 1st, t(10) = 10.67, ****p < 0.0001; for 2nd, t(10) = 10.79, ****p < 0.0001). Data are expressed as mean ± SEM, with data points representing the number of animals.

An additional Cre-dependent virus of the same serotype (AAV1) that drove expression of a fluorophore (YFP or mCherry) was co-administered in order to visualize injection quality and facilitate harvesting of cells (Figure 11C). After three weeks, mice underwent ovariectomy since OVX mice express the maximum slow EPSP amplitude (Qiu J. et al., 2016). Brain slices were prepared, and cells harvested as previously described (Qiu J. et al., 2018) and analyzed with qPCR. We found that the Trpc5 mutagenesis group displayed a reduction in relative expression of Trpc5 in Kiss1ARH neurons compared to the control group (Figure 11D). Hence, the qPCR data verified that in the sgTrpc5-targeted mice we can selectively reduce Trpc5 gene expression in targeted cells.

As predicted, mutagenesis of Trpc5 channels in Kiss1ARH neurons significantly attenuated the slow EPSP (Figures 12A, B, C) such that the postsynaptic excitation was reduced to a “trickle” of action potential firing. What we would not have predicted is that the double sgRNA mutagenesis of Trpc5 channels in Kiss1ARH neurons significantly hyperpolarized the resting membrane potential by 7 mV (Figure 12D). Moreover, the rheobase (minimum current required to induce firing) significantly increased by ∼13 % in females bearing the sgTrpc5 double mutagenesis (Figure 12E). The firing frequency versus injected current (F-I) curve for sgTrpc5 double mutagenesis Kiss1ARH neurons was also significantly attenuated (Figure 12F). In agreement with these experimental findings, current ramp simulations of our mathematical model confirmed that a decrease of the TPRC5 conductance increased the rheobase and reduce the firing activity of Kiss1ARH neurons (Figures 12G, H). In our model, this contribution of TRPC5 channels to cell excitability is attributed to intracellular calcium levels which activate TRPC5 channels independently of NKB stimulation. Finally, we employed our mathematical model to further investigate the transition from synchronous firing driven by NKB release and TRPC5 channel activation to burst firing generated by E2-mediated upregulation of endogenous conductances. As we varied the conductance of TRCP5 channels and GIRK channels under saturating extracellular signals (e.g., NKB and Dyn), we find that balance between the two conductances controls the neuronal excitability in the OVX parameterized model, with synchronous firing eliminated when TRPC5 conductance is low (red triangle) relative to the GIRK conductance (Figure 13A). Furthermore, the burst firing of the OVX+E2 parameterized model was supported by elevated h- and Ca 2+-currents (Figure 13B) as well as by the high conductance of Ca2+ channels relative to the conductance of TRPC5 channels (Figure 13C). Importantly, the simulated burst firing was analogous to the phasic burst firing recorded ex vivo in Kiss1ARH neurons (Figure 1A).

Double CRISPR mutagenesis of Trpc5 attenuates slow EPSP, increases rheobase and shifts the F-I curve.

A, high-frequency photo-stimulation (20 Hz) generated slow EPSP in Kiss1ARH neuron from ovariectomized, control mouse. Red trace is slow EPSP after low-pass filtering. B, slow EPSP in Kiss1ARH neuron from OVX, double sgTrpc5 -targeted (Trpc5 mut) mouse. C, summary of the effects of Trpc5 mutagenesis on slow EPSP amplitude in female mice. Unpaired t-test, t(27) = 5.916, ****p < 0.0001. Data are expressed as mean ± SEM, with data points representing the number of cells. D, double sgRNA mutagenesis of Trpc5 channels in Kiss1ARH neurons significantly increased the RMP (control: -64.5 ± 1.4 mV versus Trpc5 mut, 2: -71.1 ± 1.2 mV, Unpaired t-test, t(73) = 3.524, ***p = 0.0007). E, current ramp showing the increased rheobase in Kiss1ARH neuron from Trpc5 mut mice (control: 31.1 ± 1.2 pA, n = 31, versus Trpc5 mut, 35.3 ± 1.0 pA, n = 33, Unpaired t-test, t(62) = 2.777, **p = 0.0073). F, firing frequency vs. current (F-I) curves for control versus Trpc5 mut (two-way ANOVA: main effect of treatment, F(1,52) = 13.03, p = 0.0007, main effect of time, F(8,416) = 291.3, p < 0.0001, and interaction, F(8,416) = 6.254, p < 0.0001; Control, n = 26, Trpc5 mut, n = 29; post hoc Bonferroni test, *<0.05; **<0.01 and ***p < 0.005, respectively). G, model simulations of the effects of a current ramp (50 pA/sec) for OVX (left panel) and OVX female with reduced (muted) TRPC5 conductance (right panel), and H, the associated firing frequency vs. current curves. In the latter case the TRCP5 conductance was halved, which is a conservative estimation of the CRISPR state in which the Trpc5 is much more mutated in Kiss1ARH neurons (Figure 11).

Computational modeling of a Kiss1ARH neuron in the OVX and OVX + E2 state demonstrates its distinct dynamic responses.

A model of the Kiss1ARH neuron was developed and calibrated using molecular data and electrophysiological recordings of Kiss1ARH neurons from OVX and OVX+E2 mice. A. Simulations of the OVX-parameterized model demonstrating high frequency activity in response to saturating levels of NKB stimulation. The balance between GIRK and TRCP5 conductance controls the response of the neuron to NKB stimulation, with neuronal response eliminated when TRPC5 conductance is low (red triangle) relative to the GIRK conductance. B. The OVX+E2 parameterized models demonstrate sustained burst firing activity. The bursting activity that is supported by elevated h- and Ca 2+-currents (red square) as observed in OVX+E2 mice. C. In the OVX+E2 state, burst firing activity is also supported by high conductance of HVA Ca2+ channels relative to the conductance of TRPC5 channels. Representative points in the parameter space giving rise to burst firing activity are marked with red squares, whereas red triangles are used for points resulting in regular spiking. The black line separates these two regions of activity.

Discussion

E2 appears to play a critical role in transitioning the glutamatergic/peptidergic Kiss1ARH neurons from a high frequency firing mode during synchronization, which is dependent on NKB-driven activation of TRPC5 channels, to a short bursting mode that would facilitate glutamate release. E2 decreased the expression of the peptide neurotransmitters NKB (kisspeptin and dynorphin) and TRPC5 channels but increased the mRNA expression of Vglut2, voltage-activated calcium channels and Hcn channels that contribute to burst firing and glutamate release from the Kiss1ARH neurons. We determined that the increase in mRNA expression of the HVA calcium channels translated into a significant increase in whole-cell current with all of the calcium channels contributing proportionally. Most importantly the kinetics of activation and inactivation were unaltered with E2 treatment, which indicates that other post-translation modifications were not affecting channel activity. Surprisingly and somewhat counter intuitive, the BK α1 subunit was also upregulated, but based on our modeling the rapid repolarization of the Kiss1ARH neurons (i.e., the fast AHP) would facilitate higher frequency of action potential firing. Moreover, our modeling confirmed that TRPC5 channels, which generate the slow EPSP (a.k.a., plateau potential in other CNS neurons), are vital for initiating and sustaining synchronous firing of Kiss1ARH neurons, while subsequent activation of GIRK channels repolarizes Kiss1ARH neurons. E2 treatment of ovariectomized females decreased both Trpc5 and Girk2 channel mRNA expression, which in our model correlated with the reduction in sustained high frequency firing of Kiss1ARH neurons. Therefore, the synchronous high frequency firing of Kiss1ARH neurons in a low E2 milieu correlates with the pulsatile release of GnRH (LH from the pituitary gland), whereas the transition to burst firing in the presence of high circulating levels of E2 (e.g., proestrus) would facilitate the GnRH (LH) surge through its glutamatergic synaptic connection with Kiss1AVPV/PeN neurons (Lin et al., 2021; Qiu J. et al., 2016).

Core calcium conductances underlying synchronous and burst firing of Kiss1ARH neurons

TRPC5 channels are highly expressed in Kiss1ARH neurons (Figure 11), and TRPC5 channels are essentially ligand-activated calcium channels with a high permeability to calcium (PCa/PNa = 9:1) (Venkatachalam and Montell, 2007). In general, mammalian TRPC channels are activated by both G protein-coupled receptors and receptor tyrosine kinases (Ambudkar and Ong, 2007; Clapham, 2003), and are one of the major downstream effectors activated by glutamate binding to group I metabotropic glutamate receptors (mGluR1 and mGluR5) in CNS neurons (Bengtson et al., 2004; Berg et al., 2007; Faber et al., 2006; Tozzi et al., 2003). In substantia nigra dopamine neurons mGluR1 agonists induce a current that exhibits the tell-tale double-rectifying current-voltage plot of TRPC channel activation (Tozzi et al., 2003), similar to what we see with the effects of the NKB agonist senktide in Kiss1ARH neurons (Qiu Jian et al., 2021). Both mGluR1 and TacR3 are Gq-coupled to phospholipase C (PLC) activation which leads to hydrolysis of phosphatidylinositol 4,5-bisphosphate (PIP2) to diacylglycerol (DAG) and inositol 1,4,5 triphosphate (IP3), which is involved in channel activation (Birnbaumer, 2009). TacR3 (NKB) signaling has additional consequences since many K+ (e.g., GIRK, KCNQ) channels are dependent on PIP2 for channel opening, and depletion of PIP2 by PLC leads to channel closure (Brown and Passmore, 2009) (Zhang C. et al., 2013) (Whorton and MacKinnon, 2011) (Zheng et al., 2022). Therefore, depletion of PIP2 by NKB signaling would further facilitate the sustained firing of Kiss1ARH neurons during synchronization.

A plateau potential has been characterized in hippocampal and cortical neurons (Arboit et al., 2020; Zhang Z. et al., 2011) as such these neurons express biophysical properties that allow them to continue to persistently fire even after a triggering synaptic event has subsided (Zylberberg and Strowbridge, 2017). The persistent firing activity of these neurons is linked to ICAN (Zylberberg and Strowbridge, 2017), and TRPC5 channels appear to be responsible for the ICAN (Zhang Z. et al., 2011). With TacR3 activation in Kiss1ARH neurons there is an influx of Ca+2 through TRPC5 channels leading to greater build-up of [Ca2+]i that facilitates the opening of more TRPC5 channels in a self-sustaining (autocatalytic) manner (Qiu J. et al., 2016). Using the fast intracellular calcium chelator BAPTA, which has been shown to robustly inhibit TRPC5 channel activation in heterologous cells (Blair et al., 2009), we abolished the slow EPSP and persistent firing of Kiss1ARH neurons following optogenetic stimulation (Qiu Jian et al., 2021). Moreover, HVA calcium channel blockers attenuated the generation of the slow EPSP (Figure 3) so it appears that they also contribute to the ICAN since calcium influx via both LVA and HVA calcium channels can also facilitate TRPC5 channel opening in Kiss1ARH neurons. In the ovariectomized female, treatment with E2 upregulated Cav1.2, Cav2.1, Cav2.2, and Cav2.3 mRNA by 1.5-to 2-fold and Cav3.1 mRNA expression by ∼3-fold. Hence, E2 significantly increased whole-cell calcium currents Kiss1ARH neurons, which greatly enhanced the excitability and contributed to the burst firing of Kiss1ARH neurons (Figure 1). However, the amplitude of the slow EPSP with E2 treatment is only ∼25% of the amplitude in the ovariectomized state (Qiu J. et al., 2016). Therefore, there appears to be a physiologic transition of Kiss1ARH neurons from the slow EPSP firing mode in the OVX state to the burst firing mode in the presence of E2, which has important physiologic ramifications as discussed below.

TRPC5 and GIRK channels are vital for synchronization of Kiss1ARH neurons

Recently, Tian et al. demonstrated that TRPC4, a close homolog of TRPC5 sharing ∼64 percent homology, is a “coincidence detector” of neurotransmission by both Gq/11 and Gi/o-coupled receptors in lateral septal (LS) neurons (Tian et al., 2022). In whole-cell recordings of LS neurons, TRPC 4 channels mediate a strong depolarizing plateau potential that in contrast to TRPC5 channel activation in Kiss1ARH neurons, abrogates action potential firing as a result of a depolarization block. In many instances the plateau potential in LS neurons is followed by an AHP, which is dependent on the activation of Gi/o-coupled receptors. In contrast, we have not observed an AHP in Kiss1ARH neurons following the slow EPSP (Qiu J. et al., 2016). Tian and colleagues showed that the depolarizing plateau in LS neurons is codependent on activation of both Gq/11-coupled mGluR1 glutamate receptors and Gi/o-coupled γ-aminobutyric acid type B receptors, the latter activating GIRK channels. Moreover, the firing patterns in LS neurons encodes information about the relative strengths of these contrasting inputs (i.e., Gq/11 versus Gi/o) such that only mGluR1 produces weak depolarization accompanied by increased firing of LS neurons, whereas pure GABAB receptor activation hyperpolarizes the cells and abrogates firing activity. Coincident input of both mGluR1 and GABAB receptors results in a brief burst of action potentials followed by a pause in firing, and both the pause duration and firing recovery patterns reflect the relative strengths of Gq/11 versus Gi/o inputs. Importantly, Tian and colleagues computationally simulated these various scenarios with computational modeling, and similar to our modeling, utilized only TRPC4 and GIRK channels. A notable difference between the Kiss1ARH neurons and the LS neurons is that dynorphin binds to kappa-opioid receptors to open GIRK channels in the nerve terminal of Kiss1ARH neurons (presynaptic effect) (Qiu J. et al., 2016), whereas GABA binds to some GABAB receptors to hyperpolarize LS neurons (Tian et al., 2022). Therefore, although the high frequency firing activity of both LS and Kiss1ARH neurons can be modeled around TRPC and GIRK channels, the generated firing patterns are dramatically different based on the timing (co-incident activation of TRPC4 and GIRK channels in LS) and localization of the kappa opioid-coupled GIRK channels in the axon terminal of Kiss1ARH neurons. Moreover, E2-treated, ovariectomized females show a significant down-regulation of both Trpc5 and Girk2 mRNA expression in Kiss1ARH neurons (Figure 10), which is important for the physiological transitioning as described below.

Interestingly, the hypothalamic A12 (ARH) dopamine neurons show a rhythmic “oscillatory” firing behavior that transitions to a tonic firing mode with synaptic input from Thyrotropin-releasing hormone (TRH) neurons (Lyons D.J. et al., 2010) or feedback by circulating prolactin, released by pituitary lactotrophs (Lyons D. J. et al., 2012). A more recent paper has revealed that TRPC5 channels mediate the plateau potential and tonic firing in A12 dopamine neurons in response to prolactin (Blum et al., 2019). Similar to our findings with CRISPR deletion of Trpc5 in Kiss1ARH neurons (Figures 11 and 12), conditional knockout of Trpc5 in dopamine neurons abrogated the prolactin-induced plateau potential and tonic firing. Although Blum and colleagues did not model the oscillatory firing or tonic firing of the A12 dopamine neurons, their findings are consistent with our results showing that the activation of TRPC5 channels underlies the slow EPSP (plateau potential) and sustained firing.

Contribution of endogenous K+ channels to synchronized and burst firing

Beyond the ligand-gated (e.g., baclofen) GIRK channels, there are endogenous K+ channels that help sculpt the firing activity of kisspeptin neurons. We focused on the calcium-activated K+ channel family: the large-conductance, calcium-activated potassium (BK, also called BKCa, KCa1.1, MaxiK, Slo), small conductance, calcium-activated K+ (SK1, SK2, SK3) (Bond et al., 2005), and the K+ channels underlying the M-current (KCNQ, Kv7.1-7.5) (Brown and Passmore, 2009), which mediate the fast afterhyperpolarization (AHP), the intermediate AHP/slow AHP, respectively (Andrade et al., 2012) BK channels are gated by both voltage and cytoplasmic calcium and sculpt action potential firing in CNS neurons (Blatz and Magleby, 1987; Marty, 1989; Sah P., 1996; Storm, 1990). Indeed, BK channels have been shown to mediate rapid spike repolarization—i.e., the fast AHP in hippocampal CA1 pyramidal neurons (Lancaster and Nicoll, 1987; Storm, 1987). Blockade of BK channels in CA1 neurons attenuates the initial discharge frequency in response to current injection, which is attributable to suppression of the BK channel-dependent rapid spike repolarization (Lancaster and Nicoll, 1987; Storm, 1987). Blockade of BK channels is thought to increase inactivation of the spike-generating transient Na+ current and activate more of the slower K+ currents, thereby enhancing refractoriness and reducing excitability during the immediate aftermath of the first action potential (Shao et al., 1999). Thus, BK channels facilitate high-frequency burst firing of CA1 neurons. Furthermore, extracellular field recordings confirmed that BK channels contribute to high-frequency burst firing in response to excitatory synaptic input to distal dendrites in CA1 neurons (Gu et al., 2007). Therefore, BK channels appear to play an important role for early high-frequency, rapidly adapting firing in hippocampal CA1 pyramidal neurons, thus promoting the type of bursting that is characteristic of these cells in vivo during behavior. Based on our in vitro electrophysiological recordings and computational modeling we see a similar physiological phenomenon in Kiss1ARH neurons (Figure 8). Not only does E2 increase the mRNA expression of Kcnma1 (Figure 8G), but also the maximum BK (IbTx-sensitive) current by 3-fold, which would facilitate rapid repolarization during burst firing and promote glutamate release similar to hippocampal CA1 neurons.

In contrast to BK channel expression, E2 did not affect the mRNA expression of SK3 channel mRNA. SK channels underlie the apamin-sensitive component of the medium duration AHP and are responsible for repolarization following a burst of action potentials (Andrade et al., 2012; Bond et al., 2005). The activation of SK channels is voltage-independent, but SK channels have a higher affinity for Ca2+ than BK channels (Andrade et al., 2012). SK channels are tightly coupled (within 100 nm) to L-type Ca2+ channels, and BK channels (within 30 nm) are tightly coupled to N-type Ca2+ channels in hippocampal CA1 pyramidal neurons (Marrion and Tavalin, 1998). The determination of the proximity of SK and BK channels to HVA calcium channels in kisspeptin neurons will require cell-attached patch recordings. However, in Kiss1ARH neurons the SK channels may come into play during a short burst of action potentials but would become overwhelmed with the higher frequency synchronized, sustained firing as a result of NKB stimulation (Qiu J. et al., 2016). As discussed above what limits the synchronized firing of Kiss1ARH neurons is the activation of GIRK channels.

Finally, the calcium-activated slow AHP probably plays a critical role in the repolarization of Kiss1ARH after burst firing. The molecular identification of the channels mediating the slow AHP has long been an area of intense investigation (Andrade et al., 2012; Vogalis et al., 2003). A critical feature of the slow AHP is that it activates very slowly (hundreds of milliseconds) long after the rise in cytoplasmic Ca2+ (Sah P. and Clements, 1999) so an intermediate Ca2+ signaling molecule has long been thought to be involved. Indeed, Tzingounis and colleagues (Tzingounis A.V. et al., 2007) have provided compelling evidence that the diffusible calcium sensor hippocalcin is the critical intermediate molecule involved in Ca2+ sensing. The slow AHP is abrogated in hippocalcin KO mice, and transfection of hippocalcin into cultured hippocampal neurons generates a pronounced slow AHP in response to a depolarizing stimulus (Tzingounis A.V. et al., 2007). Importantly, the slow AHP is activated by Ca2+ with an EC50 ≈ 300 nM, which is well within the operational range of hippocalcin but well below that of calmodulin (Andrade et al., 2012). Finally, two seminal papers from Tzingounis and colleagues demonstrate that KCNQ 2, 3 channels are responsible for the slow AHP in hippocampal dentate neurons (Tzingounis A. V. and Nicoll, 2008), and KCNQ 5 channels are responsible for the slow AHP in CA3 neurons (Tzingounis A. V. et al., 2010). Moreover, the KCNQ channel blocker XE991 attenuates the slow AHP in CA3 neurons (Tzingounis A. V. and Nicoll, 2008). Based on these seminal findings we investigated the role of the KCNQ channels, which “classically” underlie the M-current in Kiss1ARH. The M-current was first identified in Kiss1ARH neurons by Conde and Roepke (Conde and Roepke, 2019), and presently we found that Kcnq2 mRNA is expressed in Kiss1ARH neurons and up-regulated by E2, which translated into a greater M-current in Kiss1ARH neurons (Figure 9). Incorporating the M-current into our computational model indeed supports our hypothesis that M-current, along with ISK and IBK, contributes to membrane repolarization after burst firing (Figure 9G). Importantly the slow AHP, as opposed to the fast AHP (BK) and medium AHP (SK) is highly regulated by multiple neurotransmitters (Andrade et al., 2012), which sets the stage for further modulation of the slow EPSP in Kiss1ARH neurons.

Importance of E2-driven physiological transitioning

Since the expression of the peptide neurotransmitters in Kiss1ARH neurons are down-regulated by E2, the Kiss1ARH neurons are believed to be under “inhibitory” control by E2 and are important for “negative-feedback” regulation of GnRH release (Lehman et al., 2013; Navarro V. M. et al., 2009; Smith et al., 2005) (Rance Naomi E. and Young, 1991) (Rance N.E., 2009). However, our past (Gottsch et al., 2011) and current findings document that these Kiss1ARH neurons express HVA and LVA calcium and HCN (pacemaker) channels and are excited by co-released glutamate from neighboring Kiss1ARH neurons, which indicates that these neurons have pacemaker electrophysiological properties similar to other CNS neurons (Bal and McCormick, 1993; Lüthi and McCormick, 1998). Additionally, in contrast to the neuropeptides, E2 increases Slc17a6 (Vglut2) mRNA expression, in addition to mRNA for the voltage-activated calcium and HCN channels, and increases glutamate release onto Kiss1AVPV/PeN neurons (Qiu J. et al., 2018). Interestingly, Slc17a6 mRNA expression in Kiss1ARH neurons and the probability of glutamate release are decreased along with the neuropeptides in intact versus castrated males (Nestor et al., 2016), which indicates a profound sex difference in the glutamate signaling by Kiss1ARH neurons (Nestor et al., 2016; Qiu J. et al., 2018). Obviously, in the male there is no preovulatory LH surge so there is no need for excitatory glutamatergic input to the few Kiss1AVPV neurons in the male.

In females, conditional knockout of Slc17a6 in Kiss1 neurons abrogates glutamate release from Kiss1ARH neurons (Qiu J. et al., 2018). Kiss1AVPV/PeN neurons do not express Slc17a6 and do not release glutamate. Within the Kiss1ARH neurocircuitry the lack of glutamate transmission does not diminish the slow EPSP in ovariectomized females (Qiu J. et al., 2018). Indeed, a recent publication from the Herbison lab demonstrates that glutamate generates small “synchronizing” events that are dependent on the ionotropic receptors (Han et al., 2023), but the fast neurotransmitter is unable to support the sustained firing (i.e., slow EPSP) that is necessary for peptide release and synchronization of the KNDy network. Rather, we postulate that glutamate neurotransmission is more important for excitation of Kiss1AVPV/PeN neurons and facilitating the GnRH (LH) surge with high circulating levels of E2 when peptide neurotransmitters are at a nadir and glutamate levels are high in female Kiss1ARH neurons. Indeed, low frequency (5 Hz) optogenetic stimulation of Kiss1ARH neurons, which only releases glutamate in E2-treated, ovariectomized females (Qiu J. et al., 2016), generates a surge-like increase in LH release during periods of optical stimulation (Lin et al., 2021; Voliotis et al., 2021). In a subsequent study optical stimulation of Kiss1ARH neuron terminals in the AVPV at 20 Hz, a frequency commonly used for terminal stimulation in vivo, generated a similar surge of LH (Shen et al., 2022). Additionally, intra-AVPV infusion of glutamate antagonists, AP5+CNQX, completely blocked the LH surge induced by Kiss1ARH terminal photostimulation in the AVPV (Shen et al., 2022). Therefore, there appears to be a clear role for glutamatergic transmission from the Kiss1ARH to Kiss1AVPV/PeN neurons in amplifying the LH surge in the female mouse. Finally it is important to keep in mind that even in the presence of high physiological levels of E2, the mRNA expression of Tac2 is many-fold higher than Kiss1 (Figure 10), which is essential for NKB maintaining synchronous firing of Kiss1ARH neurons, albeit at a lower frequency, across all physiological states (Qiu J. et al., 2016). Indeed, there is a progressive change from a strictly pulsatile pattern of GnRH in the hypophyseal portal circulation to one containing both pulsatile and non-pulsatile components during the development of the GnRH surge in the ewe (Evans, Dahl, Mauger, & Karsch, 1995; Evans, Dahl, Mauger, Padmanabhan, et al., 1995), and a pulsatile mode of LH secretion during the preovulatory LH surge is also evident in other species including humans (Rossmanith et al., 1990). Therefore, we believe that our cellular molecular and electrophysiological findings in combination with our computational modelling provide a foundation for understanding the complex role of Kiss1ARH neurons in controlling fertility in the mammal. Finally, our model provides the first comprehensive biophysical description of the conductances underlying the neuronal activity of KissARH neurons, which can serve as a basis for future computational modelling of the KissARH neuronal network and its interactions with other brain regions involved in the complex regulation of mammalian female reproduction.

Methods and Materials

Animals

All the animal procedures described in this study were performed in accordance with institutional guidelines based on National Institutes of Health standards and approved by the Institutional Animal Care and Use Committee at Oregon Health and Science University (OHSU).

Mice

Kiss1Cre transgenic female mice version 2 (Padilla et al., 2018) were selectively bred at OHSU. They also were crossed with heterozygous Ai32 mice (RRID:IMSR_JAX:024109, C57BL/6 background) which carry ChR2 (H134R)–EYFP gene in their Gt(ROSA)26Sor locus (Madisen et al., 2012). All animals were maintained under controlled temperature and photoperiod (lights on at 0600h and off at 1800h) and given free access to food (Lab Diets 5L0D) and water. Where specified, Kiss1Cre mice received viral injections to express channelrhodopsin 2 (ChR2) in Kiss1ARH neurons, fourteen to twenty-one days prior to each experiment as described (Qiu J. et al., 2016). Some of the females were ovariectomized seven days prior to an experiment.

Each animal was injected on day 5 following OVX with 0.25 μg E2 or vehicle, followed on day 6 with 1.50 μg E2 or vehicle and used for experiments on day 7 (Bosch et al., 2013).

AAV delivery to Kiss1Cre mice

Fourteen to twenty-one days prior to each experiment, the Kiss1Cre mice (>60 d old) received bilateral ARH injections of a Cre-dependent adeno-associated viral (AAV; serotype 1) vector encoding mCherry (AAV1-Ef1α-DIO-mCherry) or AAV1 vectors designed to encode SaCas9 and single-guide RNAs (sgRNAs) (See the SaCas9 section for specifics on the sgRNA design). Using aseptic techniques, anesthetized female mice (1.5% isoflurane/O2) received a medial skin incision to expose the surface of the skull. The glass pipette with a beveled tip (diameter = 45 μm) was filled with mineral oil, loaded with an aliquot of AAV using a Nanoject II (Drummond Scientific). ARH injection coordinates were anteroposterior (AP): −1.20 mm, mediolateral (ML): ± 0.30 mm, dorsoventral (DV): −5.80 mm (surface of brain z = 0.0 mm); 500 nl of the AAV (2.0×1012 particles/ ml) was injected (100 nl/min) into each position, and the pipette left in place for 10 min post-injection, then slowly retracted from the brain. The skin incision was closed using Vetbond (3M) and each mouse received analgesia (Rimadyl, 4-5 mg/kg, s.c.).

Generation of AAV1-FLEX-SaCas9-U6-sgTrpc5

The generation of AAV1-FLEX-SaCas9-U6-sgTrpc5 viruses were done at the University of Washington using published methods (Gore et al., 2013; Hunker et al., 2020). The constructs of sgRNAs for Trpc5 (AAV1-FLEX-SaCas9-U6sgTrpc5) were designed to target exon2 and exon7, respectively (Figure 11A and B). To achieve Trpc5 mutagenesis in Kiss1ARH neurons, Kiss1Cre mice were co-injected with AAV1-DIO-mCherry, AAV1-FLEX-SaCas9-U6-sgTrpc5-exon2, and AAV1-FLEX-SaCas9-U6-sgTrpc5-exon7 at a ratio of 10%, 45%, and 45%, respectively. Control animals were co-injected with AAV1-FLEX-SaCas9-U6-sgRosa26 and AAV1-DIO-mCherry at a ratio of 90% and 10%, respectively. AAV1-DIO-mCherry was co-injected with Cas9 vectors to confirm the targeting of injections and visualize the infected Kiss1ARH neurons. Each animal was used for experiments three weeks after the viral injection, with OVX performed one week prior to the experiments.

Visualized whole-cell patch recording

Electrophysiological and optogenetic studies were made in coronal brain slices (250 μm) containing the ARH from AAV1-EF1α-DIO-mCherry injected Kiss1Cre:GFP or Kiss1-Cre:EYFP::AI32 mice, which were vehicle-treated OVX, and E2-treated OVX females 10 weeks and older as previously described (Qiu J. et al., 2016; Qiu J. et al., 2018). Whole-cell patch recordings were performed in voltage-clamp and current-clamp as previously described (Qiu J. et al., 2018) using an Olympus BX51 W1 fixed stage scope out-fitted with epifluorescence and IR-DIC video microscopy. Patch pipettes (A-M Systems; 1.5 μm outer diameter borosilicate glass) were pulled on a Brown/Flaming puller (Sutter Instrument, model P-97) and filled with the following solution (unless otherwise specified): 128 mM potassium gluconate, 10 mM NaCl, 1 mM MgCl2, 11 mM EGTA, 10 mM HEPES, 2 mM ATP, and 0.25 mM GTP adjusted to pH 7.3 with KOH; 295 mOsm. Pipette resistances ranged from 3.5–4 MΩ. In whole-cell configuration, access resistance was less than 30 MΩ; the access resistance was 80% compensated. The input resistance was calculated by measuring the slope of the I-V relationship curve between −70 and −50 mV. Standard whole-cell patch recording procedures and pharmacological testing were performed as previously described (Qiu J. et al., 2003; Qiu J. et al., 2014). Electrophysiological signals were digitized with a Digidata 1322A (Axon Instruments) and the data were analyzed using p-Clamp software (Molecular Devices, Foster City, CA). The liquid junction potential was corrected for all data analysis.

For optogenetic stimulation, a light-induced response was evoked using a light-emitting diode (LED) 470 nm blue light source controlled by a variable 2A driver (ThorLabs, Newton, NJ) with the light path directly delivered through an Olympus 40x water-immersion lens. For high-frequency (20 Hz) stimulation the length of stimulation was 10 seconds (Qiu J. et al., 2016).

For studying the firing pattern of Kiss1ARH neurons from OVX and E2-treated OVX mice, the electrodes were filled with an internal solution consisting of the following (in mM): 120 KMeSO4, 5 Na2 phosphocreatine, 10 KCl, 2 MgSO4, 10 HEPES, 4 K2ATP, and 0.25 GTP adjusted to pH 7.3 with KOH; 290 mOsm. The bath solution as described (Ashhad and Feldman, 2020) consisted of (in mM): 124 NaCl, 3.5 KCl, 25 NaHCO3, 0.5 NaH2PO4 and 30 D-glucose,1 MgSO4, 1.5 CaCl2, gassed with 95% O2 /5% CO2 (pH 7.4, 310 mOsm).

For studying the activation/inactivation characteristics of the Ca2+ current, the electrodes were filled with an internal solution as described (Qiu J. et al., 2003) consisting of the following (in mM): 100 Cs+ gluconate, 20 TEA-Cl, 10 NaCl, 1 MgCl2, 10 HEPES, 11 EGTA, 1 ATP, 0.25 GTP, the pH was adjusted to 7.3 with CsOH at 300 mOsm. The bath solution as described (Zhang X. B. and Spergel, 2012) consisted of (in mM) 117.5 NaCl, 25 NaHCO3, 1.25 NaH2PO4, 10 TEA-Cl, 2 CaCl2, 1 MgCl2, 20 sucrose and 5 glucose, gassed with 95% O2 /5% CO2 (pH 7.4, 309 mOsm) and supplemented with 1 μM TTX, 10 μM CNQX, 50 μM AP5 and 100 μM picrotoxin. The effects of estrogen treatment on the peak current, peak current density, and activation/inactivation characteristics of calcium current were measured. Activation curves were fitted by the Boltzmann equation: I/Imax=1/{1+exp[V1/2-Vs)/k]}, where I is the peak current at the step potential Vs, Imax is the peak current amplitude, V1/2 is the step potential yielding half-maximum current, and k is the slope factor. Inactivation curves were fit with the Boltzmann equation: I/Imax = 1 - 1/ {1 + exp [(VH - V1/2)/k]}, where I is the peak current at the step potential VH, Imax is the peak current amplitude, V1/2 is the step potential at which half the current is inactivated, and k is the slope factor.

To record M-currents, pipettes were filled with an internal solution consisting of 10 mM NaCl, 128 mM K-gluconate, 1 mM MgCl, 10 mM HEPES, 1 mM ATP, 1.1 mM EGTA, and 0.25 mM GTP (pH 7.3; 290 mOsm). During voltage-clamp, we employed a standard deactivation protocol (Conde and Roepke, 2019; Roepke et al., 2011) to measure potassium currents. This involved 500-ms voltage steps ranging from –30 to –75 mV in 5-mV increments, following a 300-millisecond prepulse to –20 mV. The amplitude of the M-current relaxation or deactivation was quantified as the difference between the initial (<10 ms) and sustained current (>475 ms) of the current trace.

The bath solution for whole-cell recording of BK, SK and M currents was aCSF supplemented with 1 μM TTX, 10 μM CNQX, 50 μM AP5 and 100 μM picrotoxin.

Electrophysiological solutions/drugs

A standard artificial cerebrospinal fluid (aCSF) was used (Qiu J. et al., 2003; Qiu J. et al., 2010). All drugs were purchased from Tocris Bioscience unless otherwise specified. 1 mM TTX (Alomone Labs), 50 mM DL-2-Amino-5-phosphonopentanoic acid sodium salt (AP5), 10 mM 6-Cyano-7-nitroquinoxaline-2,3-dione disodium (CNQX), 100 mM picrotoxin, 10 mM Nifedipine (Sigma), 2 mM ω-conotoxin GVIA (ConoGVIA; Alomone Labs), 0.5 mM ω-conotoxin MVIIC (ConoMVIIC; Alomone Labs, 50 μM ω-agatoxin IVA (AgaIVA; Alomone Labs), 50 μM SNX-482 (Alomone Labs), 10 mM TTA-P2 (TTAP2; Alomone Labs), 200 mM CdCl2 (Cd2+; Sigma), 40 mM XE991 (Alomone Labs), 100 μM Iberiotoxin (Alomone Labs), 500 μM Apamin (Alomone Labs), 100 mM glutamic acid and 1 mM senktide. Stocks (1000×) were prepared in dimethylsulfoxide DMSO (picrotoxin, TTAP2 and senktide) or water (TTX, AP5, CNQX, ConoGVIA, ConoMVIIC, AgaIVA, SNX-482, CdCl2 and glutamic acid) and stored at −20°C. Aliquots of the stock solutions were stored as appropriate until needed.

Cell harvesting of dispersed Kiss1Cre neurons and real-time quantitative PCR (qPCR)

Cell harvesting and qPCR was conducted as previously described (Bosch et al., 2013). The ARH was microdissected from basal hypothalamic coronal slices obtained from female Kiss1Cre version 2 mice (Padilla et al., 2018) (n = 4-6 animals/group). The dispersed cells were visualized, patched, and then harvested (10 cells/tube) as described previously (Bosch et al., 2013). Briefly, ARH tissue was incubated in papain (7 mg/ml in oxygenated aCSF) for 50 min at 37° C then washed 4 times in low Ca2+ aCSF and two times in aCSF. For cell dispersion, Pasteur pipettes were flame polished to decreasing tip sizes and gentle trituration used to disperse the neurons onto a glass bottom dish. The plated cells were bathed in oxygenated aCSF using a peristaltic pump to keep the cells viable and clear of debris. Healthy cells with processes and a smooth cell membrane were harvested. Pipettes (World Precision Instruments; 1.5 μm outer diameter borosilicate glass) were pulled on a Brown/Flaming puller (Sutter Instrument, model P-87) to a 10 µm diameter tip. The cells were harvested using the XenoWorks Microinjector System (Sutter Instruments, Navato, CA) which provides negative pressure in the pipette and fine control to draw the cell up into the pipette. Cell pools were harvested and stored at -80°C. All cell pools were DNAse-treated using DNase1. cDNA synthesis was performed as previously described (Bosch et al., 2013).

Primers for the genes that encode for low and high voltage-gated calcium channels, TRPC5, Vglut2, large conductance calcium-activated K+ (BKα) channels, small conductance calcium-activated K+ (SK3) channels and GAPDH were designed using Clone Manager software (Sci Ed Software) to cross at least one intron-exon boundary and optimized as previously described using Power Sybr Green method (Bosch et al., 2013). Real-time qPCR controls included neuronal pools without reverse transcriptase (-RT), hypothalamic RNA with RT (+) and without RT (-), as well as water blanks. Standard curves using ARH cDNA were utilized to determine the real-time PCR efficiency (E = 10(−1/m) – 1) (Biosystems, 2006; Pfaffl, 2001). Only primers resulting in efficiencies of 90-100% were used for analysis. Primer sequences, qPCR parameters and efficiency calculations are provided in Table 1.

mRNA expression analysis

qPCR was performed on a Quantstudio 7 Flex Real-Time PCR System (Applied Biosystems) using Power SYBR Green PCR Master Mix (Applied Biosystems) according to established protocols (Bosch et al., 2013). The comparative ΔΔCT method (Livak and Schmittgen, 2001; Pfaffl, 2001; Schmittgen and Livak, 2008) was used to determine values from duplicate samples of 4 µl for the target genes and 2 µl for the reference gene GAPDH in a 20 μl reaction volume containing 1x Power SYBR Green PCR Master Mix and 0.5 μM forward and reverse primers. Three to six 10-cell pools per animal were analyzed from 4-6 animals per group. The relative linear quantity was determined using the 2-ΔΔCT equation (Livak and Schmittgen, 2001; Pfaffl, 2001; Schmittgen and Livak, 2008). Relative mRNA expression level of target genes in Kiss1Cre neurons was obtained by comparing OVX Oil-treated controls to OVX E2-treated animals. The mean Δ CT for the target genes from the OVX Oil-treated control samples was used as the calibrator. The data were expressed as n-fold change in gene expression normalized to the reference gene GAPDH and relative to the calibrator.

Experimental design and Statistical analysis

For the whole-cell patch recording, pharmacological experiments, only one cell was recorded per slice. Two to three slices were analyzed from each Kiss1Cre mouse, with at least 3-5 mice contributing to each group. For cell harvesting of dispersed Kiss1Cre-YFP neurons and qPCR measurements, 10 cells per pool and 3-6 pools from each animal were used, unless otherwise specified. Statistical comparisons between two groups were performed using an unpaired two-tailed Student’s t-test. Comparisons between more than two groups were performed using the repeated measures, multifactorial ANOVA. If a significant interaction was encountered, we then moved to the one-way ANOVA, followed by the multiple range tests as specified in the appropriate figure legends. All data were analyzed using GraphPad Prism version 6. All data are presented as mean ± standard error of the mean (SEM). Differences were considered statistically significant if the probability of error was less than 5%.

Mathematical model and simulation of neuronal behavior

A mathematical model of the Kiss1ARH neuron was developed and calibrated based on our physiological findings. Here, we employed the Hodgkin-Huxley modelling approach (Hodgkin and Huxley, 1952). Accordingly, the equation describing the membrane potential Vm of the Kiss1ARH neuron is given by where Cm is the membrane capacitance and I is the sum of 12 ionic currents: INaT and INaP are the transient and persistent sodium currents, respectively; represents the A current; represents the M current; ISK and IBK are the potassium currents through the SK and BK channels respectively; Ih the HCN current; IT the T-type calcium current; ICa represents other calcium currents (L-, N-, P/Q-, and R-type); ITRPCS represents Calcium current throught the TRPC5 channel; IGIRK potassium current through the GIRK channels. Finally, Ileak represents the contribution of leak currents. We use the Hodgkin-Huxley formalism to model current dynamics and their dependence on the membrane potential. A complete specification of all currents along with the complete table of model parameters can be found in the supplementary material. Model simulations were conducted in Matlab using the built-in numerical solver (ode45; based on an explicit Runge-Kutta (4, 5) formula). The Matlab code is available at https://git.exeter.ac.uk/mv286/kiss1-arcuate-neuron-model.